---
title: "RadNet Exploratory Data Analysis with RStudio"
author: "Wendy Bisset"
date: "August 13, 2018"
output:
  html_document:
    df_print: paged
  html_notebook:
    fig_caption: yes
editor_options:
  chunk_output_type: console
---
=============================================================

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(ggmap)
library(ggthemes)
library(choroplethr)
library(choroplethrMaps)
library(grid)
library(gridExtra)
library(GGally)
```

##Data Source and Background Information 

The US Environmental Protection Agency maintains several public databases on environmental data.[^EPA] One such database, **RadNet**,[^RadNetOverview] is a system of geographically distributed monitoring stations which sample and test for a number of radiochemical analytes (e.g., gross beta (β), cesium-137, iodine-131) in the nation's air, precipitation, and drinking water. RadNet provides historical data to estimate long-term trends in environmental radiation levels and as a means to estimate levels of radioactivity in the environment. The stations are located across the US as well as American Territories. The current database primarily consists of data collected since 1978 though some data dates back to 1973 from RadNet's precursor, ERAMS (Environmental Radiation Ambient Monitoring System).
 
[^EPA]:  <https://www.epa.gov/enviro/envirofacts-overview>
[^RadNetOverview]: <https://www.epa.gov/enviro/radnet-overview>

All data for this study was downloaded in September 2017 as .csv files from the RadNet website. According to the EPA's [Envirofacts Data Service API](https://www.epa.gov/enviro/web-services) data output is limited to 10000 rows of data at a time from a maximum of three tables. Consequently, the  data for each media type was downloaded by decade (up-to 1989, 1990-1999, 2000-2009, 2010-2017) and the resulting .csv files were loaded into RStudio and then merged[^merged] into a single R dataframe.

[^merged]: https://psychwire.wordpress.com/2011/06/03/merge-all-files-in-a-directory-using-r-into-a-single-dataframe/

```{r Load_Data, echo=FALSE, message=FALSE, warning=FALSE}
### To avoid re-appended data, dataframes are deleted when rerunning ### this chunk 
rm(air_data, drinking_water_data, Milk_data, precipitation_data, surface_water_data)

### Load/Append .csv files into 5 data frames, one for each media type
### air_data, drinking_water_data, Milk_data, precipitation_data, surface_water_data

dir_list <-  list.files('radnet_data/')

for (directory in dir_list){
  subdir <- (paste('radnet_data/', directory, sep=""))
  files_in_sub <- list.files(subdir)
  dfname <- paste(directory, '_data', sep="")  

  for (file in files_in_sub){
    file_loc <- paste(subdir, '/', file, sep="")
    # if the  dataset doesn't exist, create it
    if (!exists("db")){
        db <- read.csv(file_loc, header=TRUE)
    }
    # if the dataset does exist, append to it
    else if (exists("db")){
        temp_dataset <-read.csv(file_loc, header=TRUE)
        db <-rbind(db, temp_dataset)
       rm(temp_dataset)
    }
  }
  assign(dfname, db)
  remove(db)
}

### Cleanup Workspace
rm(dfname, dir_list, directory, file, file_loc, files_in_sub, subdir)

```

```{r consolidate_data, echo=FALSE, message=FALSE, warning=FALSE}
### First attempt to bind media datasets with loop failed
### List of dataframe names with 
### `as.list(names(which(sapply(.GlobalEnv, is.data.frame))))`
### [^df_names]:
### <http://r.789695.n4.nabble.com/getting-list-of-data-frame-names-td3864338.html>
### generates character list of dataframes (listofDF) not df object 
### dataframe names were finally hard coded into `bind_rows()`.

### convert SAMP_ID (INT) in surface_water_data to CHR as in other dfs
### ignore factor coercion to character vectors
surface_water_data$V_ERAMS_MATRIX_SAMPLE_ANALYSIS.SAMP_ID <-
  as.character(surface_water_data$V_ERAMS_MATRIX_SAMPLE_ANALYSIS.SAMP_ID)

### Consolidate media specific data into ONE dataframe: <rad_data_raw>
rad_data_raw <- bind_rows(list(air_data, drinking_water_data, 
                               Milk_data, precipitation_data,
                               surface_water_data),
                         .id ="id")
```

##Data Overview and Tidying 

The original raw consolidated dataframe has 19 variables and 504092 observations. One variable was added as a tracker in the merge operation to track the media dataframe from which the observation originated. From a cursory glance, variables of interest are monitoring station locations, sample types collected, and analytes. Dates and analysis amounts will be helpful in the investigation of these variables. 

Variable names were shortened and several variables were recast as factors with ordered levels where it made sense that an inmposed order may be useful, e.g., S(econd) < M(inute) < H(our) < D(ay) < Y(ear). In the event of possible date manipulations, the date field was changed to type DATE instead of CHR.[^as.Date]

[^as.Date]: <https://www.rdocumentation.org/packages/base/versions/3.4.1/topics/as.Date>.

```{r cleanup, echo=FALSE, message=FALSE, warning=FALSE}
#### Rename Variables ####
colnames(surface_water_data) <-gsub('.+\\.','',colnames(surface_water_data))
colnames(Milk_data) <-gsub('.+\\.','',colnames(Milk_data))
colnames(air_data) <-gsub('.+\\.','',colnames(air_data))
colnames(precipitation_data) <-gsub('.+\\.','',colnames(precipitation_data))
colnames(drinking_water_data)<-
  gsub('.+\\.','',colnames(drinking_water_data))
colnames(rad_data_raw) <-gsub('.+\\.','',colnames(rad_data_raw))

#### level/order variables of limited categories ####
### impose order, e.g., S(econd) < M(inute) < H(our) < D(ay) < Y(ear).
####################################################################
### (1) ID for original df source (numerical tie to media type)
###           1-5 :  (1) air_data, (2) drinking_water_data, 
###                  (3) Milk_data,(4)precipitation_data,
###                  (5) surface_water_data
rad_data_raw$id          <- factor(as.integer(rad_data_raw$id))
### (2) MAT_ID: material type e.g. Air, SW, DW, PPT, MILK
rad_data_raw$MAT_ID      <- factor(rad_data_raw$MAT_ID) 
### (5) ANA_UNIT: size of sample G,L,M3, MG, ML
rad_data_raw$ANA_UNIT    <- factor(rad_data_raw$ANA_UNIT)       
### (15) RESULT_UNIT: Result Units -> ACI/M3, DPM/GC, G/L, PCI/L, PCI/M3
rad_data_raw$RESULT_UNIT <- factor(rad_data_raw$RESULT_UNIT)
### (17) ANA_TYPE Analyte two types E (Element) and R (Radionuclide)
rad_data_raw$ANA_TYPE    <- factor(rad_data_raw$ANA_TYPE)
### (19) HALF_LIFE_TIME_UNIT: 5 Units of Half Life: S< M< H< D< Y
rad_data_raw$HALF_LIFE_TIME_UNIT <-
  factor(rad_data_raw$HALF_LIFE_TIME_UNIT,
         levels = c("S", "M", "H", "D", "Y"))

```

##Univariate Investigations

###Univariate Plots, Tables and Summaries

A quick summary of some of the variables shows the most common values or components of the database. Air-filter samples dominate as does the result unit pCi/m^3^ while the analytical sample size is split between liters (L) and cubic meters (m^3^).

```{r test_summaries, echo=FALSE, eval = FALSE, include=FALSE, warning=FALSE, message=FALSE}
#names(rad_data_raw)
#str(rad_data_raw)
summary(rad_data_raw$MAT_ID)
summary(rad_data_raw$RESULT_UNIT)
summary(rad_data_raw$ANA_UNIT)
```

#####Location Numbers, Cities, and States
The location numbers range from 1 to 4157, but there are only 324 unique sampling locations and these correspond to one of 289 cities or regions in the US, US territories, the Pananma Canal (PC) or Ottawa, Canada (ON). 

The entries are not evenly distributed amongst the monitoring sites. Some sites have have $\geq$ 12,000 entries while others have only a single entry. Highly monitored states have $\geq$ 20,000 entries while some areas have < 500 entries. Several cities have multiple monitoring stations. Oakridge, Tennesse tops this list with 11 different monitoring location numbers.

```{r locations, echo=FALSE, message=FALSE, warning=FALSE}
#### Variable Audit LOC_NUM, CITY_NAME, STATE_ABBR ####

### LOC_NUM -> INT  1-4157, 324 distinct locs ####
summary(rad_data_raw$LOC_NUM)
n_distinct(rad_data_raw$LOC_NUM)
ggplot(rad_data_raw, aes(LOC_NUM)) +
 geom_histogram(bins =1000, fill = "salmon")+
 labs(title = "Analysis Counts by Location Number",
      x = "Location ID", y = "COUNT")+
 theme(plot.title = element_text(hjust = 0.5))

### City Locations ####
n_distinct(rad_data_raw$CITY_NAME)
ggplot(rad_data_raw, aes(CITY_NAME)) + geom_bar()+
 labs(title = "Cities/Regions in RadNet",
      x = "CITY", y = "COUNT")+
 theme(plot.title = element_text(hjust = 0.5))

city_obs_locs <- rad_data_raw %>% 
  group_by(STATE_ABBR, CITY_NAME, LOC_NUM) %>% summarise(n = n()) %>% 
  group_by(STATE_ABBR, CITY_NAME) %>% summarise(locations = n(), 
                                                observations = sum(n)) %>% 
  arrange(desc(locations))

head(city_obs_locs, 10)
```

The top and bottom five locations, by city or state, are tabularized by number of observations for that location or area. Some locations have only 1 entry in the dataframe.

```{r EPA_regions, echo=FALSE, message=FALSE, warning=FALSE}
### CITY_NAME -> CHR, 289 Distinct City, county or EPA Region ####

### 10 EPA Regions reset to corresponding Headquarter City ####
### R01 New England (CT, ME, MA, NH, RI, VT and 10 tribal nations)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R01")]   <- "BOSTON"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R01")]  <- "MA"
### R02 NJ,NY, Puerto Rico (PR), US VI and 8 tribal nations)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R02")]   <- "New York City"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R02")]  <- "NY"
### R03 Mid-Atlantic (DE, DC, MD, PA, VA, WV)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R03")]   <- "PHILADELPHIA"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R03")]  <- "PA"
### R04 Southeast (AL, FL, GA, KY, MS, NC, SC, TN and 6 tribes)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R04")]   <- "ATLANTA"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R04")]  <- "GA"
### R05 IL, IN, MI, MN, OH, WI and 35 tribal nations                    
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R05")]   <- "CHICAGO"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R05")]  <- "IL"
### R06 South Central (AK, LA, NM, OK, TX and 66 tribal nations)      
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R06")]   <- "DALLAS"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R06")]  <- "TX"
### R07 Midwest (IA, KS, MO, NE and 9 tribal nations)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R07")]   <- "KANSAS CITY"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R07")]  <- "KS"
### R08 Mountains and Plains (CO, MT, ND, SD, UT, WY and 27 tribal nations)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R08")]  <- "DENVER"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R08")]  <- "CO"
### R09 Pacific Southwest (AZ, CA, HI, NV, Pacific Islands and 148 tribal nations)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R09")]   <- "SAN FRANCISCO"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R09")]  <- "CA"
### R10 Pacific Northwest (AK, ID, OR, WA and 271 Native Tribes)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R10")]   <- "SEATTLE"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R10")]  <- "WA"


staedte <- count(rad_data_raw,CITY_NAME)  # top cities have 5000-8000 rows
staedte <- staedte[order(staedte$n, decreasing = T),]

#### STATE_ABBR    ###############################################
###             -> 50 US States +
###                DC (District of Columbia)
###                US Territories : PR (Puerto Rico), GU (Guam), 
###                                 PC (Panama Canal), VI (Virgin Islands)
###                CNMI - Commonwealth of the No. Mariana Islands (Saipan)
###                EPA Regions R01- R10 -changed above to HDQs
###                ON Ottawa, ON
#######################################################################
### geocode does not find 'PC', changed STATE_ABBR to 'Panama' 
### Doswell SC changed to Doswell VA
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "PC")] <- "PANAMA"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "SC") &
                          (rad_data_raw$CITY_NAME == "DOSWELL")] <- "VA"

n_distinct(rad_data_raw$STATE_ABBR)

staat <- count(rad_data_raw,STATE_ABBR)  # top states
staat <- staat[order(staat$n, decreasing = T),]

mytabletheme <- gridExtra::ttheme_default(
    core = list(fg_params=list(cex = 0.5)),
    colhead = list(fg_params=list(cex = 0.5)),
    rowhead = list(fg_params=list(cex = 0.5)))


grid.arrange(tableGrob(head(staedte, n=5), theme = mytabletheme),
             tableGrob(head(staedte[order(staedte$n, decreasing = FALSE),], n=5),
                       theme = mytabletheme),
             tableGrob(head(staat, n=5), theme = mytabletheme),
             tableGrob(head(staat[order(staat$n, decreasing = FALSE),], n=5),
                       theme = mytabletheme),
             top = "City/State Most & Least Observations in RadNet Database",
             bottom = textGrob("EPA RADNet 1978-2017",
                               gp = gpar(fontface = 3, fontsize = 9),
                               hjust = 1,
                               x = 1),
             ncol = 2)
```

#####Geographical Representations
A simple search turned up the R package `ggmap`[^ggmap] which makes it possible to represent location information geographically.[^ggmap_vis]^,^[^ggmap_vis2]  The packages `choroplethr` and `choroplethrMaps` also allow for geographical representation of various values on a shaded and keyed `Choropleth Map`.[^choropleth]^,^[^choroplethr] Using the provided city and state information, the variables, latitude and longitude, were added for each location using the `geocode()` function.[^geocode]^,^[^geocode_csv]

[^ggmap]: http://cran.r-project.org/web/packages/ggmap/ggmap.pdf
[^ggmap_vis]: <https://blog.dominodatalab.com/geographic-visualization-with-rs-ggmaps/
[^ggmap-vis2]: D. Kahle and H. Wickham. ggmap: Spatial Visualization with ggplot2. The R Journal, ##5(1), 144-161. URL http://journal.r-project.org/archive/2013-1/kahle-wickham.pdf
[^geocode]: https://www.rdocumentation.org/packages/ggmap/versions/2.6.1/topics/geocode
[^geocode_csv]: http://www.storybench.org/geocode-csv-addresses-r/

[^choroplethr]: https://www.rdocumentation.org/packages/choroplethr/versions/3.6.1
[^choropleth]: https://en.wikipedia.org/wiki/Choropleth_map
[^eparegions]: https://www.epa.gov/aboutepa#pane-4

```{r csv_latlona, echo=FALSE, message=FALSE, warning=FALSE}
#### TO REDUCE RUN TIME lat/lon data obtained from Google Maps API was saved
#### to file "places_data.csv" 
#### Code for generating original file is chunk: get_latlona
#### Code chunk to obtain this info is set to "eval = FALSE" 

places <- read.csv("places_data.csv", header=TRUE)
########################################################################
```

```{r get_latlona, eval=FALSE, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
########################################################################
###  CHANGE eval=TRUE to run this code
###  Generate file with lat/lon geo data for City/State 
###  WARNING THIS CHUNK DOWNLOADS LAT/LON FROM google and takes TIME!
########################################################################

places <- data.frame("CITY"   = rad_data_raw$CITY_NAME,   
                     "STATE"  = rad_data_raw$STATE_ABBR)
places$CITY <- as.character(places$CITY)
places$STATE <- as.character(places$STATE)
places <- unique(places)
places <- arrange(places, CITY)
rownames(places) <- NULL  #reset rownames
places$citystate <- paste(places$CITY, places$STATE)
places$lat     <- NA
places$lon     <- NA
places$address <- NA


### Use tryCatch with geocode() in case of Errors from API
### to bypass error from places which have no specific LAT/LON
### A 1s pause is used between API requests, 
###########################################################################
### [Errors from Geocode]
### (https://stackoverflow.com/questions/30770328/
###  how-to-handle-error-from-geocode-ggmap-r)

miss_geo <- character()  ### Create list of geocoding potential issues/misses
for (i in 1:nrow(places)) {
  z <- 0
  repeat{
    geo_result <- tryCatch(geocode(places[i,3], output = c("latlona")),
                      warning = function(w) {
                        paste("Location Issue ", places[i,3]);
                        places[i,4] <- NA
                        places[i,5] <- NA  
                        places[i,6] <- NA
                      },
                      error = function(e) {
                          paste("Location error", places[i,3]);
                          next
                      })
    places[i,4] <- as.numeric(geo_result[2])
    places[i,5] <- as.numeric(geo_result[1])
    places[i,6] <- as.character(unlist(geo_result[3]))
    Sys.sleep(1)  # pause before next API request
    if (!is.na(places[i,6]) | z==2) break
    if (is.na(places[i,6])) {
      z = z+1
      print(paste(places[i,3], "--NA", z, i)) #check if info returned
      }
  }
  banana <- strsplit(gsub('[[:digit:]]',"", places[i,6]), ", +")[[1]]
  test_geo <-paste(toupper(banana[1]),
                     toupper(gsub('[[:blank:]]',"",banana[2])))
  if (test_geo != places[i,3]) {miss_geo <- c(miss_geo, places[i,3])}
  if (i%%5 ==0) print(i)   # visual feedback to see if code is still running
}
```

```{r fix_latlona, eval=FALSE, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
### This CHUNK is for manual Retest of GEO CODE and  Missing lat/lon data
### Check places for is.na() and find row # of NA values; 
### manually change n (row#) to resend geocode() for that location
######################################################################
count(places, is.na(places$lat))
filter(places, is.na(lat))
miss_geo # differences include county,zip code, accent marks
n=192
geo_result <- geocode(places[n,3], output = c("latlona"))
places[n,4] <- as.numeric(geo_result[2])
places[n,5] <- as.numeric(geo_result[1])
places[n,6] <- as.character(unlist(geo_result[3]))
places[n,6]

summary(places$lat)
summary(places$lon)

###  SAVE PLACES AS .csv to avoid reloading from API
write.csv(places, "places_data.csv", row.names = FALSE)

```

```{r geo_merge, echo=FALSE, message=FALSE, warning=FALSE}
#### Merge lat/lon info into rad_data_raw ####
rad_data_raw <- left_join(rad_data_raw, places, 
                          by = c("CITY_NAME" = "CITY",
                                 "STATE_ABBR" = "STATE") )
```

Most of the monitoring locations can be represented on a cropped map of North America. The eastern seaboard has the most monitoring stations. Filtering data by lat/long we find that there are 12 remaining monitoring locations, not represented on this map.

```{r northamericamap, echo=FALSE,  message=FALSE, warning=FALSE}
usa_center = as.numeric(geocode("United States"))
usa_cont_bounds = c(left = -130.0,bottom = 11.0, right = -60.0, top = 51.0)
USAMap = ggmap(get_googlemap(center=usa_center, scale=2, zoom=3, 
                             extent ="device"))
        
continental <- places %>% filter(between(lon, -130, -60) & 
                                 between(lat, 8, 55))
noncontinental <- places %>% filter(!between(lon, -130, -60))

### US Map with sampling locations
USAMap + 
  scale_y_continuous(limit = c(8,50)) +
  scale_x_continuous(limit = c(-130,-62)) +
  geom_point(data = continental, aes(x= lon, y= lat), 
             col="blue", alpha=0.4) +
  labs(caption = 'EPA RadNet Monitoring Locations 1978-2017') +
  theme_map()

grid.arrange(tableGrob(head(noncontinental, n=12), theme = mytabletheme),
             top = "Non-North American Continent RADNet Monitoring Sites",
             bottom = textGrob("EPA RADNet 1978-2017",
                               gp = gpar(fontface = 3, fontsize = 9),
                               hjust = 3,
                               x = 1))
```

Filtering again we can look closer at regional monitoring stations such as those in Hawaii and Alaska on state maps. 

```{r regionmaps, echo=FALSE,  message=FALSE, warning=FALSE}
#################################
###  HAWAII AREA MAPS
###  Using stamen map centered
#################################
HI_center <- c(left = -161, bottom = 18, right = -154, top = 23)
HImap <- get_stamenmap(bbox= HI_center, zoom = 7, maptype = "toner-lite") 
HImap <- ggmap(HImap) + theme_map()
hawaii <- noncontinental %>% filter(between(lon,-164,-150) & 
                                      between(lat,16,24))
HImap_stations <- HImap +
  geom_point(data = hawaii, aes(x= lon, y= lat), colour ="red", size = 3) +
  geom_text(data = subset(hawaii, CITY == "KAHUKU" | CITY == "KAUAI"),
            aes(label = CITY), nudge_y = 0.2, size = 2)

#################################
### ALASKA AREA MAP
#################################
AK_center = c(-149.4937, 64.20084)
AKmap <- ggmap(get_googlemap(center=AK_center, 
                             zoom=4, extent="normal")) +
  scale_y_continuous(limit = c(50, 72)) +
  scale_x_continuous(limit = c(-175,-130)) + 
  theme_map()

alaska <- noncontinental %>% filter(between(lon,-167,-130) & 
                                    between(lat,53,65))
AKmap_stations <- AKmap  +
  geom_point(data = alaska, aes(x= lon, y= lat), col="red", size = 2)

```

```{r time_sep,echo=FALSE,  message=FALSE, warning=FALSE}
###View HI, AK Maps - view doesn't work in Console if in same chunk?
###need time delay between data filtering and grid.arrange execution
grid.arrange(HImap_stations, AKmap_stations, nrow = 1, 
             top="EPA RadNet Monitoring Stations: Hawaii & Alaska")
```

<!-- Do not rerun if rad_data_raw already has region -->
```{r stateregions, echo=FALSE,  message=FALSE, warning=FALSE}
### Include *region* as recognized by choroplethr
### lower case full state name, eg. new mexico
#################################################
state_regions <- data.frame(state.abb,state.name)
state_regions$state.name <- tolower(state_regions$state.name)
colnames(state_regions) <- c("abbreviation", "region")
state_regions$abbreviation <- as.character(state_regions$abbreviation)

### Add regions for places not included in state.abb/state.name
#################################################
state_regions <- rbind(state_regions, c("abbreviation" = "PR",
                                  "region" = "puerto rico"))
state_regions <- rbind(state_regions, data.frame("abbreviation" = "DC", 
                                   "region" = "district of columbia"))
state_regions <- rbind(state_regions, data.frame("abbreviation" = "GU", 
                                   "region" = "guam"))
state_regions <- rbind(state_regions, data.frame("abbreviation" = "VI", 
                                   "region" = "virgin islands"))
state_regions <- rbind(state_regions, data.frame("abbreviation" = "ON", 
                                                 "region" = "ontario"))
state_regions <- rbind(state_regions, data.frame("abbreviation" = "PANAMA",
                                                 "region" = "panama"))
state_regions <- rbind(state_regions, data.frame("abbreviation" = "CNMI", 
                                   "region" = "mariana islands"))

places <- left_join(places, state_regions,
                    by = c("STATE" = "abbreviation"))
rad_data_raw <- left_join(rad_data_raw, state_regions,
                          by= c("STATE_ABBR" = "abbreviation"))
rm(state_regions)

```

#####Material ID -> Sample Types
A bar graph yields a better visual of the distribution of sample material. As noted above, air samples truly dominate with nearly $\frac{1}{2}$ of the entries at a quarter million observations. 

```{r variable_matID, echo=FALSE, message=FALSE, warning=FALSE}
### MAT_ID -> type CHR, 6 material types - air_data is subdivided 
###               AIR-CHARCOAL (Discontinued in 1980s) & AIR-FILTER
#######################################################################

sample_types <- ggplot(rad_data_raw, aes(MAT_ID)) + 
  geom_bar(fill = "SteelBlue") +
  scale_x_discrete(limits = c("AIR-FILTER", "PRECIPITATION",
                   "PASTEURIZED MILK", "DRINKING WATER", 
                   "SURFACE WATER", "AIR-CHARCOAL")) +
  geom_text(stat='count',aes(label=..count..), color="black")+
  coord_flip() +
  labs(title="Number of RadNet Entries by Media Type", 
       caption = 'EPA RadNet Data 1978-2017',
       x=NULL, y = "Count") +
  theme(plot.title = element_text(hjust=0.5),
        axis.title.x = element_blank(),
        legend.title = element_blank(),
        axis.title.y = element_blank(),
        plot.caption = element_text(size = rel(0.5)))
sample_types
```

#####Sample IDs and Analytes
Just over $\frac{1}{2}$ of the sample ID numbers are unique and most of these Sample IDs (84%) are single entries. The remaining samples have as many as 25 entries (replicates). Perhaps this should not be surprising as there are 61 different analytes for which each sample might be analysed. *Gross Beta* accounts for nearly $\frac{1}{2}$ of the observations, but the top 10 most analysed analytes account for ~85% of the dataframe observations.

```{r variable_sampanaID, echo=FALSE, message=FALSE, warning=FALSE}
### SAMP_ID -> type CHR, sample ID number, some are alpha-numeric, 
###                roughly 1/2 the number of observations, i.e, multiple
###                observations (analyses) for a single SAMP_ID
#######################################################################
#n_distinct(rad_data_raw$SAMP_ID)/length(rad_data_raw$SAMP_ID)
samp_replicates <- rad_data_raw %>%
                   group_by(SAMP_ID) %>%
                   summarise(n = n()) 

sample_reps <- ggplot(samp_replicates, aes(n)) + 
  geom_histogram(bins = 25,fill ="cornflowerblue") +
  scale_y_sqrt() +
  labs(title = "Distribution of Sample ID Replicates",  
       x = "Sample ID Replicates", y = "COUNT")

### ANALYTE_ID -> CHR,  61 unique analytes of interest
### Most common Gross Beta 232,387
###################################################################
analytes <- rad_data_raw %>% group_by(ANALYTE_ID) %>% 
  summarise(n = n(), percent_tot = round(n()/504092*100,2))
top_analytes <- filter(analytes, n > 9000)
top_analytes <- arrange(top_analytes, desc(n))

mytabletheme <- gridExtra::ttheme_default(
    core = list(fg_params=list(cex = 0.5)),
    colhead = list(fg_params=list(cex = 0.5)),
    rowhead = list(fg_params=list(cex = 0.5)))

colnames(top_analytes) <- c("Analyte", "n", "% Observations")

grid.arrange(sample_reps, 
             tableGrob(top_analytes, theme = mytabletheme), 
             top = "Samples & Analytes in RadNet Database",
             bottom = textGrob("EPA RADNet 1978-2017",
                               gp = gpar(fontface = 3, fontsize = 9),
                               hjust = 1,
                               x = 1),
             nrow = 1)

View(samp_replicates %>% group_by(n) %>% summarise(reps = n())
     %>% mutate(samples = reps/n))

```

#####Sample Sizes and Units of Measure
While sample size values are grouped about the values 1, 5 and 5,000, this does not provide a valid comparison of *size* as the *unit size* is needed to provide context. A 1 L sample is 1000 times the size of a 1 mL sample. The units, while clarifying relative size can also indicate sample type: i.e., mL and L are generally for liquids; mg and g for solids and ($m^3$) is for gas samples. A small fraction of entries have no designated unit (NA), while about $\frac{1}{2}$ of the measured units are for gases ($m^3$). 

This ties well to the earlier observations that the predominant samples are air-filter & gross beta. It is likely that $\frac{1}{2}$ of the observations in this database are gross beta measurements in air samples ($m^3$). The next biggest group of sample units is the liter (L) which matches well with the remaining sample types being liquids: i.e, precipitation, milk, and water (drinking & surface)

```{r variable_anaUnit_size, echo=FALSE, message=FALSE, warning=FALSE}
### ANA_SIZE -> type NUM, numeric ####
sample_sizes <- ggplot(rad_data_raw, aes(ANA_SIZE)) + 
  geom_histogram(bins = 50) +
  scale_x_log10(labels=scales::comma, breaks = c(1, 10, 1000,5000)) +
  labs(title = "Sample Sizes ... not Normalized for Unit of Measure",  
       x = "Numeric Sample Size", y = "COUNT")
sample_sizes

### ANA_UNIT ->CHR, measure unit for analysis size ####
###   (air = m3, solids = mg, g, liquids = mL, L, "" blank 3206)

table(rad_data_raw$ANA_UNIT)     # M3 followed by L

```

#####Analytical Procedures and Duration
There are 35 analytical procedure numbers with **Procedure 1** returned as the mode[^getmode] of this variable (~42% of the entries). Procedures 1 and 9 account for 74% of the entries and a small high density grouping about 120 shows procedures 118 and 119 accounting for another 19 % of the entries.  Thus, most (93%) of the entries are analyzed with one of four procedures. A duration variable indicates that while most tests take under 20 hours there are some tests/procedures which are over 80 hours.

[^getmode]:  Getmode function https://www.tutorialspoint.com/r/r_mean_median_mode.htm
```{r variable_procedures_dur, echo=FALSE, message=FALSE, warning=FALSE}
### ANA_PROC_NUM -> INT; 35 analytical procedures used for analysis ####
###               Proc Num range from 1 to 170 with mode = ProcNum 1

getmode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}

ggplot(rad_data_raw)+
  geom_violin(aes(x= "Procedure Frequency", y = ANA_PROC_NUM))

### DURATION -> NUM range 1-5000, raw duration in mins, convert to hrs ####
test_duration <- ggplot(rad_data_raw, aes(DURATION/60)) + 
  geom_histogram(bins = 75, fill = "darkseagreen") +
  scale_y_log10(labels=scales::comma) +
  labs(title = "Testing Duration", 
       caption = 'EPA RadNet Data 1978-2017',
       x = "Duration (h)", y = "COUNT")+
   theme(plot.title = element_text(hjust = 0.5),
         plot.caption = element_text(size = rel(0.5)))

test_duration

```

#####Result Amount, MDC, CSU, and Units of Measure
Result amounts like sample sizes cannot be directly compared as this is more or less meaningless without a Result Unit. However, a plot of the raw result values show the greatest density around 0.01. The overlayed plots of the minimum detectable concentration (MDC) and combined standard uncertainty (CSU) shows that both MDC and CSU have their greatest density at <0.001, a factor of 10 lower than the results. Though this would need confirmation of the unit of measurement the distribution is reassuring as the uncertainties and detection limits should be much less than the actual measurment. 

Result units themselves are distributed similar to the sample units in that about $\frac{1}{2}$ of the measured units are for gases (pCi/m^3^ and aCi/m^3^).

Sixteen percent (81,921) of the observations have no result entry, result amount is NA. This seems odd as RadNet is designed specifically for the documentation of analtye concentrations, i.e., results.  A separate dataframe was created which filters out the empty RESULT_AMOUNT observations, *rad_data*. Using a new *complete* variable (Y/N) derived from the full and NA's entries the result amounts can represented by 1 and 0, respectively which allows for proportional analysis.

```{r result_amts, echo=FALSE, message=FALSE, warning=FALSE}
## RESULT_AMOUNT -> NUM, -200 - 2.57e5, 81291 NA's
summary(rad_data_raw$RESULT_AMOUNT)
### CSU -> NUM, 0-15000 Combined Standard Uncertainty,  81921 NA's (16%) ####
summary(rad_data_raw$CSU)
### MDC -> NUM, 0-9700 Min Detectable Concentration; 218822 NA's (43%) ####
summary(rad_data_raw$MDC)

ggplot(rad_data_raw, alpha = 0.5)+
  geom_density(aes(x=RESULT_AMOUNT, fill = "RESULT"))+
  geom_density(aes(x=CSU, fill="CSU")) +
  geom_density(aes(x=MDC, fill="MDC")) +
  scale_x_log10(labels=scales::comma, limits = c(0.000015,25000),
                breaks = c(0.0001, 0.01, 1, 100)) +
  labs(title = "Results, Combined Standard Uncertainty, Min Detectable Concentration",
       subtitle ="(no Unit Normalization)", 
       x = 'RAW Values',
       y = 'Count')

### RESULT_UNIT ->  
###                (air = ACI/m3, PCI/M3 (45%); solids = g/L, DPM/GC;
###                 liquids = pCi/L (48%))                            ####
table(rad_data_raw$RESULT_UNIT)

###New variables for future analyses of empty/complete entries       ####
rad_data_raw$complete <- ifelse(is.na(rad_data_raw$RESULT_AMOUNT), "N", "Y")
rad_data_raw$Y <- ifelse(rad_data_raw$complete == "Y", 1, 0)
rad_data_raw$N <- ifelse(rad_data_raw$complete == "N", 1, 0)
   
### Create df without RESULT_AMOUNT == NA                             ####
rad_data <- rad_data_raw[complete.cases(rad_data_raw[,"RESULT_AMOUNT"]),]
```

#####Result Dates 
A Simple distribution of entries over result dates, yields typical count ranges are about 125 to 250 entries, but there are regular spikes 50% to 100% above the baseline. These spikes occur on annual intervals; until 2009 the spikes are at mid-year (7/1) afterwhich they appear year-end (12/31). Grouping the observations by date ranges gives a histogram-like plot in which you can see a drop in entries around 1989 and again 10 years later, though there is a subtle increase from 1999 to 2012. The drop in 2017 cannot be verified as the data was only downloaded through September.

```{r variable_date, echo=FALSE, message=FALSE, warning=FALSE}
###  RESULT_DATE -> CHR transform to DATE range from 7/1/1978 to 7/26/2017
###  DATE set to ISO 8601 format:  %F == "%Y-%m-%d"
#######################################################################
rad_data_raw$RESULT_DATE <- as.Date(rad_data_raw$RESULT_DATE, "%F")

analyses_bydate <- ggplot(rad_data_raw, aes(RESULT_DATE)) + 
  geom_bar(color = "steelblue ", stat = "count") +
  labs(title="Number of Analyses Results by Date", x=NULL, y = "Count") +
  theme(plot.title = element_text(hjust = 0.5))
analyses_bydate

View(rad_data_raw %>% group_by(RESULT_DATE) %>% select(RESULT_DATE, RESULT_AMOUNT) %>% 
                 summarise(n = n())  %>% filter(n>400))

### New Vectors to group data by date ranges of 6 mos, 1 yr, 2 yr       ####
rad_data_raw$date_range2 <- cut(rad_data_raw$RESULT_DATE, breaks = '2 years')
rad_data_raw$date_range1 <- cut(rad_data_raw$RESULT_DATE, breaks = '1 years')
rad_data_raw$date_range <- cut(rad_data_raw$RESULT_DATE, breaks = '6 months')

analyses_by2years <- ggplot(rad_data_raw, aes(date_range2)) + 
  geom_bar(color = "steelblue ", stat = "count") +
  labs(title="Number of Analyses Results by 2 year bins", x=NULL, y = "Count") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x= element_text(angle=20,size = rel(0.5)))
analyses_by2years

```

#####Analysis Types and Half-Lives
There are two types of analyses: radioactive (R) 97.9 % and elemental (E) 2.1%. One might think it odd that only 50% of the entries include a radioactive half-life ($t_{1/2}$). However, as noted gross beta accounts for 50% of the entries and a half-life is only defined for the single isotope of an particular element, gross beta is not isotope specific. So this is understandable and we can confirm that the half-life entries for beta are NA's. There are 56 distinct elements with half-lives, but the top 10 account for a majority of the remaining dataframe.

Half-life values by themselves are like sample size and result amount and cannot be compared directly. However, the values in this case can be easily converted to a single time unit, e.g., *years*  and then plotted. This wonderfully captures the breadth of half-lives that is involved within this group of isotopes. The shortest-lived isotope, radon-219 has $t_{1/2}$ of 3.9 seconds compared to the longest-lived isotope, lanthanum-138 with $t_{1/2}$ = 1.05x10^11^, 105 billion years.

```{r variable_halflife, echo=FALSE, message=FALSE, warning=FALSE}
### ANA_TYPE -> CHR, "E" (Element) = 10585, "R" (Radionuclide)= 493507
#######################################################################
table(rad_data_raw$ANA_TYPE)
summary(rad_data_raw %>% select(ANALYTE_ID, HALF_LIFE) %>% filter(ANALYTE_ID == "BETA"))

### HALF_LIFE -> value of an isotopes half-life, 251749 50% NA's  ####
element_half_life <- subset(rad_data_raw[ , c(11,18,19)], !is.na(HALF_LIFE))
element_half_life <- within(element_half_life,
                     ANALYTE_ORDER <- factor(ANALYTE_ID,
                                      levels=names(sort(table(ANALYTE_ID)))))
element_life <- ggplot(element_half_life,
                       aes(x = ANALYTE_ORDER)) + 
  geom_bar(fill = "SteelBlue") +
  scale_y_log10() +
  geom_text(stat='count',aes(label=..count..), color="black")+
  coord_flip() +
  labs(title="Count of Analyses for Analytes with Half Life", 
       x=NULL, y = "Count (log10)") +
  theme(plot.title = element_text(hjust = 0.5))

n_distinct(element_half_life$ANALYTE_ID)

element_life

### HALF_LIFE_TIME_UNIT -> CHR; D,H,M,S,Y and 50% blank (251749) ####
###      variable half_lives to normalized years & sorted
half_lives <- unique(element_half_life) %>% 
  mutate(HALF_LIFE_YEAR = case_when(HALF_LIFE_TIME_UNIT =="S" 
                                    ~ HALF_LIFE/3153600,
                                    HALF_LIFE_TIME_UNIT =="M" 
                                    ~ HALF_LIFE/525600,
                                    HALF_LIFE_TIME_UNIT =="H" 
                                    ~ HALF_LIFE/8760,
                                    HALF_LIFE_TIME_UNIT =="D" 
                                    ~ HALF_LIFE/365,
                                    HALF_LIFE_TIME_UNIT =="Y" 
                                    ~ HALF_LIFE/1))
half_lives <- within(half_lives, sort(HALF_LIFE_YEAR))

isotope_years <- ggplot(half_lives,
                       aes(reorder(ANALYTE_ID, -HALF_LIFE_YEAR, median),
                           HALF_LIFE_YEAR)) + 
  geom_col(fill = "SteelBlue") +
  scale_y_log10(
   breaks = scales::trans_breaks("log10", function(x) 10^x, n=7),
   labels = scales::trans_format("log10", scales::math_format(10^.x))
   ## http://ggplot2.tidyverse.org/reference/annotation_logticks.html
   ) +
  coord_flip() +
  labs(title="Analyte Isotope Half-Lives ", 
       x=NULL, y = "Half-Life Years (log10)") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.y= element_text(angle=20,size = rel(0.5)))
isotope_years

```

###Univariate Analysis and Review

#### Dataset structure:
The original dataframe had 504,092 entries with 19 variables. The variables represented factors (some levelled), numerical values, characters and dates. Some of the variables were recast for uniformity and to help data manipulation and representation. Most notably the dates were converted to date characters in ISO 8601 format.

#### Main features of interest:   
* Material ID: milk, air: filter and charcoal, water: surface and precipitation
* Location: Geographic information over monitoring (sampling) locations
* Result Amount 
    + comparative amounts 
    + empty result entries (NANs)

#### Support features in the dataset:
The main features of interest should be great in investigation of their own interactions. However, the dates should also be very useful in introducing a time component to the study.  
Result values in combination with Location IDs and dates would help isolate groups of data and identify any radioactive release events or concerns. Analytes tested should also help to subset and classify data to provide better comparisons. Additionally, Sample IDs may help to find duplicate entries or to better understand missing data.

#### New variables created from existing variables:
Several new variables were created for the handling of the geographical information, e.g., latitude and longitude. The R packages `ggmap` and `choroplethr` and the API with `geocode` all require specific formatting of input data which in turn required reformatting of variables, e.g., NM to new mexico.  

Three variables were added to clarify and provide better handling of the empty *result amount* (NA). One variable *complete* is a simple character variable with Y(es) and N(o) to identify whether there is a result value or not. Two numeric variables were added with the hopes to better quantify and classify the Y and N values.   

The dates were grouped into 3 different time spans to provide a way for a more simplified examination of the dataframe as well as to investigate duplicate Sample IDs. 

#### Data Cleansing/Unusual Distributions:
In general, most cleaning and date munging involved verifying and reformatting location data. Column names were tidied and shortened. A few database issues were found, e.g.: the abbreviation *PC* was not recognized by the Google API and was changed to "Panama Canal"; the city of Doswell, SC does not appear to exist, but several Doswell, VA observations were found in the dataframe so the single Doswell, SC entry was changed to VA. Several entries were listed as EPA Regions 1 to 10. In an effort to geo-locate the data, each region number was changed to the location of the headquarters for that EPA region.[^eparegions]. 

Because of the number of missing result entries a duplicate dataframe (rad_data) was created which filtered out the NA rows for RESULT_AMOUNT. 

A couple other features of note:  1) The periodic spike in entries occurs July 31 and December 31. Is this a side effect of the database, precipitated by some administrative speciifcation or were there really that many year-end entries, on New Year's Eve?  2) There is any interesting long tailed positive skew for the distribution of sample replicates. So as there is the possibility for multi-analyte analysis, most samples are primarily sampled only for gross beta analysis.  


##Bivariate Investigations

###Bivariate Plots 

#####Correlations
A heat map correlation matrix indicates that most of the numerical variables are not well correlated. The new variables Y/N are negatively correlated and the measurement variables (result amount, CSU and MDC) are stongly correlated.  

A general correlation matrix^[^ggpairs] on a 10% sample of the database reveals how the majority of the observations may be classified. Variables with multiple levels, such as cities and states are not easy to include in this type of matrix. To provide correlations for observations by location some grouping is necessary.

[^ggpairs]: http://koaning.io/ggally-explore-all-the-things.html

```{r corr_matrix,echo=FALSE,  message=FALSE, warning=FALSE }
numeric_cols <- c("ANA_SIZE", "ANA_PROC_NUM", "LOC_NUM",
                  "RESULT_AMOUNT", "CSU","MDC","HALF_LIFE","lat",
                  "lon", "Y", "N")
cat_cols <- c("MAT_ID","ANA_PROC_NUM",
              "RESULT_UNIT","ANA_TYPE","complete")
ggcorr(rad_data_raw[,numeric_cols], label = TRUE, label_color = 'white')

## 10% sampling of data base
set.seed(999)
test_10 <- rad_data_raw[sample(nrow(rad_data_raw), 50000, replace = FALSE),]

test_10 %>% select(cat_cols) %>% 
  ggpairs(mapping=ggplot2::aes(colour = MAT_ID)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

rad_data_raw %>% filter(ANA_TYPE == "E") %>% select(MAT_ID, ANALYTE_ID, ANA_PROC_NUM) %>% 
  group_by(MAT_ID, ANALYTE_ID, ANA_PROC_NUM) %>% 
  summarise(n=n(), proc = mean(ANA_PROC_NUM))

```

#####Monitoring Locations and Entries
When the monitoring stations are grouped within a region there is a not surprising strong correlation (r=0.87) with the number of observations. Essentially, with more monitoring stations there are more entries. However, the correlation is weaker (r=0.79) when examining the number of empty results to the number of regional monitoring stations. Using a choropleth map we can plot the number of observations entered by state and compare information by location. It is easy to pick out individual states with moderate/low number of observations and higher percentages of empty results, e.g., Kentucky with <4,500 entries has > 17% empty, while California with 23,686 entries has <15% NA results. 

```{r uschoropleth, echo=FALSE,  message=FALSE, warning=FALSE}
### Choropleth of Sample Analyses by State
bystate <- rad_data_raw %>% group_by(region) %>%
  summarise(value = n())

bystate <- left_join(bystate, (rad_data_raw %>% group_by(region, LOC_NUM) %>%
               summarise(obs_station = n()) %>% summarise(stations = n())),
                          by = "region")

bystate <- left_join(bystate, (rad_data_raw %>% filter(is.na(RESULT_AMOUNT)) %>% 
               group_by(region) %>%
               summarise(empty = n())),
                          by = "region")

bystate <- bystate[order(-bystate$stations),]
head(bystate)
#### Number of Stations and Observations(Empty)
cor_station_obs <- round(cor(bystate$stations, bystate$value, 
                             method = "pearson"),2)
cor_station_empty <- round(cor(bystate$stations, bystate$empty, 
                               method = "pearson"),2)

station_obs <- ggplot(bystate, aes(x = stations, y=value)) +
  geom_point() +
  geom_smooth(method = "lm") +
  annotate("text", x = 5, y=27500, label = cor_station_obs)

station_NAs <- ggplot(bystate, aes(x = stations, y=empty)) +
  geom_point() +
  geom_smooth(method = "lm") +
  annotate("text", x = 5, y=5000, label =  cor_station_empty)

grid.arrange(station_obs,station_NAs, ncol = 2, nrow = 1)

###Choropleths Location(geographic) and Observations(Empty)
choro_entries <-state_choropleth(bystate, 
                 title="RADNet Analyses by State") + 
                 labs(subtitle = '1978-2017',
                      caption = 'EPA RadNet Data') +
  theme(plot.caption = element_text(size = rel(0.5)),
         plot.title = element_text(hjust=0.75),
         plot.subtitle = element_text(size=rel(.75), hjust = 0.5))

### Choropleth of Empty Results by State
### New Colour Scheme
### https://www.r-bloggers.com/advanced-choroplethr-changing-color-scheme/
col.pal<-brewer_pal(palette = "GnBu")(7)

colnames(bystate)[2] <- "totalanal"
bystate$value <- round(bystate$empty/bystate$totalanal*100, digits =1)
state_nan <-StateChoropleth$new(bystate)
state_nan$title <- "Empty Result RADNet Entries by State"
state_nan$ggplot_scale <- scale_fill_manual(name="% Empty Result Entries",
                                         values=col.pal, drop=FALSE)
  
choro_nan <-state_nan$render()  +
  theme(plot.caption = element_text(size = rel(0.5)),
         plot.title = element_text(hjust=0.5),
         plot.subtitle = element_text(size=rel(.75), hjust = 0.5)) + 
  labs(subtitle = '1978-2017', caption = 'EPA RadNet Data')

choro_entries
choro_nan
```

<!-- #RCHUNK Just for creation of function shared_legend -->
```{r shared_legend,include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
#####################################################################
#### R chunk for function  with common legend function for multiple graphs: Grid.arrange 
#### http://www.guru-gis.net/
###  share-a-legend-between-multiple-plots-using-grid-arrange/
#https://andyphilips.github.io/blog/2017/04/04/single-legend-for-multiple-plots.html
#####################################################################
grid_arrange_shared_legend <- function(..., ncol = length(list(...)), 
                                       nrow = 1, position = c("bottom", "right")) {
    plots <- list(...)
    position <- match.arg(position)
    g <- ggplotGrob(plots[[1]] + 
                      theme(legend.position=position,
                            legend.key.size = unit(2,"mm")
                            ))$grobs
    legend <- g[[which(sapply(g, function(x) x$name) == "guide-box")]]
    lheight <- sum(legend$height)
    lwidth <- sum(legend$width)
    gl <-lapply(plots, function(x) x + theme(legend.position="none"))
    gl <- c(gl, ncol= ncol, nrow = nrow)
    combined <- switch(position, 
                       "bottom" = arrangeGrob (do.call(arrangeGrob, gl),
                                               legend, ncol = 1, 
                                               heights = unit.c(unit(1,"npc")-
                                                                  lheight, lheight)),
                       "right" = arrangeGrob (do.call(arrangeGrob, gl),
                                               legend, ncol = 2, 
                                               widths = unit.c(unit(1,"npc")-
                                                                  lwidth, lwidth)))
    grid.newpage()
    grid.draw(combined)
    
    #return gtable invisibly
    invisible(combined)
}
```

#####Hawaiian Subset
Regional subsets of the data are useful not only to focus on a specific geographic area, but to simplify initial data manipulations before working with the entire data set. Observations from Hawaii were filtered and using a simple state shows the majority of samples are from Honolulu. Using a simple material ID facet, we can see while Honolulu has the most varied of sample types, air-Filters are the most common material. 

```{r HI_minireviews, echo=FALSE, message=FALSE, warning=FALSE}
### REVIEW OF DATA with smaller data subset - Hawaii
##########################################################
### Map of Analyses in HI
#####################################
HI_data <- rad_data_raw %>% filter(STATE_ABBR == "HI")

HI_map_matidA <- HImap + geom_count(data = subset(HI_data,
                                                 !is.na(RESULT_AMOUNT)), 
                                   aes(x= lon, y= lat), 
                                   na.rm = TRUE) +
  labs(title=  'Count of RADNet Analyses in Hawaii',
       subtitle = '1978-2017',
       caption = 'EPA RadNet Data',
       colour = element_blank()) +
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_text(hjust=0.5),
        plot.subtitle = element_text(size=rel(.75), hjust = 0.5),
        legend.position = c(0.05, 0))

### Map of Materials and Observation Counts in HI
#####################################
HI_map_matidB <- HImap + geom_count(data = subset(HI_data,
                                                 !is.na(RESULT_AMOUNT)), 
                                   aes(x= lon, y= lat, color = MAT_ID), 
                                   na.rm = TRUE,
                                   position = position_jitter(width = 0.1, 
                                                              height = 0.1)) +
  labs(title=  'RADNet Sample Type Analyses in Hawaii',
       subtitle = '1978-2017',
       caption = 'EPA RadNet Data',
       colour = element_blank()) +
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_text(hjust=0.5),
        plot.subtitle = element_text(size=rel(.75), hjust = 0.5),
        legend.position = c(0,0), legend.box.background = element_rect())


grid.arrange(HI_map_matidA, HI_map_matidB,  ncol = 2, nrow = 1)

```

Further examination of the material types in Hawaii show that while air-filter samples are the majority of samples, very few of those entries are empty. The precipition samples have the most empty entries. In 2011 an anomalous spike of empty air-charcoal entries appears, for which there seems to be a corresponding spike of completed air-filter samples. As air-charcoal samples were phased out in the 1980s it seems possible that the charcoal samples were entered in error as air-charcoal and were then re-entered as air-filter samples. Additionally, there are no surface water analyses entered for Hawaii.

```{r HI_nans,include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
#### Comparison of empty (NA's) RESULT_AMOUNT HI ####
#HI_data %>% group_by(MAT_ID, complete) %>% summarise(n = n())

HI_matIDna <- ggplot(subset(HI_data, is.na(RESULT_AMOUNT)), 
                   aes(RESULT_DATE, fill = MAT_ID)) +
  geom_histogram(bins = 25, alpha = 0.6) +
  labs(title = 'Empty Result Entries',
       subtitle = 'Hawaii',   
       y = 'Number') +
  theme(plot.title = element_text(hjust=0.5),
        plot.subtitle = element_text(size=rel(.75), hjust = 0.5),
        axis.title.x = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(size = 5),
        legend.position = c(0.3,0.8))

HI_matIDnna <- ggplot(subset(HI_data, !is.na(RESULT_AMOUNT)),
                       aes(RESULT_DATE, fill = MAT_ID)) +
  geom_histogram(bins = 40, alpha = 0.6)+
  labs(title = 'Completed Results',
       subtitle = 'Hawaii',
       caption = 'EPA RadNet Data 1978-2017',
       y = 'Number') +
  theme(plot.title = element_text(hjust=0.5),
        axis.title.x = element_blank(),
        plot.subtitle = element_text(size=rel(.75), hjust = 0.5),
        legend.title = element_blank(),
        axis.title.y = element_blank(),
        plot.caption = element_text(size = rel(0.5)),
        legend.position = 'none')

grid.arrange(HI_matIDna, HI_matIDnna, ncol = 2, nrow = 1)

## To remove initial blank page created from shared_legend 
## save as .pdf onefile = FALSE
## https://stackoverflow.com/questions/12481267/
## in-r-how-to-prevent-blank-page-in-pdf-when-using-gridbase-to-embed-subplot-insi

#pdf("HI_mats.pdf",onefile = FALSE)
#grid_arrange_shared_legend(HI_matID, HI_matIDnans, ncol = 2, nrow = 1, position = "bottom")
#dev.off()
```

Further examination of the Hawaiian air samples confirms that around April 2011, 378 air analyses were entered. One-half as air-charcoal (190) with 128 empty and the remainder as air-filter analyses (188) with only 21 incomplete. The air-charcoal samples are from Kahuku and Kauai monitoring stations as are a number of the air-filter samples. However, upon a closer look, the air-filter sample spike is actualy earlier than the air-charcoal spike. So while an interesting detail, more information would be needed to establish if there is a relationship between the empty air-charcoal and air-filter samples.

```{r HI_airsamples, include=FALSE, echo = FALSE, message=FALSE, warning=FALSE}
HI_airentries <- subset(HI_data, RESULT_DATE > '2011-03-15' &
                                 RESULT_DATE < '2011-04-30' &
                                (MAT_ID == 'AIR-CHARCOAL' | 
                                 MAT_ID == 'AIR-FILTER'), 
                        select = c(SAMP_ID, MAT_ID, ANALYTE_ID,RESULT_AMOUNT, 
                                   CITY_NAME, RESULT_DATE, complete))
dim(HI_airentries)
HI_airentries_group <- HI_airentries %>%  group_by(MAT_ID, complete) %>%
  summarise(n = n())
HI_airentries_group 
n_distinct(HI_airentries$SAMP_ID)
HI_airentries %>%  group_by(SAMP_ID, MAT_ID) %>% summarise(n = n())
filter(HI_airentries, MAT_ID == "AIR-FILTER")  %>%  
  group_by(RESULT_DATE) %>% summarise(n = n())

HI_airsamp <- ggplot(HI_airentries,
                   aes(RESULT_DATE, fill = CITY_NAME)) +
  geom_histogram(bins = 50, alpha = 0.6) +
  facet_wrap(~MAT_ID)+
  labs(title = 'Hawaiian Air Samples Spring 2011',
       y = 'Number') +
  theme(plot.title = element_text(hjust=0.5),
        axis.title.x = element_blank(),
        axis.text.x = element_text(size = rel(.75), angle = 20, hjust=1),
        legend.title = element_blank(),
        legend.text = element_text(size = rel(0.5)),
        legend.key.size = unit(3,"mm"))
HI_airsamp

filter(rad_data_raw, MAT_ID =="AIR-CHARCOAL"  & is.na(RESULT_AMOUNT)) %>% group_by(region) %>% summarize(n=n())

```
A quick survey of the databases shows that only a few locations have empty air-charcoal entries with Alaska and Hawaii having the greatest number. 

#####Empty Entries and Completed Samples All Regions
 To examine the preponderance of empty samples a table of proportions was generated and frequency distributions of the complete/incomplete observations plotted by date. Interestingly, empty (incomplete) entries only exist from 1990 to 2011. During this 20 year interval, the ratio of empty/completed entries is fairly steady at ~0.37. 

```{r all_nans, echo=FALSE, message=FALSE, warning=FALSE}
### Change facet lables    ####
# https://stackoverflow.com/questions/3472980/ggplot-how-to-change-facet-labels
fertig <- list('N' = "Empty", 'Y' = "Complete")
fertig_labeller <- function(variable, value){
  return(fertig[value])
  }

### Complete/Empty Result Amounts by date range 6 mos  ####
nans_date_type <- prop.table(table(rad_data_raw$date_range, 
                                   rad_data_raw$MAT_ID, rad_data_raw$complete), 1)
nans_date_type <- as.data.frame(nans_date_type)
nans_date_type$Var1 <- as.Date(nans_date_type$Var1, "%F")
colnames(nans_date_type) <- c('Date_Range','Sample_Type','Complete','Frequency')

nans_freq <- ggplot(nans_date_type, aes(x= Date_Range, y = Frequency)) +
  geom_bar(stat="identity") +
  labs(y="Frequency") +
  facet_wrap(~Complete, labeller = fertig_labeller) +
  theme(axis.text.x = element_text(angle = 30, size = rel(0.7)),
        axis.title.x = element_blank(),
        legend.title = element_blank())

grid.arrange(nans_freq, top = "Proportion of RADNet Database Entries 1978-2017 Completed/Empty")

```

Using the *complete* variable the observations status can be examined in relation to the other support variables. For example, by analyte, of the 61 analytes, only 15 have observations with an empty *RESULT_AMOUNT*. Cesium-237 tops this list with 17,938 empty entries. More interesting though is Thorium-234 has 163 empty entries, but no completed entries. 
Empty entries by material ID reveals that proportionally, Air-Charcoal is the most incomplete while Air-Filter is the most complete. On an actual count basis Precipitation and Pasteurized Milk have the most blank result values. Just two analytical procedure numbers have empty entries, procedures 9 and 118. 

```{r incompletes, echo=FALSE, message=FALSE, warning=FALSE}
#### by analyte    ####
analyte_complete <- table(rad_data_raw$ANALYTE_ID, rad_data_raw$complete)
analyte_incomplete <- as.data.frame(analyte_complete)
analyte_incomplete <- spread(analyte_incomplete, Var2, Freq)
names(analyte_incomplete)[1]<- "ANALYTE"
analyte_incomplete <- analyte_incomplete[order(-analyte_incomplete$N),]
analyte_tests <-head(analyte_incomplete, n=15)

### grid.table(analyte_tests)

#### Analyte Complete/Incomplete Table
tt2 <- ttheme_default(core=list(fg_params=list(hjust=1, x=0.75,
                                               y= 0.7, cex = 0.5)),
                      rowhead=list(fg_params=list(hjust=0, x=0.75, 
                                                  cex = 0.5)))

grid.arrange(tableGrob(analyte_tests,
                       rows = rownames(analyte_tests$ANALYTE), 
                       cols = c("Analyte", "Empty", "Completed"),
                                         theme = tt2),
             left = "Analytes with Empty Entries",
             bottom = textGrob("EPA RADNet 1978-2017",
                               gp = gpar(fontface = 3, fontsize = 6),
                               hjust = 1.5,
                               x = 1))

#### by Material ID  ####
spread(rad_data_raw %>% group_by(MAT_ID, complete) %>%  summarise(count = n()), 
       complete, count) %>%  mutate(Incomplete = N/(N+Y))
#### by Procedure Number ####
spread(rad_data_raw %>% group_by(ANA_PROC_NUM, complete) %>%  summarise(count = n()),complete,
       count) %>% View()
```

#####Sample IDs - Samples and Tests
The number of unique sample IDs is 272,004, 54% of the database. If distinct is applied to additionally categories such as location, the number of samples does not change, thus sample IDs do not appear to be used across multiple locations. However, when grouped by both sample ID and result date fewer samples are duplicated, indicating that results for one Sample ID can be entered on different dates. If the date grouping is over a date range (e.g, 6 mos, 1 yrs, 2 yrs) the number of duplicate entries decreases. This limits the life-time of a Sample ID and indicates IDs are not recycled. A grouping by Sample ID and Analyte ID confirms that 98% of the data (494249) is unique within these two variables.  

```{r samp_ID_groups, echo=FALSE, message=FALSE, warning=FALSE}

## Grouped by SAMP-ID
#####################
print((c("Distinct Sample IDs:  ", 
              n_distinct(rad_data_raw$SAMP_ID))))
## analyses complete by Sample ID grouped by another variable
print(paste(c("Distinct Sample IDs by location:  ", 
              n_distinct(rad_data_raw$SAMP_ID, rad_data_raw$LOC_NUM))))
print(paste(c("Distinct Sample IDs by result date:  ",
              n_distinct(rad_data_raw$SAMP_ID, rad_data_raw$RESULT_DATE))))
print(paste(c("Distinct Sample IDs by Date Range (6 mos) and Location:  ",
              n_distinct(rad_data_raw$SAMP_ID, rad_data_raw$date_range))))
print(paste(c("Distinct Sample IDs by Date Range (1 year) and Location:  ",
              n_distinct(rad_data_raw$SAMP_ID, rad_data_raw$date_range1))))
print(paste(c("Distinct Sample IDs by Date Range (2 years) and Location:  ",
              n_distinct(rad_data_raw$SAMP_ID, rad_data_raw$date_range2))))
print(paste(c("Distinct Sample IDs by Analyte:  ",
              n_distinct(rad_data_raw$SAMP_ID, rad_data_raw$ANALYTE_ID))))
```

As noted, a sample can be tested for multiple analytes with different result dates. Each of the analytes would be a separate observation, but should have identical Sample IDs. To collect together all parts of a sample, the dataframe was grouped across 4 categories (material ID, date range (2yr), location, and sample ID).  This grouping was developed to determine a status for the sample: complete - all analytes have results, incomplete - some analytes have empty results, and empty - no results entered for any of the analytes for that sample ID. Using this status categorization there are 274677 samples, 93.5% are complete, 6.4% are incomplete, and 0.1% are empty. With the grouped data we can visualize the status of a sample by the number of tests (entries). Generally completed samples have only 1 test, those that are empty have about 5 tests while incomplete samples have a range of entered tests, mostly from 4 to 10.

```{r in_complete, echo=FALSE, message=FALSE, warning=FALSE}
## Grouped by MAT_ID, **2-year date range**, location, sample id
samples_rank <- rad_data_raw %>%
  group_by(MAT_ID, date_range2, LOC_NUM, lat, lon, SAMP_ID) %>% 
  summarise(Ytot = sum(Y), Ntot = sum(N))  %>% 
  mutate(analyses = Ytot + Ntot, pI = Ntot/analyses, pC = Ytot/analyses)

#### Notation:  Empty (no analyses, all RESULT_AMOUNT = NA), ####
##   Complete (no empty results),  Incomplete (mix of complete and empty)
samples_rank$status <- ifelse(samples_rank$Ytot == 0, "E",
                              ifelse(samples_rank$Ntot == 0, "C",
                                           "I"))
table(samples_rank$status)

#### Status - density violin plot with percent Complete, Empty, Incomplete ####

tests_complete_status <- ggplot(samples_rank, 
                                  aes(x=status, y = analyses)) +
  geom_violin(aes(fill = status)) +
  scale_fill_manual(values = c( "green", "red", "blue"),
                      labels = c("(C)omplete", "(E)mpty", "(I)ncomplete")) +
  #facet_wrap(~MAT_ID) +
  labs(title = "Status of Samples",
       caption = "EPA RadNet Data 1978-2017",
       y = "Number of Tests") +
  theme(plot.title = element_text(hjust=0.5),
        plot.caption = element_text(size = rel(0.5)),
        axis.title.x = element_blank())
tests_complete_status
```

#####Result Values by Sample Type
While it was noted that result values themselves cannot be compared directly without ensuring the same unit of measure, we can group like samples and plot values filtered by a unit. Filtering for the chief analyte *Beta* and pCi/m^3^ within the Hawaiian dataset a plot by date yields a two sharp increases above background. The dates of these events are just after the nuclear accidents at the Chernobyl and Fukushima nuclear power plants in May 1986 and March 2011. 

The entire data set was then evaluated in the same fashion and the spikes for Fukushima and Chernobyl are a bit changed. Beta result values in other locations reversed the relative heights of the peaks. In addition, two new spikes emerge, one in March 1981 and another in July 2008. 

```{r beta_date, echo=FALSE,message=FALSE, warning=FALSE}
### HI Beta Results (pCi/m3) by Date, coloured for location
#######################################################################
HI_beta <- ggplot(subset(HI_data, !is.na(RESULT_AMOUNT) &
                           RESULT_UNIT == "PCI/M3" & 
                           ANALYTE_ID =="BETA"),
                  aes(x=RESULT_DATE, y= RESULT_AMOUNT #, color = CITY_NAME
                      )) + 
  geom_point() +
  annotate("text", label = "0.894 Mar '11 \n (Fukushima)",
           x= as.Date("2011-03-25"),
           y=0.8) +
  annotate("text", label = "0.3758 May '86 \n (Chernobyl)",
           x= as.Date("1986-05-18"),
           y=0.5) +
  labs(title = 'Gross Beta in HI Samples 1978-2017', 
       caption = 'EPA RadNet Data') +
  scale_x_date(date_breaks = "5 years", labels = date_format("%Y")) +
  xlab('Date') +
  ylab('pCi/m3') +
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_text(hjust=0.5),
        legend.title = element_blank())
HI_beta

## Test Filter date to find High Points in above chart
#test <- HI_data %>% filter(RESULT_AMOUNT > 0.25 & 
#                                  ANALYTE_ID == "BETA" &
#                                  RESULT_UNIT == "PCI/M3" &
#                                  RESULT_DATE < "2011-01-01") %>% 
#                         select(RESULT_AMOUNT, RESULT_DATE)

### Plot of Beta Results (pCi/m3) all locations
#####################################

all_beta <- ggplot(subset(rad_data,
              RESULT_UNIT == "PCI/M3" &
              ANALYTE_ID =="BETA"),
       aes(x=RESULT_DATE, y= RESULT_AMOUNT #, color = STATE_ABBR
           )) + 
  geom_point() +
  annotate("text", label = "1.158 Mar '11 AK \n (Fukushima Mar '11)",
              x= as.Date("2011-03-25"),
              y= 1.9) +
  annotate("text", label = "0.703 \nJuly '08 TN",
              x= as.Date("2008-07-18"),
              y= 0.85) +
  annotate("text", label = "2.543 May '86 AL\n (Chernobyl Apr '86)",
              x= as.Date("1986-05-23"),
              y= 2.75) +
    annotate("text", label = "1.104 \nMar '81 NV \n (Tsuraga? \n Mar '81)",
              x= as.Date("1983-03-04"), #moved for display, point at 1981-03-04
              y= 1.25) +
  labs(title = 'Gross Beta in RADNet Samples 1978-2017', 
       caption = 'EPA RadNet Data') +
  scale_x_date(date_breaks = "5 years", labels = date_format("%Y")) +
  xlab('Date') +
  ylab('pCi/m3') +
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_text(hjust=0.5),
        legend.title = element_blank(),
        legend.key.size = unit(2,"mm"),
        legend.text = element_text(size = rel(0.7))
        ) +
  guides(col= guide_legend(ncol = 2))
all_beta

## Filter date to find High Points in above chart
#test <- rad_data_raw %>% filter(RESULT_AMOUNT > 1.5 & 
#                                  ANALYTE_ID == "BETA" &
#                                  RESULT_UNIT == "PCI/M3" &
#                                  RESULT_DATE < "1987-01-01") %>% 
#                         select(RESULT_AMOUNT, RESULT_DATE, STATE_ABBR, CITY_NAME)

```

Other analytes can be examined in the same way. Radium-226, the 10th most common analyte, is most often reported in picocuries per liter (pCi/L). Filtering for this subset and plotting yields a much less populated graph with consistent low values until 2012 when noisy and negative values are reported. Because of the noise in this graph further study and examination of detection limits would be required to determine the signifcance of the peak value >25 pCi/L.

```{r radium_date, echo=FALSE, message=FALSE, warning=FALSE}
rad_data_raw %>% filter(ANALYTE_ID == "RA226") %>% group_by(RESULT_UNIT) %>% summarise(n=n())
radium <- subset(rad_data, RESULT_UNIT == "PCI/L" & ANALYTE_ID =="RA226")

all_radium <- ggplot(radium, aes(x=RESULT_DATE, y= RESULT_AMOUNT)) + 
  geom_point() +
  geom_line(aes(x=RESULT_DATE, y= MDC), lty = 3, colour = "blue")+
  labs(title = 'Radium-226 in RADNet Samples 1978-2017', 
       caption = 'EPA RadNet Data') +
  scale_x_date(date_breaks = "5 years", labels = date_format("%Y")) +
  xlab('Date') + ylab('pCi/L') +
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_text(hjust=0.5),
        legend.title = element_blank(),
        legend.position = c(0.3,0.5))
all_radium
```

###Bivariate Analysis and Review
####Relationships Observed in bivariate investigations
The material ID appears to be the most convenient way to quickly subset and classify the observations. As seen in the univariate analysis Air-Filter dominates and the correlation matrix confirmed that Air-Filter samples coincide with the top categorical results: result unit *pCi/m^3^*, analytical procedure *1*, type *R*adioactive. Air-Filters aslo comprise the most complete entries, *Y*.

The location data, while informative and visually interesting, does not easily correlate to the other variables in the data. Yes, more observations are noted in areas with more monitoring stations. Not surprisingly, there are more monotoring stations located in areas of higher population and locations near known radioactive material handling, e.g. Oak Ridge, TN; Carlsbad, NM; Richland, WA. Surprisingly, the incomplete entries do not have the same correlation to location, these entries do not seem to be as simple as a percentage of the total.

In fact empty entries have a strong correlation to Air-Charcoal entries which are predominantly incomplete. Milk and precipitation samples have more incomplete entries percentage-wise than the remaining material types. 

####Interesting relationships between minor features
The *E*lement category small subset of the data and are not generally split in any other categories. Element observations are most simply *completed drinking water samples* reported in *g/L*. The element entries are split by analytical procedure with most being analysed by procedure 9 with some analysed by procedure 57. On this limited set it was easy to tie procedures 9 and 57 to the analytes potassium (K) and calcium (Ca), respectively. 

A somewhat surprising result is the weak correlation of the analytical size to the result amount values. The result value should be independent of the sample size. However, it is important to remember that the results and size are not standardized to for actual comparison, this correlation it is purely a relationship between the values not actual magnitude. While sample sizes are often specified by procedure, it should not be expected that the result values would be related to the size of the sample. In this case, I would theorize that the the relationship is forced due to limited analyses in which the result values and sample sizes possess a low variance. 

####The strongest relationship
The strongest correlations existing in the original dataframe is between the result amounts, minimal detectable concentration and combined standard uncertainty values. This is to be anticipated as both MDC and CSU are determined/calculated through the analyses of samples and results. A strong negative (-1) correlation between the Y/N status variables confirms that these variables were set up correctly as if the sample is complete (1) it cannot also be incomplete (0) and vice versa.

##Multivariate Investigations

###Multivariate Plots Section

It is easy to facet and fill on earlier plots by some of the variables to give more relational information between the variables. When the simple histogram is facted by material ID, air-filter entries dominate and as anticipated an analyte ID fill yields gross beta as the predominant analyte. As plotted by year, we notice that pasteurized milk and precipitation were analysed more often in the 1980s-1990s, but appear to have been discontinued/deprioritized whereas the number of air-filter samples has picked up in the last 20 years. 

With one simple change the distribution of samples over the various monitoring stations is plotted. When using the numerical ID the distribution looks fairly uneven. However, the distribution is more even if region is plotted, so the right skewed assignment of location IDs yields a skewed visual of geographic distribution.

```{r analytes_facet, echo=FALSE, message=FALSE, warning=FALSE}

ggplot(rad_data, aes(RESULT_DATE, fill = ANALYTE_ID)) + 
  geom_histogram() +
  facet_wrap(~rad_data$MAT_ID) 

ggplot(rad_data, aes(LOC_NUM, fill = ANALYTE_ID)) + 
  geom_bar(stat="count") +
  facet_wrap(~rad_data$MAT_ID)
summary(rad_data$LOC_NUM) #skewed LOC_NUM ID assignments 

ggplot(rad_data, aes(region, fill = ANALYTE_ID)) + 
  geom_bar(stat="count") +
  facet_wrap(~rad_data$MAT_ID) 

```

If we focus only on the second tier analytes, the histogram also shows that these less common analytes are also sampled with fairly regular distribution across the monitoring sites. 

```{r otheranalytes_geo, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(subset(rad_data, ANALYTE_ID %in%
                         c("CS137","H3","K40","BA140","I131")), 
              aes(region, fill = ANALYTE_ID)) + 
  geom_bar() +
  scale_y_sqrt() +
  labs(title = "",  
       x = "LOCATION NUMBER", y = "COUNT")
```

####Result Amounts and Locations - Beta Analyses 
Using the beta analyes alone we can facet by location to get a bigger picture view of the release events noted earlie. Within the Hawaiian data set the locations are limited to Hawaiian cities and it is easy to see that the Chernobyl event was picked up at the Honolulu station(s), which appears to be the only station monotoring at that time, but for Fukushima Kahuku and Kauai both saw spikes in gross beta.

Presenting the entire dataset in the same manner we can elucidate that the July 2008 appears localized to Tennessee, while the March 1981 incident was picked up at several monitoring stations.

Research did not yield any information on any nuclear incidents during 2008.  The 1981 spike was entered March 3 by a Nevada station however, the only nuclear accident found during this time was at the Tusraga Power Plant[^tsuruga]. Reports have this incident as being on March 8th[^tsuruga8] or 9th.[^tsuruga9], however, the reports were covered up for some 40 days and earlier issues at the power plant went unreported, perhaps these are related. 

[^tsuruga]: https://en.wikipedia.org/wiki/Tsuruga_Nuclear_Power_Plant
[^tsuruga8]: http://timshorrock.com/wp-content/uploads/Chronology-of-1981-Tsuruga-Accident-from-Japanese-Press.pdf
[^tsuruga9]: https://www.history.com/this-day-in-history/japanese-power-plant-leaks-radioactive-waste

```{r beta_regions, echo=FALSE}
#### HI Beta Results (pCi/m3) by Date, Coloured for location #######
HI_beta <- ggplot(subset(HI_data, !is.na(RESULT_AMOUNT) &
                           RESULT_UNIT == "PCI/M3" & 
                           ANALYTE_ID =="BETA"),
                  aes(x=RESULT_DATE, y= RESULT_AMOUNT, color = CITY_NAME)) + 
  geom_point(alpha=0.5) +
  annotate("text", label = "0.894 Mar '11 \n (Fukushima)",
           x= as.Date("2011-03-25"),
           y=0.8) +
  annotate("text", label = "0.3758 May '86 \n (Chernobyl)",
           x= as.Date("1986-05-18"),
           y=0.5) +
  labs(title = 'Gross Beta in HI Samples 1978-2017', 
       caption = 'EPA RadNet Data') +
  scale_x_date(date_breaks = "5 years", labels = date_format("%Y")) +
  xlab('Date') +
  ylab('pCi/m3') +
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_text(hjust=0.5),
        legend.title = element_blank())
HI_beta

####Plot of Beta Results Everywhere (pCi/m3), coloured by state####
all_beta <- ggplot(subset(rad_data,
              RESULT_UNIT == "PCI/M3" &
              ANALYTE_ID =="BETA"),
       aes(x=RESULT_DATE, y= RESULT_AMOUNT, color = STATE_ABBR)) + 
  geom_point(alpha = 0.6) +
  annotate("text", label = "1.158 Mar '11 AK \n (Fukushima Mar '11)",
              x= as.Date("2011-03-25"),
              y= 1.9) +
  annotate("text", label = "0.703 \nJuly '08 TN",
              x= as.Date("2008-07-18"),
              y= 0.9) +
  annotate("text", label = "2.543 May '86 AL\n (Chernobyl Apr '86)",
              x= as.Date("1986-05-23"),
              y= 2.75) +
    annotate("text", label = "Mar '81\n1.104 - NV \n (Tsuraga?)",
              x= as.Date("1983-03-01"), #moved for display, point at 1981-03-04
              y= 1.5) +
  labs(title = 'Gross Beta in RADNet Samples 1978-2017', 
       caption = 'EPA RadNet Data') +
  scale_x_date(date_breaks = "5 years", labels = date_format("%Y")) +
  xlab('Date') +
  ylab('pCi/m3') +
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_text(hjust=0.5),
        legend.title = element_blank(),
        legend.key.size = unit(2,"mm"),
        legend.text = element_text(size = rel(0.7))
        ) +
  guides(col= guide_legend(ncol = 2))
all_beta
```

To focus on the Fukushima incident, data was filtered and grouped for the month after the incident (March to April 11, 2011). The summarized data shows that max beta measurements across the US are definitely higher on the west coast during this month versus measurements made in the eastern continental US. The same data in the previous year was also summarized and those values are less than a tenth of the 2011 values. Additionally, the pattern across the US is different with *maximum* values being higher across the northern border instead of the coast.

```{r fukushima, echo=FALSE, message=FALSE, warning=FALSE}
#### FUKUSHIMA RESULTS IN-DEPTH March 11, 2011####
fukushima <- rad_data %>% 
  filter(RESULT_DATE >= "2011-03-11" & RESULT_DATE <= "2011-04-11" &  
         ANALYTE_ID == "BETA"  &
         RESULT_UNIT == "PCI/M3") %>% 
  group_by(region) %>% 
  summarise(mean = mean(RESULT_AMOUNT), n = n(), 
            value = max(RESULT_AMOUNT))

#### Choropleth of Beta Air Results March 2011####
#show_col(brewer_pal(palette = "PuBu")(7))
col.pal<-brewer_pal(palette = "PuRd")(7)

choro_fukushima1<-StateChoropleth$new(fukushima)
choro_fukushima1$ggplot_scale <- scale_fill_manual(name="pCi/m3",
                                         values=col.pal, drop=FALSE)
choro_fukushima1$show_labels = FALSE
choro_F1 <- choro_fukushima1$render()  +
  theme(plot.caption = element_text(size = rel(0.5)),
         plot.title = element_blank(),
         plot.subtitle = element_text(size=rel(.75), hjust = 0.5),
         legend.key.size = unit(2.5,"mm")) + 
  labs(subtitle = 'After Fukushima March 11 - April 11 2011', caption = 'EPA RadNet Data')  


#### Baseline March-April 2010####
baseline <- rad_data %>% 
  filter(RESULT_DATE >= "2010-03-11" & RESULT_DATE <= "2010-04-11" &  
         ANALYTE_ID == "BETA"  &
         RESULT_UNIT == "PCI/M3") %>% 
  group_by(region) %>% 
  summarise(mean = mean(RESULT_AMOUNT), n = n(), 
            value = max(RESULT_AMOUNT))

col.pal<-brewer_pal(palette = "PuBu")(7)
choro_grundlinie<-StateChoropleth$new(baseline)
choro_grundlinie$title <- "Baseline 2010"
choro_grundlinie$ggplot_scale <- scale_fill_manual(name="pCi/m3",
                                         values=col.pal, drop=FALSE)
choro_grundlinie$show_labels = FALSE
  
choro_grund <- choro_grundlinie$render()  +
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_blank(),
        plot.subtitle = element_text(size=rel(.75), hjust = 0.5),
        legend.key.size = unit(2.5,"mm")) + 
  labs(subtitle = 'Baseline March 11 - April 11 2010', caption = 'EPA RadNet Data') 

grid.arrange(choro_F1, choro_grund, nrow = 2,
             top = "Max Beta Activity in Air-Samples")

```

Localized results are seen by plotting the maximum Beta results by monitoring station. There are not many measurements for other radionuclides, but plot of the other analytes (pCi/m^3^) shows areas up to 37 pCi/m^3^. 

```{r fukushima2, echo=FALSE, message=FALSE, warning=FALSE}
#### Fukushima Beta max by monitoring station###################
fukushima_points <- rad_data %>% 
  filter(RESULT_DATE >= "2011-03-11" & RESULT_DATE <= "2011-04-11" &  
         ANALYTE_ID == "BETA" & 
         RESULT_UNIT == "PCI/M3") %>% 
  group_by(region,citystate) %>% 
  summarise(mean = mean(RESULT_AMOUNT), n = n(), 
            value = max(RESULT_AMOUNT), lat = mean(lat), lon = mean(lon))

#### Google Map Fukushima Beta max by monitoring station###################
monitors_fukushima <- USAMap + scale_y_continuous(limit = c(8.5,49.5)) +
  scale_x_continuous(limit = c(-130,-62)) +
  geom_point(data = fukushima_points,
             aes(x= lon, y= lat, size = value),
             col = 'red4') +
  labs(title = "Beta Activity",
       subtitle = 'March 11 - April 11 2011', caption = 'EPA RadNet Data',
       size = 'pCi/L')+
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_text(hjust=0.5),
        plot.subtitle = element_text(size=rel(.75), hjust = 0.5),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.08,0.18),
        legend.key.size = unit(2,"mm")
        )
monitors_fukushima

#### Fukushima Other Analytes max by monitoring station###################
fukushima_other <- rad_data %>% 
  filter(RESULT_DATE >= "2011-03-11" & RESULT_DATE <= "2011-04-11" &  
         ANALYTE_ID != "BETA" & 
         RESULT_UNIT == "PCI/M3") %>% 
  group_by(region, citystate, ANALYTE_ID) %>% 
  summarise(mean = mean(RESULT_AMOUNT), n = n(), 
            value = max(RESULT_AMOUNT), lat = mean(lat), lon = mean(lon))

#### Google Map Fukushima non-Beta max by monitoring station###################
monitors_fukushima_other <- USAMap + 
  scale_y_continuous(limit = c(8.5,49.5)) +
  scale_x_continuous(limit = c(-130,-62)) +
  geom_point(data = subset(fukushima_other, value > 1),
             aes(x= lon, y= lat, size = value, colour = ANALYTE_ID),
             position = position_jitter(width = 0.55, height = 0.55)) +
  labs(title = "Other Radionuclide Activity",
       subtitle = 'March 11 - April 11 2011', caption = 'EPA RadNet Data',
       size = 'pCi/L',
       colour = element_blank()) +
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_text(hjust=0.5),
        plot.subtitle = element_text(size=rel(.75), hjust = 0.5),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.08,0.25),
        legend.key.size = unit(2,"mm")
        ) +
  scale_size(limits = c(1,50), breaks= c(10,30,50))+
  scale_color_brewer(type = 'qual', palette = "Set1")

monitors_fukushima_other
grid.arrange(monitors_fukushima, monitors_fukushima_other, ncol = 2, nrow = 1,
             top = "Air Filter Samples After Fukushima")


```

The max beta activity measurement in Alaska after Fukushima were almost twice the US-contiguous response, 1.15 pCi/m^3^  vs.  0.6 pCi/m^3^ max. In Alaska other radionuclides are detected at even higher values, 144 pCi/m^3^ for Bismuth-212, 126 pCi/m^3^ for Lead-212 and 59 pCi/m^3^ for Thallium-208.
```{r fukushima_AK, echo=FALSE, message=FALSE, warning=FALSE}
AK_center = c(-149.4937, 64.20084)
AKMap <- ggmap(get_googlemap(center=AK_center,
                             zoom=4, extent="normal"))

AKMap_analytes <- AKMap + 
  geom_point(data = fukushima_other, 
             aes(x= lon, y= lat,color = ANALYTE_ID, size  = value),
             #alpha = 0.8,
             na.rm = TRUE, 
             position = position_jitter(width = 0.8, height = 0.75))+
   geom_point(data = fukushima_points, 
             aes(x= lon, y= lat, size  = value, colour ="beta"),
             #alpha = 0.8,
             na.rm = TRUE, 
             position = position_jitter(width = 0.8, height = 0.75))+
  labs(title=  'Alaska Radioactive Analyses after Fukushima',
       subtitle = 'March 11 - April 11, 2011',
       caption = 'EPA RadNet Data',
       size = "pCi/m3",
       colour = element_blank()) +
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_text(hjust=0.5),
        axis.title = element_blank(), 
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        plot.subtitle = element_text(size=rel(.75), hjust = 0.5),
        legend.box.just = "right",
        legend.key.size = unit(3, "mm")) +
  scale_size(range = c(0,6)) 

AKMap_analytes
```

####Sample Status/Incomplete Empties
Using a simple material ID facet shows that the individual empty observations which we know comprise 16% of the raw database and occured between 1990 and 2010 are mostly precipitation/milk samples with very few air-filter samples. However, when we group by sample ID to the 272004 distinct samples, only 285 samples (0.1%) are entirely empty while only 6% are incomplete. And a deep dive view of the empty samples reveals that they are predominantly drinking water samples between 1998-2008.

```{r complete_status, echo=FALSE, message=FALSE, warning=FALSE}
#### NA Result Amounts in all observations ####
nans_freq <- ggplot(nans_date_type, aes(x= Date_Range, y = Frequency, fill=Sample_Type)) +
  geom_bar(stat="identity") +
  labs(y="Frequency") +
  facet_wrap(~Complete, labeller = fertig_labeller) +
  theme(axis.text.x = element_text(angle = 30, size = rel(0.7)),
        axis.title.x = element_blank(),
        legend.title = element_blank())
nans_freq

#### Sample Status by date, MAT_ID Fill barplot ####
table(samples_rank$status)
sample_status <-ggplot(samples_rank,
                       aes(x=date_range2,fill= MAT_ID)) +
  geom_bar() +
    facet_grid(~samples_rank$status)+
  labs(title = "Sample Entry Status",
       caption = "EPA RadNet Data 1978-2017")+
  theme(plot.title = element_text(hjust=0.5),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 20),
        plot.caption = element_text(size = rel(0.5)),
        legend.title = element_blank(),
        legend.key.size = unit(3, "mm"),
        legend.text =  element_text(size=rel(0.7)),
        legend.position = c(0.15,0.80))
sample_status

#### Empty Samples by date, MAT_ID Fill barplot ####
sample_empty <-ggplot(subset(samples_rank, status =="E"),
                       aes(x=date_range2,fill= MAT_ID)) +
  geom_bar() +
  labs(title = "Empty Sample Entries",
       caption = "EPA RadNet Data 1978-2017")+
  theme(plot.title = element_text(hjust=0.5),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 20),
        plot.caption = element_text(size = rel(0.5)),
        legend.title = element_blank(),
        legend.key.size = unit(3, "mm"),
        legend.text =  element_text(size=rel(0.7)),
        legend.position = c(0.15,0.80))
sample_empty

#### Percent Sample Complete by Material ####
## gather percent to colour by pI/pC - complimentary values, doubles #Samples
samples_rank <- gather(samples_rank, "percent", n, c("pI", "pC"))

percent_complete_mat <- ggplot(samples_rank, 
                               aes(x= percent, y = n, fill = percent)) +
  geom_boxplot(notch = TRUE,
               alpha = 0.6,
               outlier.colour = 'brown',
               outlier.size = 1) +
  scale_fill_manual(values = c( "darkgreen", "darkorange"),
                    labels = c("Complete", "Incomplete")) +
  labs(title = "Percent Complete/Incomplete Samples by Material Type",
       caption = "EPA RadNet Data 1978-2017",
       y = "Percent") +
  facet_wrap(~MAT_ID) +
  theme(axis.title.x = element_blank(),
        plot.caption = element_text(size = rel(0.5)) )
percent_complete_mat 

```

A quick map of the contiguous US shows even though there are relatively few incomplete and empty samples they are spread across the monitoring stations. 

```{r emptyH2O_locs, echo=FALSE, message=FALSE, warning=FALSE}
##reload USA Map centered with Alaska
usa_center = c(-115,39)
USAMap = ggmap(get_googlemap(center=usa_center, scale=2, zoom=3, 
                             extent ="device"))
loc_empty_h20 <- USAMap + 
  scale_y_continuous(limit = c(8.5,65)) +
  scale_x_continuous(limit = c(-167,-63)) +
  geom_point(data = subset(samples_rank, MAT_ID == "DRINKING WATER" & status != "C"),
             aes(x= lon, y= lat, col = status, size = n),
             position = position_jitter(width = 0.7, height = 0.7)) +
  labs(title = "Drinking Water Samples - Incomplete or Empty",
       subtitle = '1990 = 2010', caption = 'EPA RadNet Data',
       size = 'percent',
       colour = element_blank()) +
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_text(hjust=0.5),
        plot.subtitle = element_text(size=rel(.75), hjust = 0.5),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.08,0.25),
        legend.key.size = unit(2,"mm")
        ) +
  scale_size(limits = c(0,1), range = c(1,2)) +
  scale_color_brewer(type = 'qual', palette = "Set1")
loc_empty_h20
```

###Multivariate Analysis and Review

Examining the relationships between multiple variables confirmed deductions made via the cursory univariate review. The majority observations between material ID and analyte are in fact correlated. One caution is to be aware of the data and its set up. The distribution of samples appeared to have some geographic importance when examined by location number. However, remember the univariate analysis showing the extreme right skewed assignment of location numbers led to a closer examination. And while some locations do have more observations than others the distribution is more even when looking by actual geographic locations or regions. 

####Interesting and Surprising Feature Interactions

The data observations without a result entry seems to have some interesting story and appeared at first to be a fairly significant portion of the database. A major step in this investigation led to the grouping of the observations by sample which revealed that most of the empty observations were not duplicates just multiple analyte tests on the same sample. So there was some information, but just incomplete. Largely, these samples are not localized to any site with the exception of the air-charcoal samples in Hawaii. 
Analyte tests in a sampling plan or were they previously tested and phased out  
If these are optional analyte tests it is better to consider the missing results in context of the entire sample. Is the sample completley analyzed, that is all assigned analytes have results, are they incomplete having some results, or are they empty with no results? 

------

##Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One

```{r Plot_One, echo=FALSE, message=FALSE, warning=FALSE }
#### USA plot with Monitoring Stations Sized by #Observations, Colored by #Stations
city_obs_locs <- city_obs_locs %>% 
  left_join(select(places, lat, lon, CITY, STATE),
            by = c("CITY_NAME" = "CITY",
                   "STATE_ABBR" = "STATE") )
city_obs_locs$locations <- as.factor(city_obs_locs$locations)
city_obs_locs <- city_obs_locs[order(city_obs_locs$locations, decreasing = F),]

radnet_overview <- USAMap + 
  scale_y_continuous(limit = c(8.5,65)) +
  scale_x_continuous(limit = c(-167,-63)) +
  # scale_y_continuous(limit = c(8.5,49.5)) +
  # scale_x_continuous(limit = c(-130,-62)) +
  geom_point(data = city_obs_locs,
             aes(x= lon, y= lat, size = observations, 
                 colour = locations)) +
             #position = position_jitter(width = 0.55, height = 0.55)) +
  labs(title = "RadNet Station Database Entries",
       subtitle = '1978-2017', caption = 'EPA RadNet Data',
       size = 'Entries',
       colour = 'Monitoring Stations') +
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_text(hjust=0.5),
        plot.subtitle = element_text(size=rel(.75), hjust = 0.5),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.15,0.25),
        legend.key.size = unit(4,"mm"),
        legend.box.background = element_rect(),
        legend.spacing = unit(0,"mm")
        ) +
  #scale_size(limits = c(1,50), breaks= c(10,30,50))+
  scale_color_brewer(type = 'seq', palette = "PRGn")

radnet_overview

```

### Description One
This map provides a nice simple synopsis of the RadNet Network from 1978 to 2017. North American cities with monitoring stations are marked with a colored dot indicating the number of stations in that city. Additionally, the size of the dot conveys the total number of entries into the database for the city. It is difficult to express this much information in a clear way, and while this map leaves out a number of monitoring stations in the Pacific Ocean it give a nice visual overview. 

### Plot Two
####Results by date with annotation
```{r Plot_Two, echo=FALSE, message=FALSE, warning=FALSE }
rad_data_raw$RESULT_DATE <- as.Date(rad_data_raw$RESULT_DATE, "%F")

#### DATE SPIKES ####
analysis_counts <- c(510,565,480,490,1025,760,325,312 )
analysis_dates <- c("1996-07-01","1997-07-01", "1998-07-01","1999-07-01",
                    "2009-12-31", "2010-12-31","2011-12-31", "2012-12-31")
label_df <- data.frame(analysis_counts, analysis_dates)
label_df$analysis_dates <- as.Date(label_df$analysis_dates)

date_spikes <- ggplot(rad_data_raw, aes(RESULT_DATE)) + 
  geom_line(color = "steelblue ", stat = "count") +
  scale_x_date(date_breaks = "5 year", date_minor_breaks = "year")+
  labs(title="Number of Analyses Results by Date", x=NULL, y = "Count") +
  geom_point(data = label_df, aes(x=label_df$analysis_dates,
                                  y=label_df$analysis_counts), color = "red")+
  theme(axis.text.x = element_text(angle=30),
        plot.title = element_text(hjust = 0.5))
date_spikes 
```

### Description Two


### Plot Three
####Status Faceted by Material Type
```{r Plot_Three, echo=FALSE, message=FALSE, warning=FALSE }
## Violin Status by Mat_ID and Number of Analyses
tests_complete_status <- ggplot(samples_rank, 
                                  aes(x=status, y = analyses)) +
  geom_violin(aes(fill = status)) +
  scale_fill_manual(values = c( "green", "red", "blue"),
                      labels = c("(C)omplete", "(E)mpty", "(I)ncomplete")) +
  facet_wrap(~MAT_ID) +
  labs(title = "Status of Sample Testing by Material Type",
       caption = "EPA RadNet Data 1978-2017",
       y = "Number of Analyses") +
  theme(plot.title = element_text(hjust=0.5),
        plot.caption = element_text(size = rel(0.5)),
        axis.title.x = element_blank())
tests_complete_status
```

### Description Three
These double-faceted bar-plots are clean elegant way to clarify the status of samples within the database and provide easy classification possibilities. At the top level the samples are identified either as *complete* (all entries with the same sample ID have result), *empty* (all entries for the sample ID are NA) or *incomplete* (mixed result values with NA's). The separation by material type allows an easy comparison between these high level categories of sample types. Finally, by using a violin bar-plot a density function of the number of analyses (entries) per sample is readily illustrated. 
------

#Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

In attempting to map the monitoring locations a few issues were found with the data. For example, the abbreviation *PC* was not recognized by the API and was changed to the full Panama Canal; the city of Doswell, SC does not appear to exist, but several Doswell, VA observations were found in the dataframe so the single entry was changed from SC to VA. Additionally several entries were listed only as EPA Regions (1-10). In an effort to geo-locate the data, the region number was updated to the location of the headquarters for that EPA region.[^eparegions]. Also, rather than inundate the API with multiple requests for the same location a separate dataframe was created listing only unique city-state combinations. This removed the instances where a city has multiple monitoring stations (LOC_NUM). 


In order to use the function `state_choropleth{choroplethr}`, the dataframe must contain a column named *region* matching the naming convention found for *region* in `state.map`. The convention for US state names is the full state name in lower case. The values to be mapped must be contained in column named *value*. Several data sets include data related to the USA and using the charactor vectors `state.abb` and `state.name` it was simple to make a dataframe which could be joined as an appropriately formatted *region* column using state abbreviation column already contained in the EPA RadNet dataset. However, as the state data does not include the territories several regions were added manually to complete complete the fields.






#Miscellany to remove as needed


```{r HI_facets, eval= FALSE, include=FALSE,echo=FALSE, message=FALSE, warning=FALS}
#####################################  MULTIVARIATE Faceting HI ??
##filter beta data
HI_beta <- subset(HI_data, !is.na(RESULT_AMOUNT) &
                           RESULT_UNIT == "PCI/M3" & 
                           ANALYTE_ID =="BETA")

summary(HI_data$RESULT_UNIT)
### Counts of Hawaiian results by year, faceted by City ~ Material Type

HI_samples <- ggplot(HI_beta, 
                     aes(x=RESULT_DATE, y= RESULT_AMOUNT, color = CITY_NAME))+
  geom_point() +
  #facet_grid(~HI_beta$MAT_ID) +
  labs(title = 'RAD Analyses in HI Cities by Sample Type', 
       caption = 'EPA RadNet Data', cex.main = 0.50,
       fill = 'Material Type') +
  # xlab('Date') +
  # ylab('pCi/L') +
  theme(plot.title = element_text(hjust = 0.5),
        plot.caption = element_text(size = rel(0.5)),
        axis.text.x = element_text(size=7, angle=45),
        axis.title.x = element_blank(),
        axis.title.x.top = element_text(size = 6),
        legend.position = "none"
  )
HI_samples

HI_analytes <- ggplot(HI_data, aes(RESULT_DATE, fill = ANALYTE_ID)) + 
  geom_histogram() +
  facet_wrap(~HI_data$MAT_ID) +
  labs(title = 'HI Analytes by Sample Type', 
       caption = 'EPA RadNet Data', cex.main = 0.50,
       fill = 'Analyte Type') +
  xlab('Date') +
  ylab('count') +
    theme(plot.title = element_text(hjust = 0.5),
        plot.caption = element_text(size = rel(0.5)),
        axis.text.x = element_text(size=7, angle=45),
        axis.title.x = element_blank(),
        axis.title.x.top = element_text(size = 6),
        legend.position = "bottom")
HI_analytes


  labs(title = 'HI Analytes by Sample Type', 
       caption = 'EPA RadNet Data', cex.main = 0.50,
       fill = 'Analyte Type') +
  xlab('Date') +
  ylab('count') +
    theme(plot.title = element_text(hjust = 0.5),
        plot.caption = element_text(size = rel(0.5)),
        axis.text.x = element_text(size=7, angle=45),
        axis.title.x = element_blank(),
        axis.title.x.top = element_text(size = 6),
        legend.position = "bottom")
HI_analytes
```

```{r csu_mdc_lambda, eval= FALSE, include=FALSE,echo=FALSE, message=FALSE, warning=FALSE}
##Univariate Analyses of limited variables 
##CSU Combined Standard Uncertainty (95% = 2xCSU)
summary(rad_data$CSU)
ggplot(subset(rad_data, RESULT_UNIT == "PCI/L"), 
       aes(x=RESULT_DATE, y= CSU, color = ANALYTE_ID)) +
  geom_point() +
  labs(title = "CSU by year",
       y="pCi/L", x =  "YEAR")

##MDC Minimum Detectable Concentration
summary(rad_data$MDC)
ggplot(rad_data, aes(x=RESULT_DATE, y= MDC, color = ANALYTE_ID)) +
  geom_point() +
  labs(title = "MDC by year",
       y="all units", x =  "YEAR")

ggplot(subset(rad_data, RESULT_UNIT == "PCI/L"), 
       aes(x=RESULT_DATE, y= MDC, color = ANALYTE_ID)) +
  geom_point() +
  labs(title = "MDC (pCi/L) by year",
       y="pCi/L", x =  "YEAR")


## Half-Life
summary(rad_data$HALF_LIFE)
count(rad_data, HALF_LIFE_TIME_UNIT)

ggplot(rad_data, aes(x=RESULT_DATE, y= HALF_LIFE, color = ANALYTE_ID)) +
  geom_point() +
  labs(title = "Half-Lives of Analytes",
       y="all units", x =  "YEAR")
ggplot(subset(rad_data, HALF_LIFE_TIME_UNIT =="Y"), aes(x=RESULT_DATE, y= HALF_LIFE, color = ANALYTE_ID)) +
  geom_point() +
  labs(title = "Half-Lives of Analytes",
       y="Half-life Years", x =  "YEAR")

```

```{r milk_k40, eval= FALSE, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
### Random Investigations
#####################################

pastuerized_milk <- ggplot(subset(rad_data, MAT_ID == "PASTEURIZED MILK"),
                           aes(x=RESULT_DATE, y= RESULT_AMOUNT, color = RESULT_UNIT)) + 
  geom_point(alpha = 0.5) +
  labs(title = 'Pasteurized Milk Samples 1978-2017', 
       caption = 'EPA RadNet Data') +
  scale_x_date(date_breaks = "5 years", labels = date_format("%Y")) +
  xlab('Date') +
  ylab('All Units') +
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_text(hjust=0.5),
        legend.title = element_blank(),
        axis.text.x = element_text(angle=30))

count(subset(rad_data, ANALYTE_ID == "K40"))

potassium_air <- ggplot(subset(rad_data, 
                              (RESULT_UNIT == "PCI/M3" |
                                            RESULT_UNIT == "ACI/M3") &
                              (ANALYTE_ID == "K" | ANALYTE_ID == "K40")),
                         aes(x=RESULT_DATE, y= RESULT_AMOUNT, color = RESULT_UNIT)) + 
  geom_point() +
  labs(title = 'K40', 
       caption = 'EPA RadNet Data') +
  scale_x_date(date_breaks = "1 years", labels = date_format("%Y")) +
  xlab('Date') +
  ylab('aCi/m3 or pCi/m3') +
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_text(hjust=0.5),
        legend.title = element_blank(),
        axis.text.x = element_text(angle=30))
potassium_air


detection_limit_k40 <- rad_data %>% 
  filter(ANALYTE_ID == "K40" &
         (RESULT_UNIT == "PCI/M3" | RESULT_UNIT == "ACI/M3")) %>% 
  group_by(citystate) %>% 
  summarise(mean = mean(RESULT_AMOUNT), n = n(), 
            value = max(RESULT_AMOUNT), lat = mean(lat), lon = mean(lon))
  
```

```{r beta_airfilter, eval= FALSE, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE }

### Group by BETA
#########################
beta_data <- subset(rad_data, ANALYTE_ID == "BETA")

airbeta <- subset(beta_data, MAT_ID == "AIR-FILTER" & !is.na(CSU) &
                 RESULT_UNIT == "PCI/M3")

ggplot(airbeta) +
  geom_point(aes(x=RESULT_DATE, y= RESULT_AMOUNT), colour = "red4", size =1) +
  #geom_errorbar(aes(ymin= airbeta$RESULT_AMOUNT - airbeta$CSU, 
  #                  ymax= airbeta$RESULT_AMOUNT + airbeta$CSU),
  #             colour="black", width=.1) +
  labs(title = 'BETA Results pCi/L, Air-Filter Samples', 
       caption = 'EPA RadNet Data') +
  scale_x_date(date_breaks = "5 years", labels = date_format("%Y")) +
  scale_y_continuous(limits = c(-.05,.25)) +
  xlab('Date') +
  ylab('pCi/M3') +
  geom_errorbar(aes(x=RESULT_DATE, y= RESULT_AMOUNT + CSU), colour = "yellow")+ 
  #geom_line(aes(x=RESULT_DATE, y= MDC), colour = "green")+ 
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_text(hjust=0.5),
        legend.title = element_blank(),
        axis.text.x = element_text(angle=30))


ggplot(subset(beta_data, RESULT_DATE > "2010-12-31"),
       aes(x=RESULT_DATE, y= RESULT_AMOUNT)) +
  geom_errorbar(aes(ymin = RESULT_AMOUNT - CSU, 
                    ymax = RESULT_AMOUNT + CSU),
                    colour="black", width=.1) +
  geom_line() +
  geom_point()
```
