---
title: "RadNet Exploratory Data Analysis with RStudio"
output:
  pdf_document: default
  html_notebook: default
  editor_options: 
    chunk_output_type: console
---
*by Wendy Bisset*
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages in this code chunk
# Parameter "echo" = FALSE prevents the code from displaying in the knitted #HTML output. 
# Set echo=FALSE for all code chunks in file, unless it makes sense to show #the code that generated a particular plot.
# Parameters for "message" and "warning" should also be set to FALSE
# for code chunks you have verified 

#install.packages("ggthemes")
#install.packages(c("choroplethr","choroplethrMaps"))

library(ggplot2)
library(dplyr)
library(scales)
library(ggmap)
library(ggthemes)
library(choroplethr)
library(choroplethrMaps)
library(grid)
library(gridExtra)
## citation('ggmap')
##To cite ggmap in publications, please use:
##[^ggmap]:D. Kahle and H. Wickham. ggmap: Spatial Visualization with ggplot2. The R Journal, ##5(1), 144-161. URL http://journal.r-project.org/archive/2013-1/kahle-wickham.pdf
```

#Data Source and Background Information

The US Environmental Protection Agency maintains several public databases on environmental data.[^EPA]:<https://www.epa.gov/enviro/envirofacts-overview> One such database, ***RadNet***,[^RadNetOverview]:<https://www.epa.gov/enviro/radnet-overview> is a system of geographically distributed stations which sample and test for a number of radiochemical analytes (e.g., iodine-131, gross beta (β)) in the nation's air, precipitation, and drinking water. These stations are located across the US as well as American Territories. The current RadNet database primarily consists of data collected since 1978 though some data dates back to 1973 from RadNet's precursor, ERAMS (Environmental Radiation Ambient Monitoring System).

While the stations of RadNet normally sample on a regular schedule, if there is concern for a significant radiation release, the frequency of sampling rate can be increased. The site mentions five instances in which  sampling rates were increased: 1) 1979 Three Mile Island nuclear reactor accident in the U.S., 2) 1986 Chernobyl nuclear reactor accident in the Soviet Union, 3) 1999 Tokaimura nuclear fuel processing facility accident in Japan, 4) 2000 wildfires in Los Alamos and Hanford U.S., and 5) 2001 terrorist attacks in the U.S.

RadNet helps provide historical data needed to estimate long-term trends in environmental radiation levels and also provides a means to estimate levels of radioactivity in the environment. This data helps establish background radiation as well as identify radioactive fallout from atomic weapons testing, nuclear accidents, or other intrusions of radioactive materials. There is a timeline on the RadNet website[^RadNetTimeline]:<https://www.epa.gov/radnet/history-radnet> which lists changes in the monitoring system and events that may have impacted the measurements.  

#Downloading, Importing and Merging Data Files

All data was downloaded directly from the [RadNet](https://www.epa.gov/enviro/radnet-customized-search) website as .csv files. According to the EPA's [Envirofacts Data Service API](https://www.epa.gov/enviro/web-services) data output is limited to 10000 rows of data at a time from a maximum of three tables. Consequently, the RadNet data for each media type was downloaded in chunks by decade (up-to 1989, 1990-1999, 2000-2009, 2010-2017).  The resulting .csv files were loaded into RStudio and then [merged](https://psychwire.wordpress.com/2011/06/03/merge-all-files-in-a-directory-using-r-into-a-single-dataframe/) into a single R dataframe.

In total 19 variables were downloaded from a combination of three RadNet Database Tables. Data obtained from each table is broadly outlined below: 

1. Media
   + Drinking Water (DW)
   + Surface Water (SW) program  (terminated March 1999)
   + Precipitation
   + Milk (program terminated 2014)
   + Air Data combined from: 
     - Air-Charcoal (data only found in 2010-2017)
     - Air-Filter
2. Sampling Location
   + 324 Location IDs across US and Territories
   + ID number
   + city
   + state
3. Types of Radionuclides and Radiation (61 analytes (e.g., )
   + analyte
   + analyte half-life
   + analyte procedure number 
   + result amount 
   
   Variables also include units of measure, e.g. size of sample (g, L) or time of decay (sec, min, hour, day, year). No filters or limitations were placed on location or types of radionuclides analysed. Other variables such as project numbers, surface water locations, sample collection start etc., exist in the RadNet database, but in effort to to control file size and project scope, these data were not downloaded.

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_Data}
### !! DELETE existing dataframes if rerunning this chunk 
### otherwise data will be re-appended

### List files in Data Folder 
#list.files('radnet_data/', recursive = TRUE)

### Load/Append decade .csv files into 5 data frames, one for each media type
### air_data, drinking_water_data, Milk_data, precipitation_data,
### surface_water_data

dir_list <-  list.files('radnet_data/')

for (directory in dir_list){
  subdir <- (paste('radnet_data/', directory, sep=""))
  print(subdir)
  files_in_sub <- list.files(subdir)
  dfname <- paste(directory, '_data', sep="")  

  for (file in files_in_sub){
    file_loc <- paste(subdir, '/', file, sep="")
    # print(file_loc)
    # if the  dataset doesn't exist, create it
    if (!exists("db")){
        db <- read.csv(file_loc, header=TRUE)
    }
    # if the dataset does exist, append to it
    else if (exists("db")){
        temp_dataset <-read.csv(file_loc, header=TRUE)
        db <-rbind(db, temp_dataset)
       rm(temp_dataset)
    }
  }
  assign(dfname, db)
  remove(db)
}

### Cleanup Workspace
rm(dfname, dir_list, directory, file, file_loc, files_in_sub, subdir)

```



```{r echo=FALSE, message=FALSE, warning=FALSE, consolidate_data}
### First attempt to bind media datasets with loop failed
### List of dataframe names with 
### `as.list(names(which(sapply(.GlobalEnv, is.data.frame))))`
### [^df_names]:
### <http://r.789695.n4.nabble.com/getting-list-of-data-frame-names-td3864338.html>
### generates character list of dataframes (listofDF)
### which is passed to bind_rows() as CHAR representation
### not df object 
### dataframe names were finally hard coded into `bind_rows()`.

### SAMP_ID iS INT in surface_water_data and CHR in all other dataframes 
### convert before df consolidation into one dataframe
### ignore factor coercion to character vectors
surface_water_data$V_ERAMS_MATRIX_SAMPLE_ANALYSIS.SAMP_ID <-
  as.character(
    surface_water_data$V_ERAMS_MATRIX_SAMPLE_ANALYSIS.SAMP_ID)

### Consolidate media data into one dataframe: <rad_data_raw>
rad_data_raw <- bind_rows(list(air_data, drinking_water_data, 
                               Milk_data, precipitation_data,
                               surface_water_data
                               ),
                         .id ="id")
```

#Data Overview

A quick inspection of the dataframe and the contents of each of the variables was undertaken in order to understand the breadth of the data and potential problems.
At this stage the raw dataframe has 20 variables and 504092 observations. One variable *id* was added as part of the dataframes merge to easily identify the original source (Media Type) of the data.

#### (@)Data Cleanup and Wrangling (Recast Variable Types)

All the variables were prepended with the name of SQL tables frome which they originated, e.g. V_ERAMS_MATRIX_SAMPLE_ANALYSIS.MAT_ID.  To shorten the variable names for ease of programming, all characters up to and including the period were removed.

Several variables were recast as factors with ordered levels where it made sense that an inmposed order may be useful, e.g., S(econd) < M(inute) < H(our) < D(ay) < Y(ear). All factors/level were lost in `bind_rows()`

```{r echo=FALSE, message=FALSE, warning=FALSE, cleanup}
### Rename Variables
#######################################################################
colnames(surface_water_data) <-gsub('.+\\.','',colnames(surface_water_data))
colnames(Milk_data) <-gsub('.+\\.','',colnames(Milk_data))
colnames(air_data) <-gsub('.+\\.','',colnames(air_data))
colnames(precipitation_data) <-gsub('.+\\.','',colnames(precipitation_data))
colnames(drinking_water_data)<-
  gsub('.+\\.','',colnames(drinking_water_data))
colnames(rad_data_raw) <-gsub('.+\\.','',colnames(rad_data_raw))


### DF rad_data_raw  Summary
#######################################################################
names(rad_data_raw)
dim(rad_data_raw)
str(rad_data_raw)
head(rad_data_raw)


### level/order variables of limited categories where it may be useful to 
### impose order, e.g., S(econd) < M(inute) < H(our) < D(ay) < Y(ear).
########################################################################

### (1) ID for original df source (numerical tie to media type)
rad_data_raw$id          <- factor(as.integer(rad_data_raw$id))
### (2) MAT_ID: material type e.g. Air, SW, DW, PPT, MILK
rad_data_raw$MAT_ID      <- factor(rad_data_raw$MAT_ID) 
### (5) ANA_UNIT: size of sample G,L,M3, MG, ML
rad_data_raw$ANA_UNIT    <- factor(rad_data_raw$ANA_UNIT)       
### (15) RESULT_UNIT: 5 Result measures -> ACI/M3, DPM/GC, G/L, PCI/L, PCI/M3
rad_data_raw$RESULT_UNIT <- factor(rad_data_raw$RESULT_UNIT)
### (17) ANA_TYPE Analyte two types E (Element) and R (Radionuclide)
rad_data_raw$ANA_TYPE    <- factor(rad_data_raw$ANA_TYPE)
### (19) HALF_LIFE_TIME_UNIT: Five time Measures of Half Life: S, M, H, D, Y
### Create ordered factor for Time units sec<min<hour<day<year
rad_data_raw$HALF_LIFE_TIME_UNIT <-
  factor(rad_data_raw$HALF_LIFE_TIME_UNIT,
         levels = c("S", "M", "H", "D", "Y"))



### (1) id -> type CHR - ties merged data to original MEDIA dataframes 
###           1-5 :  (1) air_data, (2) drinking_water_data, 
###                  (3) Milk_data,(4)precipitation_data,
###                  (5) surface_water_data
#######################################################################
#table(rad_data_raw$id)

```

#### (@)Variable Audit

  In the event of possible date manipulations, the date field was changed to type DATE instead of CHR.
[^as.Date]:<https://www.rdocumentation.org/packages/base/versions/3.4.1/topics/as.Date>.

A simple analyses of each variable helps to understand a bit more about the data. The half million observations in the dataframe are not evenly distributed between the 5 sample (material) types. One-half of the entries are air samples while the next largest category is precipitation with only 22% of the entries. The remaining categories contain about 28% of the observations. 

Additionally, each observation does not necessarily indicate a new sample. Only $$\frac{1}{2}$$ of the sample id numbers are unique. Samples ids are replicated as many as 25, though most are used $$<$$11 times. This is not be too surprising given that there are 61 different analytes, ranging from gross alpha to zirconium-95, for which each sample might be analysed. Of these analytes Beta ranks highest accounting for nearly $$\frac{1}{2}$$ of the observations. In fact, the top 10 most analysed analytes account for 85% of the dataframe observations.  

Becuase result values are not normalized to a single unit of measure it is important to check that values compared directly are on the same scale. The *unit size*, besides providing a comparison baseline, also reflects the state of matter for different sample types or analytes: mL and L for liquids; mg and g for solids. Not surprising, the unit for gases ($$m^3$$) is the most prevalent, which correlates nicely with the fact that half of the entries are AIR-FILTER/CHARCOAL observations as noted earlier. 

Interestingly, 16% (81,291) of the rows in the database have no result (NA). These rows were not filtered in the event some information could be deduced about these samples or the about RadNet in general. The missing values do not appear to belong to any specific group, e.g., date range, material type, analyte, monitoring station. 

There are 34 analytical procedure numbers which can be correlated to the different sample type or analyte. For example, procedure **1** is returned as the mode of this variable and is therefore likely tied to the analysis of the air samples, being as the counts are . A duration variable shows that while most tests are under 20 hours there are some that took $$>$$80 hours.


```{r echo=FALSE, message=FALSE, warning=FALSE, variable_review1}
###  Variable Audits 2-7

### (2) MAT_ID -> type CHR, 6 material types analysed - air_data is subdivided 
###               AIR-CHARCOAL (Discontinued in 1980s) & AIR-FILTER
#######################################################################
#count(rad_data_raw, MAT_ID)       # Predominantly AIR-Filter 244,590 
#table(rad_data_raw$MAT_ID)
sample_types <- ggplot(rad_data_raw, aes(MAT_ID)) + 
  geom_bar(fill = "SteelBlue") +
  scale_x_discrete(limits = c("AIR-FILTER", "PRECIPITATION",
                   "PASTEURIZED MILK", "DRINKING WATER", 
                   "SURFACE WATER", "AIR-CHARCOAL")) +
  geom_text(stat='count',aes(label=..count..), color="black")+
  coord_flip() +
  labs(title="Number of Analyses by Media Type", x=NULL, y = "Count") +
  theme(plot.title = element_text(hjust = 0.5))
sample_types

### (3) SAMP_ID -> type CHR, sample ID number, some are alpha-numeric, 
###                roughly 1/2 the number of observations, i.e, multiple
###                observations (analyses) for a single SAMP_ID
#######################################################################
# n_distinct(rad_data_raw$SAMP_ID)
# n_distinct(rad_data_raw$SAMP_ID)/length(rad_data_raw$SAMP_ID)
samp_replicates <- rad_data_raw %>%
                   group_by(SAMP_ID, LOC_NUM) %>%
                   summarise(n = n())
#table(samp_replicates$n)
sample_reps <- ggplot(samp_replicates, aes(n)) + 
  geom_histogram(bins = 25,fill ="cornflowerblue", col ="black") +
  scale_y_sqrt() +
  labs(title = "Distribution of Sample Replicates",  
       x = "Number of Sample Replicates", y = "COUNT")
sample_reps

### (4) ANA_SIZE -> type NUM, numeric 
#######################################################################
sample_sizes <- ggplot(rad_data_raw, aes(ANA_SIZE)) + 
  geom_histogram(bins = 20) +
  scale_x_log10(labels=scales::comma) +
  labs(title = "Sample Sizes ... not Normalized for Unit of Measure",  
       x = "Numeric Size", y = "COUNT")
sample_sizes

### (5) ANA_UNIT ->CHR, measurement unit for analysis size 
###   (air = m3, solids = mg, g, liquids = mL, L, "" blank 3206)
#######################################################################
table(rad_data_raw$ANA_UNIT)
#count(rad_data_raw, ANA_UNIT)     # M3 followed by L
# ggplot(rad_data_raw, aes(ANA_UNIT)) + 
#   geom_bar() +
#   labs(title = "Count of Sample Unit Sizes",  
#        x = "Unit", y = "COUNT")

### (6) ANA_PROC_NUM -> INT; 34 analytical procedures used for analysis
###               Proc Num range from 1 to 170 with mode = ProcNum 1
#######################################################################
### Calculate the mode using the getmode function
### https://www.tutorialspoint.com/r/r_mean_median_mode.htm
getmode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}
getmode(rad_data_raw$ANA_PROC_NUM)
#count(rad_data_raw, ANA_PROC_NUM) # Procedure #1 217744
#n_distinct(rad_data_raw$ANA_PROC_NUM)
#table(rad_data_raw$ANA_PROC_NUM)
procedures <- ggplot(rad_data_raw, aes(ANA_PROC_NUM)) +
  geom_histogram(bins = 200, fill ="cornflowerblue", col ="black") +
  scale_x_continuous(limits = c(0,200))+
  labs(title = "Analytical Procedure Counts",  
       x = "Procedure Number", y = "COUNT")+
  theme(plot.title = element_text(hjust = 0.5))
procedures

### (7) DURATION -> NUM range 1-5000, duration of count in minutes
###                 division of Duration/60 to give hours
#######################################################################
#summary(rad_data_raw$DURATION)
#table(rad_data_raw$DURATION)
test_duration <- ggplot(rad_data_raw, aes(DURATION/60)) + 
  geom_histogram(bins = 75, fill = "darkseagreen") +
  scale_y_log10(labels=scales::comma) +
  labs(title = "Testing Count Duration",  
       x = "Duration (h)", y = "COUNT")+
   theme(plot.title = element_text(hjust = 0.5))
test_duration
```


There is a bit of geographic data within the data set. There are 324 different sampling locations in 289 cities or EPA Regions which are spread through the 50 United States, the District of Columbia, several US territories (Puerto Rico (PR), Guam (GU), Virgin Islands (VI), Saipan - Commonwealth of the Northern Mariana Islands (CNMI)), as well as the Pananma Canal (PC) and Ottawa, Canada (ON). Each site has a numberical ID and summary statstics on grouped locations shows that the samples are not evenly distributed amongst the monitoring sites. In fact a few sites only reported 1 observation while  one location has more that 12,000 entries. 

One amusing oberservation made when evaluating the date data (result date) is that while there is fairly consistent number of entries from 1978 to 2017 there are regular intervals in which the numbers spike above background. Closer inspection revealed that these spikes are annual, occuring at mid-year (7/1) up to 2010 and year-end (12/31) thereafter. As the dates are consistent I surmise it is likely an administrative consequence (fiscal year end) as opposed to an annual environmental or testing phenomenon.


```{r, echo=FALSE, message=FALSE, warning=FALSE, variable_review2}
###  Variable Audits 8-10

### (8) LOC_NUM -> INT ID of test location 1-4157, 324 distinct locations
#######################################################################
# summary(rad_data_raw$LOC_NUM)
# n_distinct(rad_data_raw$LOC_NUM)
# count(rad_data_raw, LOC_NUM)
# table(rad_data_raw$LOC_NUM)
# freq_LOC <- rad_data_raw %>% group_by(LOC_NUM) %>% summarize( n = n())
# summary(freq_LOC)

### Histogram of observations by location number
# ggplot(rad_data_raw, aes(LOC_NUM)) +
#  geom_histogram(bins =1000, fill = "salmon")+
#  labs(title = "Analysis Counts by Location Number",  
#       x = "Location ID", y = "COUNT")+
#  theme(plot.title = element_text(hjust = 0.5))
### Density plot 
#ggplot(rad_data_raw, aes(LOC_NUM)) + 
#  geom_density()+
#  labs(title = "Distribution of Analyses by Location"
#       x = "Location ID", y = "Density")+
#  theme(plot.title = element_text(hjust = 0.5))


### (9) CITY_NAME -> CHR, 289 Distinct City, county or EPA Region associated 
###                  with sampling location
#######################################################################
#n_distinct(rad_data_raw$CITY_NAME)
#count(rad_data_raw, CITY_NAME)    # top cities have 5000-8000 rows

### (10) STATE_ABBR-> CHR; 67 state, territory or country of 
###                        sampling location
###                   50 US States +
###                   DC (District of Columbia)
###                   US Territories : PR (Puerto Rico), GU (Guam), 
###                                    PC (Panama Canal), VI (Virgin Islands)
###                   CNMI - Commonwealth of the No. Mariana Islands (Saipan)
###                  EPA Regions R01- R10
###                   ON Ottawa, ON
#######################################################################
#n_distinct(rad_data_raw$STATE_ABBR)
#table(rad_data_raw$STATE_ABBR)
#count(rad_data_raw, STATE_ABBR)   # varied top states with 10K - 15K rows
```


```{r echo=FALSE, message=FALSE, warning=FALSE, variable_review3 }
###  Variable Audits 11-20
### (11) ANALYTE_ID -> CHR, Analyte of interest, 61 unique analytes
###                    Abbreviation and isotope e.g., CO60 = Cobalt 60
### Most common Gross Beta 232,387
###################################################################
#n_distinct(rad_data_raw$ANALYTE_ID)
#table(rad_data_raw$ANALYTE_ID)
#count(rad_data_raw, ANALYTE_NAME) 
analytes <- rad_data_raw %>% group_by(ANALYTE_ID) %>% summarise(n = n())
top_analytes <- filter(analytes, n > 9000)
arrange(top_analytes, desc(n))

### (12) RESULT_AMOUNT -> NUM, range -200-257000;
###                        amount, activity or concentraion of analyte 
###                        81921 NA's (16%)
### ROWS without RESULT_AMOUNT: 81,291 of the 504,092 observations have
### NA for RESULT AMOUNT
#######################################################################
#summary(rad_data_raw$RESULT_AMOUNT)
#sum(is.na(rad_data_raw$RESULT_AMOUNT))/nrow(rad_data_raw)
#count(rad_data_raw, !is.na(rad_data_raw$RESULT_AMOUNT))

### ?? Filter out the NA rows??
#rad_data_filt <- rad_data[complete.cases(rad_data[,"RESULT_AMOUNT"]),]
#######################################################################

### (13) CSU -> NUM, 0-15000 Combined Standard Uncertainty, 
###                       81921 NA's (16%)
#######################################################################
#summary(rad_data_raw$CSU)

### (14) MDC -> NUM, 0-9700 Min Detectable Concentration
###                       218822 NA's (43%)
#######################################################################
#summary(rad_data_raw$MDC)
#sum(is.na(rad_data_raw$MDC))/nrow(rad_data_raw)

### (15) RESULT_UNIT -> unit of measure for result 
###                (air = ACI/m3, PCI/M3 (45%); solids = g/L, DPM/GC;
###                 liquids = pCi/L (48%))
#######################################################################
#table(rad_data_raw$RESULT_UNIT)
#sum(rad_data_raw$RESULT_UNIT == "PCI/L")/nrow(rad_data_raw)
#sum(rad_data_raw$RESULT_UNIT == "PCI/M3")/nrow(rad_data_raw)
#count(rad_data_raw, RESULT_UNIT)  # PCI/M3 air followed by PCI/L liquid

### (16) RESULT_DATE -> CHR transform to DATE range from 7/1/1978 to 7/26/2017
###      Set DATE to ISO 8601 format:  %F == "%Y-%m-%d"
#######################################################################
#summary(rad_data_raw$RESULT_DATE)
rad_data_raw$RESULT_DATE <- as.Date(rad_data_raw$RESULT_DATE, "%F")
analysis_counts <- c(510,565,480,490,1025,760,325,312 )
analysis_dates <- c("1996-07-01","1997-07-01", "1998-07-01","1999-07-01",
                    "2009-12-31", "2010-12-31","2011-12-31", "2012-12-31")
label_df <- data.frame(analysis_counts, analysis_dates)
label_df$analysis_dates <- as.Date(label_df$analysis_dates)

bydate_analyses <- ggplot(rad_data_raw, aes(RESULT_DATE)) + 
  geom_line(color = "steelblue ", stat = "count") +
  scale_x_date(date_breaks = "5 year", date_minor_breaks = "year")+
  labs(title="Number of Analyses Results by Date", x=NULL, y = "Count") +
  geom_point(data = label_df, aes(x=label_df$analysis_dates,
                                  y=label_df$analysis_counts), color = "red") +
  annotate("text", label = "Yearly Spikes in Analysis
           every July 1st until 2010\nthen December 31st ",
            x= as.Date("1994-01-01"),
            y= 875) +
  theme(axis.text.x = element_text(angle=30),
        plot.title = element_text(hjust = 0.5))

bydate_analyses 

### (17) ANA_TYPE -> CHR, "E" (Element) = 10585, "R" (Radionuclide)= 493507
#######################################################################
#table(rad_data_raw$ANA_TYPE)

### (18) HALF_LIFE -> quantity part of an elements half-life
###                   251749 50% NA's
#######################################################################
#summary(rad_data_raw$HALF_LIFE)
#sum(is.na(rad_data_raw$HALF_LIFE))/nrow(rad_data_raw)
element_half_life <- subset(rad_data_raw[ , c(11,18,19)], !is.na(HALF_LIFE))
element_half_life <- within(element_half_life,
                     ANALYTE_ORDER <- factor(ANALYTE_ID,
                                      levels=names(sort(table(ANALYTE_ID)))))
element_lambda <- ggplot(element_half_life,
                       aes(x = ANALYTE_ORDER)) + 
  geom_bar(fill = "SteelBlue") +
  scale_y_log10() +
  geom_text(stat='count',aes(label=..count..), color="black")+
  coord_flip() +
  labs(title="Count of Analyses for Analytes with Half Life", 
       x=NULL, y = "Count (log10)") +
  theme(plot.title = element_text(hjust = 0.5))
element_lambda

### (19) HALF_LIFE_TIME_UNIT -> CHR; D,H,M,S,Y and 50% blank (251749)
###     add new variable normalize Half-Lives to years
#######################################################################
table(rad_data_raw$HALF_LIFE_TIME_UNIT)
half_lives <- unique(element_half_life) %>% 
  mutate(HALF_LIFE_YEAR = case_when(HALF_LIFE_TIME_UNIT =="S" 
                                    ~ HALF_LIFE/3153600,
                                    HALF_LIFE_TIME_UNIT =="M" 
                                    ~ HALF_LIFE/525600,
                                    HALF_LIFE_TIME_UNIT =="H" 
                                    ~ HALF_LIFE/8760,
                                    HALF_LIFE_TIME_UNIT =="D" 
                                    ~ HALF_LIFE/365,
                                    HALF_LIFE_TIME_UNIT =="Y" 
                                    ~ HALF_LIFE/1))
half_lives <- within(half_lives, sort(HALF_LIFE_YEAR))
lambdas <- ggplot(half_lives,
                       aes(reorder(ANALYTE_ID, -HALF_LIFE_YEAR, median),
                           HALF_LIFE_YEAR)) + 
  geom_col(fill = "SteelBlue") +
  scale_y_log10(
   breaks = scales::trans_breaks("log10", function(x) 10^x, n=7),
   labels = scales::trans_format("log10", scales::math_format(10^.x))
   ## http://ggplot2.tidyverse.org/reference/annotation_logticks.html
   ) +
  coord_flip() +
  labs(title="Analyte Half Lives ", 
       x=NULL, y = "Half-Life Years (log10)") +
  theme(plot.title = element_text(hjust = 0.5))
lambdas 

### (20) ANALYTE_NAME -> full name of (11) ANALYTE_ID
###                      e.g, Gross Beta, Uranium-235
#######################################################################
#n_distinct(rad_data_raw$ANALYTE_NAME)

```
 

```{r echo=FALSE, include=FALSE, test_summaries}
### Review info for cleaned rad_data
head(rad_data_raw)
dim(rad_data_raw)
names(rad_data_raw)
summary(rad_data_raw)

summary(rad_data_raw$ANA_UNIT)
summary(rad_data_raw$RESULT_UNIT)
summary(rad_data_raw$HALF_LIFE_TIME_UNIT)
levels(rad_data_raw$HALF_LIFE_TIME_UNIT)
```

#### (@)Geocoding Locations and Geospatial Graphing
With the package `ggmap`[^ggmap](https://cran.r-project.org/web/packages/ggmap/ggmap.pdf) it is possible to represent data and information geographically.  Using the function `geocode()`[^geocode](https://www.rdocumentation.org/packages/ggmap/versions/2.6.1/topics/geocode),[^geocode_csv](http://www.storybench.org/geocode-csv-addresses-r/) the latitude and longitude of locations were be obtained from the city and state information. Additionally the packages `choroplethr` and `choroplethrMaps` allow for various values to be represented on a shaded and keyed `Choropleth Map`[^choropleth](https://en.wikipedia.org/wiki/Choropleth_map).  

```{r echo=FALSE, include=FALSE, test_code}
### Test CHUNK run of `geocode()`
### returns lat/lon and address of first row in dataframe
citystate <- paste(rad_data[1,which( colnames(rad_data)=="STATE_ABBR")],
                   rad_data[1,which( colnames(rad_data)=="CITY_NAME")])
geocode(citystate, output = "latlona", source="dsk") #datasciencetoolkit.org
geocode("Corozal, Panama", output = "latlona")
geocode("Kauai, HI",  output = "latlona", source="google")
geocodeQueryCheck(userType = "free")
```

In an attempt to map the sampling locations a few issues were found in the data entry. The state abbreviation PC for Panama Canal was not recognized by the API and consequently needed to be changed. A single entry identified as Doswell, SC does not appear to exist, but there are several Doswell, VA observations so this entry was changed. Several entries were listed as EPA regions 1-10. In an effort to geo-locate the city-state of each region was changed to the headquarters location for that EPA region as listed on the EPA website[^eparegions](https://www.epa.gov/aboutepa#pane-4).

```{r echo=FALSE, message=FALSE, warning=FALSE, geo_fixes}
### Fix location errors:
### geocode does not find 'PC', changed STATE_ABBR to 'Panama' 
### One entry misentered? Doswell SC changed to Doswell VA, LOC_NUM does not
### match any other entry, but latlong.net cannot find location either
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "PC")] <- "PANAMA"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "SC") &
                          (rad_data_raw$CITY_NAME == "DOSWELL")] <- "VA"

### The 10 EPA Regions are mapped to LOC_NUM 59-68
### Set EPA REGIONS to corresponding Headquarter City
####################################################################
### R01 New England (CT, ME, MA, NH, RI, VT and 10 tribal nations)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R01")]   <- "BOSTON"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R01")]  <- "MA"
### R02 NJ,NY, Puerto Rico (PR), US VI and 8 tribal nations)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R02")]   <- "New York City"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R02")]  <- "NY"
### R03 Mid-Atlantic (DE, DC, MD, PA, VA, WV)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R03")]   <- "PHILADELPHIA"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R03")]  <- "PA"
### R04 Southeast (AL, FL, GA, KY, MS, NC, SC, TN and 6 tribes)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R04")]   <- "ATLANTA"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R04")]  <- "GA"
### R05 IL, IN, MI, MN, OH, WI and 35 tribal nations                    
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R05")]   <- "CHICAGO"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R05")]  <- "IL"
### R06 South Central (AK, LA, NM, OK, TX and 66 tribal nations)      
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R06")]   <- "DALLAS"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R06")]  <- "TX"
### R07 Midwest (IA, KS, MO, NE and 9 tribal nations)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R07")]   <- "KANSAS CITY"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R07")]  <- "KS"
### R08 Mountains and Plains (CO, MT, ND, SD, UT, WY and 27 tribal nations)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R08")]  <- "DENVER"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R08")]  <- "CO"
### R09 Pacific Southwest (AZ, CA, HI, NV, Pacific Islands and 148 tribal nations)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R09")]   <- "SAN FRANCISCO"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R09")]  <- "CA"
### R10 Pacific Northwest (AK, ID, OR, WA and 271 Native Tribes)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R10")]   <- "SEATTLE"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R10")]  <- "WA"
                      
```

Rather than inundate the API with multiple requests for the same location a separate dataframe was created listing only unique city-state combinations.  This removes the instances where a city has multiple monitoring stations (LOC_NUM). 

```{r echo=FALSE, message=FALSE, warning=FALSE, csv_latlona}
#### The file "places_data.csv" was made by downloading lat/lon data from tbe google API
#### The code for obtaining and cleaning the geo data is in the following 2 chunks 
#### (get_latlona and adjust_latlon)
#### Due to the time needed to run, a .csv file was generated and the code chunks were #### set to "include = FALSE" - they can be run by removing the include statement

places <- read.csv("places_data.csv", header=TRUE)
########################################################################
```

```{r include=FALSE, echo=FALSE, message=FALSE, warning=FALSE, get_latlona}
########################################################################
### WARNING THIS CHUNK DOWNLOADS LAT/LON FROM google and takes TIME!
########################################################################

### Generate dataframe/list of City/State for geocoding
### Not using LOC_NUM as several cities half multiple monitoring stations
###########################################################################
places <- data.frame("CITY"   = rad_data_raw$CITY_NAME,   
                     "STATE"  = rad_data_raw$STATE_ABBR)
places$CITY <- as.character(places$CITY)
places$STATE <- as.character(places$STATE)
places <- unique(places)
places <- arrange(places, CITY)
rownames(places) <- NULL  #reset rownames
places$citystate <- paste(places$CITY, places$STATE)
places$lat     <- NA
places$lon     <- NA
places$address <- NA


### Use tryCatch with geocode() in case of Errors from API
### to bypass error from places which have no specific LAT/LON
### A 1s pause is used between API requests, some requests were not 
### returning results first try this significantly delays runtime
###########################################################################
### [Errors from Geocode]
### (https://stackoverflow.com/questions/30770328/
###  how-to-handle-error-from-geocode-ggmap-r)

miss_geo <- character()  ### Create list of geocoding potential issues/misses
for (i in 1:nrow(places)) {
  z <- 0
  repeat{
    geo_result <- tryCatch(geocode(places[i,3], output = c("latlona")),
                      warning = function(w) {
                        paste("Location Issue ", places[i,3]);
                        places[i,4] <- NA
                        places[i,5] <- NA  
                        places[i,6] <- NA
                      },
                      error = function(e) {
                          paste("Location error", places[i,3]);
                          next
                      })
    places[i,4] <- as.numeric(geo_result[2])
    places[i,5] <- as.numeric(geo_result[1])
    places[i,6] <- as.character(unlist(geo_result[3]))
    Sys.sleep(1)  # pause before next API request
    if (!is.na(places[i,6]) | z==2) break
    if (is.na(places[i,6])) {
      z = z+1
      print(paste(places[i,3], "--NA", z, i)) #check if info returned
      }
  }
  banana <- strsplit(gsub('[[:digit:]]',"", places[i,6]), ", +")[[1]]
  test_geo <-paste(toupper(banana[1]),
                     toupper(gsub('[[:blank:]]',"",banana[2])))
  if (test_geo != places[i,3]) {miss_geo <- c(miss_geo, places[i,3])}
  if (i%%5 ==0) print(i)   # visual feedback to see if code is still running
}
```

```{r echo=FALSE, include = FALSE, adjust_latlon }
### This CHUNK rechecks the places database and ouputs a .csv file
### For RETESTING GEO CODE And Getting Missing lat/lon
### Check places for is.na(); 
### manually change n (row#) to resend geocode() for that location
######################################################################
# count(places, is.na(places$lat))
# filter(places, is.na(lat))
# miss_geo # differences include county,zip code, accent marks
# n=140
# geo_result <- geocode(places[n,3], output = c("latlona"))
# places[n,4] <- as.numeric(geo_result[2])
# places[n,5] <- as.numeric(geo_result[1])
# places[n,6] <- as.character(unlist(geo_result[3]))
# places[n,6]

summary(places$lat)
summary(places$lon)

###  SAVE PLACES AS .csv to avoid reloading from API
write.csv(places, "places_data.csv", row.names = FALSE)

### Revised latlong data from latlong.net - 
### geocoding from source = "dsk" puts Honolulu in ocean 
###           and Kauai on the wrong island as found during mapping
###########################################################################
#places$lat[places$CITY == "HONOLULU"] = 21.306944
#places$lon[places$CITY == "HONOLULU"] = -157.858333
#places$lat[places$CITY == "KAUAI"] = 22.096440
#places$lon[places$CITY == "KAUAI"] = -159.526124
```


```{r}
### Merge lat/lon info into rad_data_raw
#########################################
rad_data_raw <- left_join(rad_data_raw, places, 
                          by = c("CITY_NAME" = "CITY",
                                 "STATE_ABBR" = "STATE") )
```


##Graphic Visualization with R's `ggmap`

<https://blog.dominodatalab.com/geographic-visualization-with-rs-ggmaps/>
<https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/ggmap/ggmapCheatsheet.pdf>
--------

```{r, include=FALSE, testmapping}

###  Test mapping

USAmap2<-get_map(location='united states', zoom=3, maptype = "terrain",
             source='google',color='color', urlonly = TRUE)
USAmap2<-get_openstreetmap(bbox=usa_cont_bounds,scale = 475000000,
                           color='color')
 
USAmap2

myMap <- get_stamenmap(bbox=usa_cont_bounds)
ggmap(myMap)


##qmplot(lon, lat, data = places)

us <- c(left = -125, bottom = 25.75, right = -67, top = 49)
map <- get_stamenmap(us, zoom = 5, maptype = "toner-lite")
ggmap(map) + geom_point(data = continental, aes(x= lon, y= lat),
                    col="blue", alpha=0.4)
```

The latitude/longitude points of all the monitoring stations can be visualized over the appropriate maps. In order to make the mapping clear, the map areas are sized to be neat and manageable. Consequently, Alaska, Hawaii and the other islands in the western Pacific Ocean are treated separately with Google Maps. The state choroplethrMap include AK and HI as an inset however, US territories and other regions are difficult to include on choropleth maps.

```{r echo=FALSE,  message=FALSE, warning=FALSE, northamericamap}
usa_center = as.numeric(geocode("United States"))
usa_cont_bounds = c(left = -130.0,bottom = 11.0, right = -60.0, top = 51.0)
USAMap = ggmap(get_googlemap(center=usa_center, scale=2, zoom=3, 
                             extent ="device"))
        
continental <- places %>% filter(between(lon, -130, -60) & 
                                      between(lat, 8, 55))
noncontinental <- places %>% filter(!between(lon, -130, -60))

### US Map with sampling locations
USAMap + 
  scale_y_continuous(limit = c(8,50)) +
  scale_x_continuous(limit = c(-130,-62)) +
  geom_point(data = continental, aes(x= lon, y= lat), 
             col="blue", alpha=0.4) +
  labs(caption = 'EPA RadNet Monitoring Locations 1978-2017') +
  theme_map()

```

In order to use the function `state_choropleth{choroplethr}`, the dataframe must contain a column named "region" matching the naming convention found for "region" in `state.map`. The convention for US states names is the full state name in lower case. The values to be mapped must be contained in column named "value". Several data sets include data related to the USA and using the charactor vectors `state.abb` and `state.name` it was simple to make a dataframe which could be joined as an appropriately formatted "region" column using state abbreviation column already contained in the EPA RadNet dataset. However, as the state data does not include the territories several regions were added just to complete complete the fields. 

```{r echo=FALSE,  message=FALSE, warning=FALSE, stateregions}
### Include *region* as recognized by choroplethr
### lower case full state name, eg. new mexico
#################################################
state_regions <- data.frame(state.abb,state.name)
state_regions$state.name <- tolower(state_regions$state.name)
colnames(state_regions) <- c("abbreviation", "region")
state_regions$abbreviation <- as.character(state_regions$abbreviation)

### Add regions for places not included in state.abb/state.name
#################################################
state_regions <- rbind(state_regions, c("abbreviation" = "PR",
                                  "region" = "puerto rico"))
state_regions <- rbind(state_regions, data.frame("abbreviation" = "DC", 
                                   "region" = "district of columbia"))
state_regions <- rbind(state_regions, data.frame("abbreviation" = "GU", 
                                   "region" = "guam"))
state_regions <- rbind(state_regions, data.frame("abbreviation" = "VI", 
                                   "region" = "virgin islands"))
state_regions <- rbind(state_regions, data.frame("abbreviation" = "ON", 
                                                 "region" = "ontario"))
state_regions <- rbind(state_regions, data.frame("abbreviation" = "PANAMA",
                                                 "region" = "panama"))
state_regions <- rbind(state_regions, data.frame("abbreviation" = "CNMI", 
                                   "region" = "mariana islands"))

places <- left_join(places, state_regions,
                    by = c("STATE" = "abbreviation"))
rad_data_raw <- left_join(rad_data_raw, state_regions,
                          by= c("STATE_ABBR" = "abbreviation"))
rm(state_regions)

```

Using the choropleth we can easily compare information by state. For example, while the number of observations entered into the RadNet database correlates well to the density of monitoring stations as seen in the Monitoring station map the number of empty results is not a simple portion of the total number of samples entered by a state. Some states with moderate/low number of observations have a higher percentage of empty results, e.g., Kentucky with <4,500 entries has > 17% empty, while California with many more samples (>19,000) has <14.8% NA results. 

```{r echo=FALSE,  message=FALSE, warning=FALSE, uschoropleth}

### Choropleth of Sample Analyses by State
bystate_numanalyses <- rad_data_raw %>% group_by(region) %>%
  summarise(value = n())

choro1 <-state_choropleth(bystate_numanalyses, 
                 title="Number of RADNet Analyses by State") + 
                 labs(subtitle = '1978-2017',
                      caption = 'EPA RadNet Data') +
  theme(plot.caption = element_text(size = rel(0.5)),
         plot.title = element_text(hjust=0.5),
         plot.subtitle = element_text(size=rel(.75), hjust = 0.5))
choro1

### Choropleth of Empty Results by State
### New Colour Scheme
### https://www.r-bloggers.com/advanced-choroplethr-changing-color-scheme/
bystate_NaN <- rad_data_raw %>% filter(is.na(RESULT_AMOUNT)) %>% 
               group_by(region) %>%
               summarise(empty = n())
bystate_NaN$totalanal <- bystate_numanalyses$value
bystate_NaN$value <- round(bystate_NaN$empty/bystate_NaN$totalanal*100, digits =1)

col.pal<-brewer_pal(palette = "GnBu")(7)
#show_col(brewer_pal(palette = "Reds")(7))

state_nan <-StateChoropleth$new(bystate_NaN)
state_nan$title <- "Empty Result RADNet Entries by State"
state_nan$ggplot_scale <- scale_fill_manual(name="% Empty Result Entries",
                                         values=col.pal, drop=FALSE)
  
choro2 <-state_nan$render()  +
  theme(plot.caption = element_text(size = rel(0.5)),
         plot.title = element_text(hjust=0.5),
         plot.subtitle = element_text(size=rel(.75), hjust = 0.5)) + 
  labs(subtitle = '1978-2017', caption = 'EPA RadNet Data')
choro2

grid.arrange(choro1,choro2, ncol=1)
```

```{r echo=FALSE,  message=FALSE, warning=FALSE, HAWAII1}
###  HAWAII AREA MAPS
###  Using google maps centered
#################################
HI_center = c(-157.9, 21.2)
HIMap_g <- ggmap(get_googlemap(center=HI_center, zoom=7, extent="normal"))+
  scale_y_continuous(limit = c(19, 23)) +
  scale_x_continuous(limit = c(-161,-154)) + 
  theme_map()

hawaii <- noncontinental %>% filter(between(lon,-164,-150) & 
                                    between(lat,16,24))
HIMap_g  + 
  geom_point(data = hawaii, aes(x= lon, y= lat), colour ="red", size = 3)


```

```{r,echo=FALSE, include = false, message=FALSE, warning=FALSE, HAWAII2}
###  HAWAII AREA MAP
###  Using Stamen maps with bounding box
########################################
HI <- c(left = -161, bottom = 18, right = -154, top = 23)
#HIbox <- c( -161, 23, -154, 18)
HImap <- get_stamenmap(bbox= HI, zoom = 7, maptype = "toner-lite") 
HImap <- ggmap(HImap) + theme_map()

HImap +
  geom_point(data = hawaii, aes(x= lon, y= lat), colour ="red", size = 3)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, ALASKA}
### ALASKA AREA MAP
#####################
AK_center = c(-149.4937, 64.20084)
AKMap <- ggmap(get_googlemap(center=AK_center, 
                             zoom=4, extent="normal")) +
  scale_y_continuous(limit = c(50, 72)) +
  scale_x_continuous(limit = c(-175,-130)) + 
  theme_map()

alaska <- noncontinental %>% filter(between(lon,-167,-130) & 
                                    between(lat,53,65))
AKMap  + geom_point(data = alaska, aes(x= lon, y= lat),
                    col="purple4", alpha=0.9)
```


```{r echo=FALSE,  message=FALSE, warning=FALSE, micronesia}

### MICRONESIA AREA MAP
################################
#geocode("Songsong", output = "latlon") #center of Marianna Islands (MIsle)
MIsle_center <- c(145.1459, 14.14199)
MIsleMap <- ggmap(get_googlemap(center=MIsle_center, 
                               zoom=8, extent="normal")) +
  scale_y_continuous(limit = c(13, 15.5)) +
  scale_x_continuous(limit = c(144.4, 146.5)) +
  theme_map()

mariannas <- noncontinental %>% filter(between(lon,140,150))
MIsleMap  + geom_point(data = mariannas, aes(x= lon, y= lat),
                    col="deeppink", alpha=0.9)
```


```{r, echo=FALSE, message=FALSE, warning=FALSE, HI_minireviews}
### REVIEW OF DATA with smaller Subsets: HI/AK, micronesia
##########################################################

### Map of Materials Analysed in HI
#####################################
HI_data <- rad_data_raw %>% filter(STATE_ABBR == "HI")

HI_map_matid <- HImap + geom_count(data = subset(HI_data,
                                             !is.na(RESULT_AMOUNT)), 
             aes(x= lon, y= lat, color = MAT_ID), na.rm = TRUE,
             position = position_jitter(width = 0.1, height = 0.1)) +
             theme(legend.title = element_blank())
HI_map_matid

### Observations comparisons of no RESULT_AMOUNT
############################################
# table(HI_data$ANALYTE_ID, is.na(HI_data$RESULT_AMOUNT))
# count(HI_data, !is.na(HI_data$RESULT_AMOUNT)) 
#length(unique(HI_data$SAMP_ID))

HI_matID <- ggplot(subset(HI_data, is.na(RESULT_AMOUNT)), 
                   aes(RESULT_DATE, fill = MAT_ID)) +
  geom_histogram(bins = 50, alpha = 0.6, position = "identity")

HI_matIDnans <- ggplot(subset(HI_data, !is.na(RESULT_AMOUNT)),
                       aes(RESULT_DATE, fill = MAT_ID)) +
  geom_histogram(bins = 50, alpha = 0.6, position = "identity")

HI_analytes <- ggplot(subset(HI_data, !is.na(RESULT_AMOUNT)), 
                     aes(RESULT_DATE, fill = ANALYTE_NAME)) + 
  geom_histogram(bins = 50, alpha = 0.6)

HI_milk <- ggplot(subset(HI_data, (!is.na(RESULT_AMOUNT) &
                                     RESULT_UNIT == 'PCI/L' &
                                     MAT_ID == 'PASTEURIZED MILK')),
                  aes(RESULT_DATE, RESULT_AMOUNT, colour = ANALYTE_ID)) +
  geom_point() +
  labs(title = "Pasteurized Milk in HI",  
       x = "Year", y = "pCi/L")

HI_milk_h2o <- ggplot(subset(HI_data, ( !is.na(RESULT_AMOUNT) &
                                     RESULT_UNIT == 'PCI/L' & 
                                       (MAT_ID == 'PASTEURIZED MILK' |
                                        MAT_ID == "DRINKING WATER"))),
                      aes(RESULT_DATE, RESULT_AMOUNT, colour = MAT_ID)) +
  geom_point() +
  labs(title = "Pasteurized Milk & Drinking Water in HI",  
       x = "Year", y = "pCi/L")

HI_air <- ggplot(subset(HI_data, ( !is.na(RESULT_AMOUNT) &
                           RESULT_UNIT == 'PCI/M3' & 
                           (MAT_ID == 'AIR-FILTER' |
                              MAT_ID == "AIR-CHARCOAL"))),
              aes(RESULT_DATE, RESULT_AMOUNT, colour = CITY_NAME)) +
  geom_point() +
  labs(title = "Air Samples in Hawaii",  
       x = "Year", y = "pCi/m3")

HI_rain <- ggplot(subset(HI_data, ( !is.na(RESULT_AMOUNT) &
                                      RESULT_UNIT == 'PCI/L' & 
                           (MAT_ID == 'PRECIPITATION'))),
              aes(RESULT_DATE, RESULT_AMOUNT, colour = ANALYTE_ID)) +
  geom_point() +
  labs(title = "Precipitation Samples in HI",  
       x = "Year", y = "pCi/L")

HI_K40H3 <- ggplot(subset(HI_data, ( !is.na(RESULT_AMOUNT) &
                           RESULT_UNIT == 'PCI/L' & 
                           (ANALYTE_ID == 'K40' | ANALYTE_ID == "H3"))),
              aes(RESULT_DATE, RESULT_AMOUNT, colour = ANALYTE_ID)) +
  geom_point() +
  labs(title = "Potassium-40 & Tritium Analyses in Hawaii",  
       x = "Year", y = "pCi/L")

grid.arrange(HI_milk, HI_milk_h2o, HI_rain, HI_K40H3, ncol = 2)

HI_analytes <- ggplot(HI_data, aes(ANALYTE_ID)) + geom_bar(fill = "blue") +
  labs(title = 'RadNet Samples by Analyte', 
       caption = 'EPA RadNet Data 1978-2017') +
   xlab('Analyte') +
   ylab('Number of Samples') +
  theme(axis.text.x=element_text(angle=50,hjust=1))
```
 
```{r echo=FALSE, message=FALSE, warning=FALSE, overview_Plots}
### MAT_ID - sample types 
### Histograms filled by distributions of MAT_ID
################################################
all_matID <- ggplot(subset(rad_data_raw, !is.na(RESULT_AMOUNT)),
                           aes(RESULT_DATE, fill = MAT_ID)) +
  geom_histogram(bins = 50, alpha = 0.5, position = "identity") +
  labs(title='Number of Analyses by Material Type')+
  scale_x_date(limits = c(as.Date("1978-07-01"), as.Date("2017-07-26"))) +
  xlab('Date') +
  ylab('Count') +
  theme(legend.title = element_text())+
  scale_fill_brewer(type = "qual", palette = 1)

all_matIDnans <- ggplot(subset(rad_data_raw, is.na(RESULT_AMOUNT)),
                       aes(RESULT_DATE, fill = MAT_ID)) +
  geom_histogram(bins = 50, alpha = 0.5, position = "identity") +
  labs(title ='Missing Results by Material Type') +
  scale_x_date(limits = c(as.Date("1978-07-01"), as.Date("2017-07-26"))) +
  xlab('Date') +
  ylab('Count')+
  theme(legend.title = element_blank())+
  scale_fill_brewer(type = "qual", palette = 1)

all_byloc <- ggplot(rad_data_raw, aes(LOC_NUM, fill = MAT_ID)) +
  geom_histogram(bins = 50)+
  labs(title='Sample Types by Location ID') +
  xlab('Location ID') +
  ylab('Count')+
  theme(axis.text.x=element_text(angle=50,hjust=1)) +
  scale_fill_brewer(type = "qual", palette = 1)

### Plot of Result Amounts from all sources, units
################################################
all_result <- ggplot(subset(rad_data_raw, !is.na(RESULT_AMOUNT)), 
              aes(RESULT_AMOUNT, fill = MAT_ID,
                  title = "''")) +
         geom_histogram(bins = 50, position = "identity", alpha = 0.5) +
         scale_y_log10() +
         scale_x_log10() +
         labs(title = "Result Values by Material", 
              x = 'Raw Result Value (no unit normalization)',
              y = 'Count') +
           scale_fill_brewer(type = "qual", palette = 1)
#####################################################################
#### Grid.arrange common legend function
#### ref:  http://www.guru-gis.net/share-a-legend-between-multiple-plots-using-grid-arrange/
#####################################################################
grid_arrange_shared_legend <- function(...) {
    plots <- list(...)
    g <- ggplotGrob(plots[[1]] + theme(legend.position="bottom"))$grobs
    legend <- g[[which(sapply(g, function(x) x$name) == "guide-box")]]
    lheight <- sum(legend$height)
    grid.arrange(
        do.call(arrangeGrob, lapply(plots, function(x)
            x + theme(legend.position="none"))),
        legend,
        ncol = 1,
        heights = unit.c(unit(1, "npc") - lheight, lheight))
}
 
grid_arrange_shared_legend(all_matID, all_matIDnans, all_byloc, all_result)

### RESULT_AMOUNT
### Facet Wrap Result Amount as related to Unit Measure and Material Type
#######################################################################
rad_data <- rad_data_raw %>% filter(!is.na(RESULT_AMOUNT))
all_result_byunit <- ggplot(rad_data,   
                            aes(x = RESULT_DATE, y = RESULT_AMOUNT,
                                color = MAT_ID)) +
  geom_point() +
  facet_wrap( ~ rad_data$RESULT_UNIT, 
              scales = "free") +
  labs(title = 
         'Results of RAD Analyses by Unit Measure',
       caption = 'EPA RadNet Data',
       colour = 'Material Type') +
  xlab('Date') +
  ylab('Result Amount') +
  theme(
    plot.caption = element_text(size = rel(0.5)),
    plot.title = element_text(hjust = 0.5),
    legend.text = element_text(size = rel(0.5)),
    legend.position = c(0.85, 0.2),
    legend.title.align = 0.5
  )

all_result_grid <- ggplot(rad_data,   
                            aes(x = RESULT_DATE, y = RESULT_AMOUNT,
                                color = ANALYTE_ID)) +
  geom_point() +
  facet_grid(rad_data$RESULT_UNIT ~ rad_data$MAT_ID, 
              scales = "free", labeller = label_wrap_gen(multi_line = TRUE)) +
  labs(title = 
         'RADNET Analyses by Unit and Material Type',
       caption = 'EPA RadNet Data',
       colour = 'Material Type') +
  xlab('Date') +
  ylab('Result Amount') +
  theme(
    plot.caption = element_text(size = rel(0.5)),
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(size=7, angle=45),
    axis.title.x = element_blank(),
    axis.title.x.top = element_text(size = 6),
    legend.position = "bottom",
    legend.title = element_blank()
  )
all_result_grid

all_result_bymatID <-
  ggplot(rad_data,
         aes(x = RESULT_DATE, y = RESULT_AMOUNT, color = RESULT_UNIT)) +
  geom_point() +
  facet_wrap(~ rad_data$MAT_ID, scales = "free") +
  labs(title = 'Results of RAD Analyses by Material Type',
       caption = 'EPA RadNet Data', colour = 'Result Unit') +
  xlab('Date') + ylab('Result Amount') +
  geom_jitter(width = 0.1, height = 0.1) +
  theme(
    plot.caption = element_text(size = rel(0.5)),
    plot.title = element_text(hjust = 0.5),
    legend.text = element_text(size = rel(0.5)),
    legend.position = "bottom",
    legend.title.align = 0.5
  )
  
all_result_bymatID

### HI Beta Results (pCi/m3) by Date, coloured for location
#######################################################################
HI_beta <- ggplot(subset(HI_data, !is.na(RESULT_AMOUNT) &
                           RESULT_UNIT == "PCI/M3" & 
                           ANALYTE_ID =="BETA"),
                  aes(x=RESULT_DATE, y= RESULT_AMOUNT, color = CITY_NAME)) + 
  geom_point() +
  annotate("text", label = "0.894 Mar '11 (Fukushima)",
           x= as.Date("2011-03-25"),
           y=0.894-0.035) +
  annotate("text", label = "0.3758 May '86 (Chernobyl)",
           x= as.Date("1986-05-18"),
           y=0.3758 + 0.04) +
  labs(title = 'Gross Beta in HI Air Filter Samples 1978-2017', 
       caption = 'EPA RadNet Data') +
  scale_x_date(date_breaks = "5 years", labels = date_format("%Y")) +
  xlab('Date') +
  ylab('pCi/m3') +
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_text(hjust=0.5),
        legend.title = element_blank())
HI_beta

## Filter date to find High Points in above chart
#test <- HI_data %>% filter(RESULT_AMOUNT > 0.25 & 
#                                  ANALYTE_ID == "BETA" &
#                                  RESULT_UNIT == "PCI/M3" &
#                                  RESULT_DATE < "2011-01-01") %>% 
#                         select(RESULT_AMOUNT, RESULT_DATE)

### Plot of Beta Results Everywhere (pCi/m3)
#####################################

all_beta <- ggplot(subset(rad_data_raw, !is.na(RESULT_AMOUNT) &
              RESULT_UNIT == "PCI/M3" &
              ANALYTE_ID =="BETA"),
       aes(x=RESULT_DATE, y= RESULT_AMOUNT, color = STATE_ABBR)) + 
  geom_point() +
  annotate("text", label = "1.158 Mar '11 AK \n (Fukushima Mar '11)",
              x= as.Date("2011-03-25"),
              y= 1.5) +
  annotate("text", label = "0.703 July '08 TN",
              x= as.Date("2008-07-18"),
              y= 0.8) +
  annotate("text", label = "2.543 May '86 AL\n (Chernobyl Apr '86)",
              x= as.Date("1986-05-23"),
              y= 2.35 + 0.04) +
    annotate("text", label = "1.104 \nMar '81 NV \n (Tsuraga? Mar '81)",
              x= as.Date("1984-03-04"), #moved for display, point at 1981-03-04
              y= 1.25) +
  labs(title = 'Gross Beta in RADNet Air Filter Samples 1978-2017', 
       caption = 'EPA RadNet Data') +
  scale_x_date(date_breaks = "5 years", labels = date_format("%Y")) +
  xlab('Date') +
  ylab('pCi/m3') +
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_text(hjust=0.5),
        legend.title = element_blank())
all_beta
## Filter date to find High Points in above chart
#test <- rad_data_raw %>% filter(RESULT_AMOUNT > 1.5 & 
#                                  ANALYTE_ID == "BETA" &
#                                  RESULT_UNIT == "PCI/M3" &
#                                  RESULT_DATE < "1987-01-01") %>% 
#                         select(RESULT_AMOUNT, RESULT_DATE, STATE_ABBR, CITY_NAME)

```



```{r, echo=FALSE, message=FALSE, warning=FALSE, fukushima}
##############################
### FUKUSHIMA RESULTS IN-DEPTH March 11, 2011
##############################
fukushima <- rad_data_raw %>% 
  filter(RESULT_DATE >= "2011-03-11" & RESULT_DATE <= "2011-04-11" &  
         ANALYTE_ID == "BETA" & 
         !is.na(RESULT_AMOUNT) &
         RESULT_UNIT == "PCI/M3") %>% 
  group_by(region) %>% 
  summarise(mean = mean(RESULT_AMOUNT), n = n(), 
            value = max(RESULT_AMOUNT))

### Choropleth of Beta Air Results March 2011
############################################
#show_col(brewer_pal(palette = "PuRd")(7))
col.pal<-brewer_pal(palette = "PuRd")(7)

choro_fukushima1<-StateChoropleth$new(fukushima)
choro_fukushima1$title <- "Max Gross Beta (pCi/m3) for Air Samples"
choro_fukushima1$ggplot_scale <- scale_fill_manual(name="pCi/m3",
                                         values=col.pal, drop=FALSE)
  
choro_fukushima1$render()  +
  theme(plot.caption = element_text(size = rel(0.5)),
         plot.title = element_text(hjust=0.5),
         plot.subtitle = element_text(size=rel(.75), hjust = 0.5)) + 
  labs(subtitle = 'March 11 - April 11 2011', caption = 'EPA RadNet Data')                            
### Google Map Fukushima Beta max by monitoring station
#######################
fukushima_points <- rad_data_raw %>% 
  filter(RESULT_DATE >= "2011-03-11" & RESULT_DATE <= "2011-04-11" &  
         ANALYTE_ID == "BETA" &
        !is.na(RESULT_AMOUNT) &
         RESULT_UNIT == "PCI/M3") %>% 
  group_by(citystate) %>% 
  summarise(mean = mean(RESULT_AMOUNT), n = n(), 
            value = max(RESULT_AMOUNT), lat = mean(lat), lon = mean(lon))

monitors_fukushima <- USAMap + scale_y_continuous(limit = c(8,50)) +
  scale_x_continuous(limit = c(-130,-62)) +
  geom_point(data = fukushima_points,
             aes(x= lon, y= lat, size = value),
             col = 'red4') +
  labs(title = "Air Filter - Beta Activity ",
       subtitle = 'March 11 - April 11 2011', caption = 'EPA RadNet Data',
       size = 'pCi/L')+
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_text(hjust=0.5),
        plot.subtitle = element_text(size=rel(.75), hjust = 0.5),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.1,0.15)
        )
monitors_fukushima


### Random Investigations
#####################################

pastuerized_milk <- ggplot(subset(rad_data, MAT_ID == "PASTEURIZED MILK"),
                           aes(x=RESULT_DATE, y= RESULT_AMOUNT, color = RESULT_UNIT)) + 
  geom_point(alpha = 0.5) +
  labs(title = 'Pasteurized Milk Samples 1978-2017', 
       caption = 'EPA RadNet Data') +
  scale_x_date(date_breaks = "5 years", labels = date_format("%Y")) +
  xlab('Date') +
  ylab('All Units') +
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_text(hjust=0.5),
        legend.title = element_blank(),
        axis.text.x = element_text(angle=30))

count(subset(rad_data, ANALYTE_ID == "K40"))

potassium_air <- ggplot(subset(rad_data, (RESULT_UNIT == "PCI/M3" |
                                            RESULT_UNIT == "ACI/M3") &
                                  (ANALYTE_ID == "K" | ANALYTE_ID == "K40")),
                         aes(x=RESULT_DATE, y= RESULT_AMOUNT, color = RESULT_UNIT)) + 
  geom_point() +
  labs(title = 'K40', 
       caption = 'EPA RadNet Data') +
  scale_x_date(date_breaks = "1 years", labels = date_format("%Y")) +
  xlab('Date') +
  ylab('aCi/m3 or pCi/m3') +
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_text(hjust=0.5),
        legend.title = element_blank(),
        axis.text.x = element_text(angle=30))
potassium_air


detection_limit_k40 <- rad_data %>% 
  filter(ANALYTE_ID == "K40" &
         (RESULT_UNIT == "PCI/M3" | RESULT_UNIT == "ACI/M3")) %>% 
  group_by(citystate) %>% 
  summarise(mean = mean(RESULT_AMOUNT), n = n(), 
            value = max(RESULT_AMOUNT), lat = mean(lat), lon = mean(lon))
  
```


```{r, echo=false, groups_for_review}

### Group by BETA
#########################
beta_data <- subset(rad_data, ANALYTE_ID == "BETA")

airbeta <- subset(beta_data, MAT_ID == "AIR-FILTER" & !is.na(CSU) &
                 RESULT_UNIT == "ACI/M3")

ggplot(airbeta) +
  geom_point(aes(x=RESULT_DATE, y= RESULT_AMOUNT), colour = "red4", size =1) + 
  #geom_errorbar(aes(ymin= airbeta$RESULT_AMOUNT - airbeta$CSU, 
  #                  ymax= airbeta$RESULT_AMOUNT + airbeta$CSU),
  #             colour="black", width=.1) +
  labs(title = 'BETA Results pCi/L, Air-Filter Samples', 
       caption = 'EPA RadNet Data') +
  scale_x_date(date_breaks = "5 years", labels = date_format("%Y")) +
  scale_y_continuous(limits = c(-.05,.25)) +
  xlab('Date') +
  ylab('pCi/M3') +
  geom_errorbar(aes(x=RESULT_DATE, y= RESULT_AMOUNT + CSU), colour = "yellow")+ 
  #geom_line(aes(x=RESULT_DATE, y= MDC), colour = "green")+ 
  theme(plot.caption = element_text(size = rel(0.5)),
        plot.title = element_text(hjust=0.5),
        legend.title = element_blank(),
        axis.text.x = element_text(angle=30))


ggplot(subset(beta_data, RESULT_DATE > "2010-12-31"),
       aes(x=RESULT_DATE, y= RESULT_AMOUNT)) +
  geom_errorbar(aes(ymin = RESULT_AMOUNT - CSU, 
                    ymax = RESULT_AMOUNT + CSU),
                    colour="black", width=.1) +
  geom_line() +
  geom_point()
```


```{r, more_mini_reviews}
### Counts of results by year, faceted by City ~ Material Type
HI_samples <- ggplot(HI_data, aes(RESULT_DATE, fill = MAT_ID)) + 
  geom_histogram() +
  facet_grid(HI_data$MAT_ID~HI_data$CITY_NAME) +
  labs(title = 'Number of RAD Analyses in HI Cities by Sample Type', 
       caption = 'EPA RadNet Data', cex.main = 0.50,
       fill = 'Material Type') +
  xlab('Date') +
  ylab('count') +
  theme(plot.title = element_text(hjust = 0.5),
        plot.caption = element_text(size = rel(0.5)),
        axis.text.x = element_text(size=7, angle=45),
        axis.title.x = element_blank(),
        axis.title.x.top = element_text(size = 6),
        legend.position = "none"
  )


HI_analytes <- ggplot(HI_data, aes(RESULT_DATE, fill = ANALYTE_ID)) + 
  geom_histogram() +
  facet_wrap(~HI_data$MAT_ID) +
  labs(title = 'HI Analytes by Sample Type', 
       caption = 'EPA RadNet Data', cex.main = 0.50,
       fill = 'Material Type') +
  xlab('Date') +
  ylab('count') +
    theme(plot.title = element_text(hjust = 0.5),
        plot.caption = element_text(size = rel(0.5)),
        axis.text.x = element_text(size=7, angle=45),
        axis.title.x = element_blank(),
        axis.title.x.top = element_text(size = 6),
        legend.position = "bottom")

ggplot(subset(HI_data, HI_data$ANALYTE_ID =="BETA"), 
       aes(RESULT_AMOUNT, fill = RESULT_UNIT)) +   
       geom_histogram(bins = 15)
ggplot(subset(HI_data, HI_data$RESULT_UNIT =="PCI/L"), 
       aes(RESULT_AMOUNT, fill = MAT_ID)) +   
       geom_histogram(bins = 15)


ggplot(subset(rad_data, rad_data$ANALYTE_ID =="BETA"), 
       aes(RESULT_AMOUNT, fill = RESULT_UNIT)) +   
       geom_histogram(bins = 50)
ggplot(subset(rad_data, rad_data$RESULT_UNIT =="PCI/L"), 
       aes(RESULT_AMOUNT, fill = MAT_ID)) +   
       geom_histogram(bins = 50)

summary(subset(rad_data, rad_data$RESULT_UNIT =="PCI/L" ))


ggplot(subset(HI_data, !is.na(HI_data$RESULT_AMOUNT)), 
       aes(ANA_PROC_NUM) )+   
       geom_histogram(bins = 25)

ggplot(HI_data, aes(ANA_PROC_NUM, RESULT_AMOUNT)) + geom_col()


summary(HI_data$ANA_PROC_NUM)

```




```{r}

##CSU Combined Standard Uncertainty (95% = 2xCSU)
summary(rad_data$CSU)
ggplot(subset(rad_data, RESULT_UNIT == "PCI/L"), 
       aes(x=RESULT_DATE, y= CSU, color = ANALYTE_ID)) +
  geom_point() +
  labs(title = "CSU by year",
       y="pCi/L", x =  "YEAR")

##MDC Minimum Detectable Concentration
summary(rad_data$MDC)
ggplot(rad_data, aes(x=RESULT_DATE, y= MDC, color = ANALYTE_ID)) +
  geom_point() +
  labs(title = "MDC by year",
       y="all units", x =  "YEAR")

ggplot(subset(rad_data, RESULT_UNIT == "PCI/L"), 
       aes(x=RESULT_DATE, y= MDC, color = ANALYTE_ID)) +
  geom_point() +
  labs(title = "MDC (pCi/L) by year",
       y="pCi/L", x =  "YEAR")


## Half-Life
summary(rad_data_filt$HALF_LIFE)
count(rad_data_filt, HALF_LIFE_TIME_UNIT)

ggplot(rad_data_filt, aes(x=RESULT_DATE, y= HALF_LIFE, color = ANALYTE_ID)) +
  geom_point() +
  labs(title = "Half-Lives of Analytes",
       y="all units", x =  "YEAR")
ggplot(subset(rad_data_filt, HALF_LIFE_TIME_UNIT =="Y"), aes(x=RESULT_DATE, y= HALF_LIFE, color = ANALYTE_ID)) +
  geom_point() +
  labs(title = "Half-Lives of Analytes",
       y="Half-life Years", x =  "YEAR")

##Duration (count time in minutes)
summary(rad_data_filt$DURATION)
count(rad_data_filt, DURATION)

rad_data_filt$LOC_NUM <- as.numeric(rad_data_filt$LOC_NUM)
hist(rad_data_filt$LOC_NUM)

#  geom_histogram() +
#  labs(title = "Testing Duration",
#       y="?", x =  "Duration")
```
```{r}
ggplot(rad_data_filt, aes(LOC_NUM)) +
  geom_histogram()
```

# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!
