---
title: "RadNet Exploratory Data Analysis"
output:
  html_notebook: default
  pdf_document: default
  editor_options: 
    chunk_output_type: inline
    chunk_output_type: console
---
*by Wendy Bisset*
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages in this code chunk
# Parameter "echo" = FALSE prevents the code from displaying in the knitted #HTML output. 
# Set echo=FALSE for all code chunks in file, unless it makes sense to show #the code that generated a particular plot.
# Parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as #you want it to. This will clean up the flow of your report.

library(ggplot2)
library(dplyr)
library(scales)
library(ggmap)

## citation('ggmap')
##To cite ggmap in publications, please use:
##[^ggmap]:D. Kahle and H. Wickham. ggmap: Spatial Visualization with ggplot2. The R Journal, ##5(1), 144-161. URL http://journal.r-project.org/archive/2013-1/kahle-wickham.pdf
```

#Data Source and Background Information

The US Environmental Protection Agency maintains several public databases on environmental data.[^EPA]:<https://www.epa.gov/enviro/envirofacts-overview> One such database, ***RadNet***,[^RadNetOverview]:<https://www.epa.gov/enviro/radnet-overview> is a system of geographically distributed stations which monitor a number of radiochemical analytes (e.g., iodine-131, gross beta (β)) in the nation's air, precipitation, and drinking water. These stations are located across the US as well as American Territories. The current RadNet database primarily consists of data collected since 1978 though some data dated back to 1973 from RadNet's precursor, ERAMS (Environmental Radiation Ambient Monitoring System) is included.

While the stations of RadNet normally sample on a regular schedule, if there is concern for a significant radiation release the frequency of sampling rate can be increased. For example, there have been five times in which the sampling rates were increased: 1) 1979 Three Mile Island nuclear reactor accident in the U.S., 2) 1986 Chernobyl nuclear reactor accident in the Soviet Union, 3) 1999 Tokaimura nuclear fuel processing facility accident in Japan, 4) 2000 wildfires in Los Alamos and Hanford U.S., and 5) 2001 terrorist attacks in the U.S.

RadNet helps provide historical data needed to estimate long-term trends in environmental radiation levels and also provides a means to estimate levels of radioactivity in the environment. This data helps establish background radiation as well as identify radioactive fallout from atomic weapons testing, nuclear accidents, or other intrusions of radioactive materials. There is a timeline on the RadNet website[^RadNetTimeline]:<https://www.epa.gov/radnet/history-radnet> which lists changes in the monitoring system and events that may have impacted the measurements.  

#Downloading, Importing and Merging Data Files

All data was downloaded directly from the [RadNet](https://www.epa.gov/enviro/radnet-customized-search) website as .csv files.
According to the EPA's [Envirofacts Data Service API](https://www.epa.gov/enviro/web-services) data output is limited to 10000 rows of data at a time from a maximum of three tables. Consequently, the RadNet data was downloaded by decade in five chunks (up-to 1989, 1990-1999, 2000-2009, 2010-2017).  The resulting .csv files were then [merged](https://psychwire.wordpress.com/2011/06/03/merge-all-files-in-a-directory-using-r-into-a-single-dataframe/) into a single R Dataframe.

In total 19 variables were downloaded from a combination of three RadNet Database Tables. Data obtained from each table is broadly outlined below: 

1. Media
   + Drinking Water (DW)
   + Surface Water (SW) program  (terminated March 1999)
   + Precipitation
   + Milk (program terminated 2014)
   + Air Data combined from: 
     - Air-Charcoal (data only found in 2010-2017)
     - Air-Filter
2. Sampling Location
   + 324 Location IDs across US and Territories
   + ID number
   + city
   + state
3. Types of Radionuclides and Radiation (61 analytes (e.g., )
   + analyte
   + analyte half-life
   + analyte procedure number 
   + result amount 
   
   Variables also include units of measure, e.g. size of sample (g, L) or time of decay (sec, min, hour, day, year). No filters or limitations were placed on location or types of radionuclides analysed. Other variables such as project numbers, surface water locations, sample collection start and end etc., exist in the RadNet database, but in effort to to control file size and project scope, these data were not downloaded.

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_Data}
##!! DELETE existing dataframes if rerunning this chunk otherwise data will be re-appended

## List files (.csv) in Data Folder 
## list.files('radnet_data/', recursive = TRUE)

##Load/Merge the Data into 5 data frames for each media type
dir_list <-  list.files('radnet_data/')

for (directory in dir_list){
  subdir <- (paste('radnet_data/', directory, sep=""))
  print(subdir)
  files_in_sub <- list.files(subdir)
  dfname <- paste(directory, '_data', sep="")  

  for (file in files_in_sub){
    file_loc <- paste(subdir, '/', file, sep="")
    # print(file_loc)
    # if the merged dataset doesn't exist, create it
    if (!exists("db")){
        db <- read.csv(file_loc, header=TRUE)
    }
    # if the merged dataset does exist, append to it
    else if (exists("db")){
        temp_dataset <-read.csv(file_loc, header=TRUE)
        db <-rbind(db, temp_dataset)
       rm(temp_dataset)
    }
  }
  assign(dfname, db)
  remove(db)
}

##Cleanup Workspace
rm(dfname, dir_list, directory, file, file_loc, files_in_sub, subdir)

```

It was attempted to bind the media datasets using a loop, but it was discovered that getting a list of dataframe names using `listofDF <- as.list(names(which(sapply(.GlobalEnv, is.data.frame))))`[^df_names]: <http://r.789695.n4.nabble.com/getting-list-of-data-frame-names-td3864338.html>
generates a  character list of dataframes (listofDF). Using this list in bind_rows passes CHAR representation of the dataframe and not the object consequently an error is generated. Thus the dataframe names were ultimately hard coded into the argument of `bind_rows()`.

```{r echo=FALSE, message=FALSE, warning=FALSE, consolidate_data}
#SAMP_ID iS INT in surface_water_data and CHR in all other dataframes #convert before df consolidation into one dataframe
#ignoring factor coercion to character vectors
surface_water_data$V_ERAMS_MATRIX_SAMPLE_ANALYSIS.SAMP_ID <-
  as.character(
    surface_water_data$V_ERAMS_MATRIX_SAMPLE_ANALYSIS.SAMP_ID)

#consolidate media data into one rad_data_raw dataframe
rad_data_raw <- bind_rows(list(air_data, drinking_water_data, 
                               Milk_data, precipitation_data,
                               surface_water_data
                               ),
                         .id ="id")
```

All the variables were prepended with the name of SQL tables frome which they originated, e.g. V_ERAMS_MATRIX_SAMPLE_ANALYSIS.MAT_ID.  To shorten the variable names for ease of programming, all characters up to and including the period were removed. 

```{r echo=FALSE, message=FALSE, warning=FALSE, short_colnames}
colnames(surface_water_data) <-gsub('.+\\.','',colnames(surface_water_data))
colnames(Milk_data) <-gsub('.+\\.','',colnames(Milk_data))
colnames(air_data) <-gsub('.+\\.','',colnames(air_data))
colnames(precipitation_data) <-gsub('.+\\.','',colnames(precipitation_data))
colnames(drinking_water_data)<-
  gsub('.+\\.','',colnames(drinking_water_data))
colnames(rad_data_raw) <-gsub('.+\\.','',colnames(rad_data_raw))
```

**Univariate Plots Section**

> **Tip**: In this section, you should perform some preliminary exploration >of your dataset. Run some summaries of the data and create univariate >plots to understand the structure of the individual variables in your >dataset. Don't forget to add a comment after each plot or closely-related group of plots!
>There should be multiple code chunks and text sections; the first one below is just to help you get started.**
***
#Data Overview

A quick inspection of the dataframe and the contents of each of the variables was undertaken in order to understand the breadth of the data and potential problems.
At this stage the raw dataframe has 20 variables and 504092 observations.  One variable *id* was added as part of the dataframes merge to easily identify the original source (Media Type) of the data.   

```{r echo=FALSE, echo=FALSE, message=FALSE, warning=FALSE, DF_overview}
### DF rad_data_raw  Summary
#######################################################################
names(rad_data_raw)
dim(rad_data_raw)
str(rad_data_raw)
head(rad_data_raw)

### (1) id -> type CHR - ties merged data to original MEDIA dataframes 
###           1-5 :  (1) air_data, (2) drinking_water_data, 
###                  (3) Milk_data,(4)precipitation_data,
###                  (5) surface_water_data
#######################################################################
table(rad_data_raw$id)

```

#### (@)Variable Audit

  In the event of possible date manipulations, the date field was changed to type DATE instead of CHR.
[^as.Date]:<https://www.rdocumentation.org/packages/base/versions/3.4.1/topics/as.Date>

```{r echo=FALSE, message=FALSE, warning=FALSE, variable_review}
###  Variable Audits 2-20

### (2) MAT_ID -> type CHR, 6 material types analysed - five media types 
###               listed above where air_data is further subdivided into 
###               AIR-CHARCOAL (Discontinued in 1980s) & AIR-FILTER
#######################################################################
table(rad_data_raw$MAT_ID)
ggplot(rad_data_raw, aes(MAT_ID)) + 
  geom_bar(fill = "SteelBlue") +
  scale_x_discrete(limits = c("AIR-FILTER", "PRECIPITATION",
                   "PASTEURIZED MILK", "DRINKING WATER", 
                   "SURFACE WATER", "AIR-CHARCOAL")) +
    geom_text(stat='count',aes(label=..count..), color="black")+
  coord_flip() +
  labs(title="Number of Analyses by Media Type", x=NULL, y = "Count") +
  theme(plot.title = element_text(hjust = 0.5))

### (3) SAMP_ID -> type CHR, sample ID number, some are alpha-numeric, 
###                roughly 1/2 the number of observations, i.e, multiple
###                observations (analyses) for a single SAMP_ID
#######################################################################
n_distinct(rad_data_raw$SAMP_ID)
n_distinct(rad_data_raw$SAMP_ID)/length(rad_data_raw$SAMP_ID)
samp_replicates <- rad_data_raw %>% 
                   group_by(SAMP_ID, LOC_NUM) %>% 
                   summarise(n = n())
n_distinct(samp_replicates$SAMP_ID)
table(samp_replicates$n)
ggplot(samp_replicates, aes(n)) + 
  geom_histogram(bins = 25) +
  scale_y_sqrt() +
  labs(title = "Distribution of Sample Replicates",  
       x = "Number of Sample Replicates", y = "COUNT")

### (4) ANA_SIZE -> type NUM, numeric 
#######################################################################
ggplot(rad_data_raw, aes(ANA_SIZE)) + 
  geom_histogram(bins = 20) +
  scale_x_log10(labels=scales::comma) +
  labs(title = "Sample Sizes ... not Normalized for Unit of Measure",  
       x = "Numeric Size", y = "COUNT")

### (5) ANA_UNIT ->CHR, measurement unit for analysis size 
###                 (air = m3, solids = mg, g, liquids = mL, L, "" blank 3206)
#######################################################################
table(rad_data_raw$ANA_UNIT)
ggplot(rad_data_raw, aes(ANA_UNIT)) + 
  geom_bar() +
  labs(title = "Count of Sample Unit Sizes",  
       x = "Unit", y = "COUNT")

### (6) ANA_PROC_NUM -> INT; 34 analytical procedures used for analysis
###                     Proc Num range from 1 to 170 with mode = ProcNum 1
#######################################################################
### Calculate the mode using the getmode function
### https://www.tutorialspoint.com/r/r_mean_median_mode.htm
getmode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}
getmode(rad_data_raw$ANA_PROC_NUM)

n_distinct(rad_data_raw$ANA_PROC_NUM)
table(rad_data_raw$ANA_PROC_NUM)
ggplot(rad_data_raw, aes(ANA_PROC_NUM)) +
  geom_histogram(bins = 200, fill ="cornflowerblue", col ="black") +
  scale_x_continuous(limits = c(0,200))+
  labs(title = "Analytical Procedure Counts",  
       x = "Procedure Number", y = "COUNT")+
  theme(plot.title = element_text(hjust = 0.5))

### (7) DURATION -> NUM range 1-5000, duration of count in minutes
###                 division of Duration/60 to give hours
#######################################################################
summary(rad_data_raw$DURATION)
#table(rad_data_raw$DURATION)
ggplot(rad_data_raw, aes(DURATION/60)) + 
  geom_histogram(bins = 75, fill = "darkseagreen") +
  scale_y_log10(labels=scales::comma) +
  labs(title = "Testing Count Duration",  
       x = "Duration (h)", y = "COUNT")+
   theme(plot.title = element_text(hjust = 0.5))

### (8) LOC_NUM -> INT ID of test location 1-4157, 324 distinct locations
#######################################################################
summary(rad_data_raw$LOC_NUM)
n_distinct(rad_data_raw$LOC_NUM)
table(rad_data_raw$LOC_NUM)
ggplot(rad_data_raw, aes(LOC_NUM)) + 
  geom_histogram(bins = 50, fill = "salmon")+
  scale_y_sqrt() +
  labs(title = "Location Number Sampling Counts",  
       x = "Location ID", y = "COUNT")+
   theme(plot.title = element_text(hjust = 0.5))
  
### (9) CITY_NAME -> CHR, 289 Distinct City, county or EPA Region associated 
###                  with sampling location
#######################################################################
#n_distinct(rad_data_raw$CITY_NAME)

### (10) STATE_ABBR-> CHR; 67 state, territory or country of 
###                        sampling location
###                   50 US States +
###                   DC (District of Columbia)
###                   US Territories : PR (Puerto Rico), GU (Guam), 
###                                    PC (Panama Canal), VI (Virgin Islands)
###                   CNMI - Commonwealth of the No. Mariana Islands (Saipan)
###                  EPA Regions R01- R10
###                   ON Ottawa, ON
#######################################################################
#n_distinct(rad_data_raw$STATE_ABBR)
#table(rad_data_raw$STATE_ABBR)

### (11) ANALYTE_ID -> CHR, Analyte of interest, 61 unique analytes
###                    Abbreviation and isotope e.g., CO60 = Cobalt 60
#######################################################################
#n_distinct(rad_data_raw$ANALYTE_ID)

### (12) RESULT_AMOUNT -> NUM, range -200-257000;
###                        amount, activity or concentraion of analyte 
###                        81921 NA's (16%)
#######################################################################
#summary(rad_data_raw$RESULT_AMOUNT)
#sum(is.na(rad_data_raw$RESULT_AMOUNT))/nrow(rad_data_raw)

### (13) CSU -> NUM, 0-15000 Combined Standard Uncertainty, 
###                       81921 NA's (16%)
#######################################################################
#summary(rad_data_raw$CSU)

### (14) MDC -> NUM, 0-9700 Min Detectable Concentration
###                       218822 NA's (43%)
#######################################################################
#summary(rad_data_raw$MDC)
#sum(is.na(rad_data_raw$MDC))/nrow(rad_data_raw)

### (15) RESULT_UNIT -> unit of measure for result 
###                (air = ACI/m3, PCI/M3 (45%); solids = g/L, DPM/GC;
###                 liquids = pCi/L (48%))
#######################################################################
#table(rad_data_raw$RESULT_UNIT)
#sum(rad_data_raw$RESULT_UNIT == "PCI/L")/nrow(rad_data_raw)
#sum(rad_data_raw$RESULT_UNIT == "PCI/M3")/nrow(rad_data_raw)

### (16) RESULT_DATE -> CHR transform to DATE 
###                     range from 7/1/1978 to 7/26/2017
#######################################################################
summary(rad_data_raw$RESULT_DATE)
# Set DATE to ISO format, where %F == "%Y-%m-%d" ISO 8601 format
rad_data_raw$RESULT_DATE <- as.Date(rad_data_raw$RESULT_DATE, "%F")
summary(rad_data_raw$RESULT_DATE)

### (17) ANA_TYPE -> CHR, "E" (Element) = 10585, "R" (Radionuclide)= 493507
#######################################################################
#table(rad_data_raw$ANA_TYPE)

### (18) HALF_LIFE -> Type = unknown?; quantity part of an elements half-life
###                   251749 50% NA's
#######################################################################
summary(rad_data_raw$HALF_LIFE)
sum(is.na(rad_data_raw$HALF_LIFE))/nrow(rad_data_raw)

### (19) HALF_LIFE_TIME_UNIT -> CHR; D,H,M,S,Y and 50% blank (251749)
#######################################################################
table(rad_data_raw$HALF_LIFE_TIME_UNIT)

### (20) ANALYTE_NAME -> CHR; 61 unique full descriptive name of analyte
###                      e.g, Gross Beta, Uranium-235
###                      Expansion on (11) ANALYTE_ID
#######################################################################
n_distinct(rad_data_raw$ANALYTE_NAME)

```
 
#### (@)Data Cleanup and Wrangling (Recast Variables Types)
Many of variables were levelled factors in the original data, but as the dataframes were merged unequal level factors were to coerced to CHR data type. As some of these variables do have limited categories, such as material type, measurement units, or time units which can be addressed as factors, it may be useful to recast these variable to factors. In the case of time units it is possible to also impose an order, e.g., S(econd) < M(inute) < H(our) < D(ay) < Y(ear). Where possible the variables are recast and levelled in the context of the compiled superset dataframe in order to improve future manipulations and analysis. 

```{r echo=FALSE,message=FALSE, warning=FALSE, factor_variables}
### Recast variables in final "rad_data" (consolidated) dataframe

### (1) ID for original df source (numerical tie to media type)
rad_data_raw$id          <- factor(as.integer(rad_data_raw$id))
### (2) MAT_ID: material type e.g. Air, SW, DW, PPT, MILK
rad_data_raw$MAT_ID      <- factor(rad_data_raw$MAT_ID) 
### (5) ANA_UNIT: size of sample G,L,M3, MG, ML
rad_data_raw$ANA_UNIT    <- factor(rad_data_raw$ANA_UNIT)       
### (15) RESULT_UNIT: 5 Result measures -> ACI/M3, DPM/GC, G/L, PCI/L, PCI/M3
rad_data_raw$RESULT_UNIT <- factor(rad_data_raw$RESULT_UNIT)
### (17) ANA_TYPE Analyte two types E (Element) and R (Radionuclide)
rad_data_raw$ANA_TYPE    <- factor(rad_data_raw$ANA_TYPE)
### (19) HALF_LIFE_TIME_UNIT: Five time Measures of Half Life: S, M, H, D, Y
### Create ordered factor for Time units sec<min<hour<day<year
rad_data_raw$HALF_LIFE_TIME_UNIT <- factor(rad_data_raw$HALF_LIFE_TIME_UNIT, 
                                       levels = c("S", "M", "H", "D", "Y"))
# (20) ANALYTE_NAME: 61 different Analytes
rad_data_raw$ANALYTE_NAME <- factor(rad_data_raw$ANALYTE_NAME)

### Other potential factor variables                                       
### (8)Location ID (324 unique IDs)
#rad_data$LOC_NUM     <- factor(rad_data$LOC_NUM)
### (9)City Name 289 different cities
#rad_data$CITY_NAME   <- factor(rad_data$CITY_NAME)
### (10) State IDs, 67 different states
#rad_data$STATE_ABBR  <- factor(rad_data$STATE_ABBR)
### (11) 61 different Analytes ANALYTE_ID abbreviated ANALYTE_NAME
#rad_data$ANALYTE_ID  <- factor(rad_data$ANALYTE_ID)
```

```{r echo=FALSE, include = FALSE}
#### *DO NOT INCLUDE CHUNK IN REPORT*
#### Attempt to convert SAMP_ID to numeric, 
#### In order to easily handle samples and possibly compare multiple results on #### the same sample it was attempted to convert the SAMP_ID to numeric type, 
#### however, some SAMP_ID values are prefaced with alpha character and variable was left alone. 

rad_data$sample <- as.numeric(rad_data$SAMP_ID)
summary(rad_data$sample)

Milk_data %>% 
            select(SAMP_ID, RESULT_AMOUNT, LOC_NUM) %>%     
            filter(is.na(as.numeric(SAMP_ID))) %>% 
            count(SAMP_ID)
str(rad_data$SAMP_ID)
```


```{r echo=FALSE, include=FALSE}
### Review info for cleaned rad_data
head(rad_data_raw)
dim(rad_data_raw)
names(rad_data_raw)
summary(rad_data_raw)

summary(rad_data_raw$ANA_UNIT)
summary(rad_data_raw$RESULT_UNIT)
summary(rad_data_raw$HALF_LIFE_TIME_UNIT)
levels(rad_data_raw$HALF_LIFE_TIME_UNIT)
```

#### (@)Geocoding Locations
Using the package `ggmap`[^ggmap](https://cran.r-project.org/web/packages/ggmap/ggmap.pdf) it is possible to represent data and information geographically.  Using the function `geocode()`[^geocode](https://www.rdocumentation.org/packages/ggmap/versions/2.6.1/topics/geocode),[^geocode_csv](http://www.storybench.org/geocode-csv-addresses-r/) the latitude and longitude of locations may be obtained from the city and state information. 

```{r echo=FALSE, include=FALSE}
### Test CHUNK run of `geocode()`
### returns lat/lon and address of first row in dataframe
citystate <- paste(rad_data[1,which( colnames(rad_data)=="STATE_ABBR")],
                   rad_data[1,which( colnames(rad_data)=="CITY_NAME")])
geocode(citystate, output = "latlona", source="dsk") #datasciencetoolkit.org
geocode("Corozal, Panama", output = "latlona")
geocode("Kauai, HI",  output = "latlona", source="google")
geocodeQueryCheck(userType = "free")
```

In an attempt to map the sampling locations a few issues were found in the data entry. The state abbreviation PC for Panama Canal was not recognized in the API and consequently needed to be changed. A single entry identified as Doswell, SC does not appear to exist, but there are several Doswell, VA observations so this entry was changed. Several entries are listed by their EPA regions 1-10, in an effort to geo-locate these regions the city-state of the regions were changed to the headquarters location for that region.  Regions 01-10 are mapped to LOC_NUM 59-68, respectively and consequently may be distinguished from the city locations if needed.

```{r echo-FALSE, geo_fixes}
### Fix location errors:
### geocode does not find 'PC', changed STATE_ABBR to 'Panama' 
### One entry misentered? Doswell SC changed to Doswell VA, LOC_NUM does not
### match any other entry, but latlong.net cannot find location either
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "PC")] <- "PANAMA"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "SC") &
                          (rad_data_raw$CITY_NAME == "DOSWELL")] <- "VA"

### The 10 Regions are mapped to LOC_NUM 59-68
### Set EPA REGIONS to Headquarter Cities
### R01 New England (CT, ME, MA, NH, RI, VT and 10 tribal nations)
rad_data_raw$CITY_NAME[(rad_data_raw$CITY_NAME == "R01")]   <- "BOSTON"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R01")]  <- "MA"
### R02 NJ,NY, Puerto Rico (PR), US VI and 8 tribal nations)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R02")]   <- "New York City"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R02")]  <- "NY"
### R03 Mid-Atlantic (DE, DC, MD, PA, VA, WV)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R03")]   <- "PHILADELPHIA"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R03")]  <- "PA"
### R04 Southeast (AL, FL, GA, KY, MS, NC, SC, TN and 6 tribes)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R04")]   <- "ATLANTA"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R04")]  <- "GA"
### R05 IL, IN, MI, MN, OH, WI and 35 tribal nations                    
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R05")]   <- "CHICAGO"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R05")]  <- "IL"
### R06 South Central (AK, LA, NM, OK, TX and 66 tribal nations)      
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R06")]   <- "DALLAS"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R06")]  <- "TX"
### R07 Midwest (IA, KS, MO, NE and 9 tribal nations)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R07")]   <- "KANSAS CITY"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R07")]  <- "KS"
### R08 Mountains and Plains (CO, MT, ND, SD, UT, WY and 27 tribal nations)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R08")]  <- "DENVER"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R08")]  <- "CO"
### R09 Pacific Southwest (AZ, CA, HI, NV, Pacific Islands and 148 tribal nations)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R09")]   <- "SAN FRANCISCO"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R09")]  <- "CA"
### R10 Pacific Northwest (AK, ID, OR, WA and 271 Native Tribes)
rad_data_raw$CITY_NAME[(rad_data_raw$STATE_ABBR == "R10")]   <- "SEATTLE"
rad_data_raw$STATE_ABBR[(rad_data_raw$STATE_ABBR == "R10")]  <- "WA"
                        
```

Rather than inundate the API with multiple requests for the same location a separate dataframe was created listing only unique city-state combinations.  This removes the instances where a city may have multiple monitoring stations (LOC_NUM). 

```{r echo=FALSE, message=FALSE, warning=FALSE, get_latlona}
### Generate dataframe/list of City/State for geocoding
### Not using LOC_NUM as several cities half multiple monitoring stations
###########################################################################
places <- data.frame("CITY"   = rad_data_raw$CITY_NAME,   
                     "STATE"  = rad_data_raw$STATE_ABBR)
places$CITY <- as.character(places$CITY)
places$STATE <- as.character(places$STATE)
places <- unique(places)
places <- arrange(places, CITY)
rownames(places) <- NULL  #reset rownames
places$citystate <- paste(places$CITY, places$STATE)
places$lat     <- NA
places$lon     <- NA
places$address <- NA

##TESTING GEO CODE And check result
n=140
geo_result <- geocode(places[n,3], output = c("latlona"))
places[n,4] <- as.numeric(geo_result[2])
places[n,5] <- as.numeric(geo_result[1])
places[n,6] <- as.character(unlist(geo_result[3]))
places[n,6]

### Use tryCatch with geocode() in case of Errors from API
### to bypass error from places which have no specific LAT/LON
### a pause (1sec) was used to help not to overload API, some entries were not 
### returning results with 292 results this delays runtime ~5 min
###########################################################################
### How to handle [Errors from Geocode]
### (https://stackoverflow.com/questions/30770328/
###  how-to-handle-error-from-geocode-ggmap-r)

miss_geo <- character()  ### Create list of geocoding potential issues/misses
for (i in 1:58) {
  z <- 0
  repeat{
    geo_result <- tryCatch(geocode(places[i,3], output = c("latlona")),
                      warning = function(w) {
                        paste("Location Issue ", places[i,3]);
                        places[i,4] <- NA
                        places[i,5] <- NA  
                        places[i,6] <- NA
                      },
                      error = function(e) {
                          paste("Location error", places[i,3]);
                          next
                      })
    places[i,4] <- as.numeric(geo_result[2])
    places[i,5] <- as.numeric(geo_result[1])
    places[i,6] <- as.character(unlist(geo_result[3]))
    Sys.sleep(1)  # pause before next API request
    if (!is.na(places[i,6]) | z==2) break
    if (is.na(places[i,6])) {
      z = z+1
      #print(paste(places[i,3], "--NA", z, i))
      }
  }
  banana <- strsplit(gsub('[[:digit:]]',"", places[i,6]), ", +")[[1]]
  test_geo <-paste(toupper(banana[1]),
                     toupper(gsub('[[:blank:]]',"",banana[2])))
  if (test_geo != places[i,3]) {miss_geo <- c(miss_geo, places[i,3])}
  #if (i%%5 ==0) print(i)
}

### Merge lat/lon infor into rad_data_raw
#########################################

rad_data_raw <- left_join(rad_data_raw, places, 
                          by = c("CITY_NAME" = "CITY",
                                 "STATE_ABBR" = "STATE") )

```


```{r echo=FALSE, include = FALSE, adjust_latlon }
summary(places$lat)
summary(places$lon)

### Revised latlong data from latlong.net - 
### geocoding from source = "dsk" puts Honolulu in ocean 
###           and Kauai on the wrong island as found during mapping
###########################################################################
#places$lat[places$CITY == "HONOLULU"] = 21.306944
#places$lon[places$CITY == "HONOLULU"] = -157.858333
#places$lat[places$CITY == "KAUAI"] = 22.096440
#places$lon[places$CITY == "KAUAI"] = -159.526124
```

##Graphic Visualization with R's ggmap

<https://blog.dominodatalab.com/geographic-visualization-with-rs-ggmaps/>
<https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/ggmap/ggmapCheatsheet.pdf>
--------

```{r echo=FALSE, maptest}
#qmap(location = "New Mexico State University")
usa_center = as.numeric(geocode("United States"))
usa_cont_bounds = c(left = -130.0,bottom = 11.0, right = -60.0, top = 51.0)
USAMap = ggmap(get_googlemap(center=usa_center, scale=2, zoom=3, 
                             extent ="device"))
USAMap +
  scale_y_continuous(limit = c(8,50)) +
  scale_x_continuous(limit = c(-130,-62))          

continental <- places %>% filter(between(lon, -130, -60) & 
                                      between(lat, 8, 55))
noncontinental <- places %>% filter(!between(lon, -130, -60))

USAMap + 
  scale_y_continuous(limit = c(8,50)) +
  scale_x_continuous(limit = c(-130,-62)) +
  geom_point(data = continental, aes(x= lon, y= lat), col="blue", alpha=0.4) 
  
```
```{r}

###Test mapping
USAmap2<-get_map(location='united states', zoom=3, maptype = "terrain",
             source='google',color='color', urlonly = TRUE)
USAmap2<-get_openstreetmap(bbox=usa_cont_bounds,scale = 475000000,
                           color='color')
 
USAmap2

myMap <- get_stamenmap(bbox=usa_cont_bounds)
ggmap(myMap)

```


```{r echo=FALSE}
#qmplot(lon, lat, data = places)

us <- c(left = -125, bottom = 25.75, right = -67, top = 49)
map <- get_stamenmap(us, zoom = 5, maptype = "toner-lite")
ggmap(map) + geom_point(data = continental, aes(x= lon, y= lat),
                    col="blue", alpha=0.4)
```



```{r echo=FALSE, HAWAII}
#HAWAII AREA MAPS
HI_center = c(-157.9, 21.2)
HIMap = ggmap(get_googlemap(center=HI_center, zoom=7, extent="normal"))+
  scale_y_continuous(limit = c(19, 23)) #+
  scale_x_continuous(limit = c(-161,-154))
HIMap

hawaii <- noncontinental %>% filter(between(lon,-164,-150) & 
                                    between(lat,16,24))
HIMap  + geom_point(data = hawaii, aes(x= lon, y= lat),
                    col="red", alpha=0.9)
HI_data <- rad_data %>% filter(STATE_ABBR == "HI")
```

```{r}
#bounding box HI 
HI <- c(left = -161, bottom = 18, right = -154, top = 23)
HI <- c( -161, 23, -154, 18)
HImap <- get_stamenmap(bbox= HI, zoom = 7, maptype = "toner-lite")
ggmap(HImap) + 
  geom_point(data = hawaii, aes(x= lon, y= lat),
                    col="red", alpha=0.9)
```


```{r echo=FALSE, ALASKA}

### ALASKA AREA MAPS
#geocode("Alaska", output = "latlon")
AK_center = c(-149.4937, 64.20084)
AKMap = ggmap(get_googlemap(center=AK_center, zoom=4, extent="normal")) #+
  scale_y_continuous(limit = c(,)) #+
  scale_x_continuous(limit = c())
AKMap

alaska <- noncontinental %>% filter(between(lon,-167,-130) & 
                                    between(lat,53,65))
AKMap  + geom_point(data = alaska, aes(x= lon, y= lat),
                    col="purple4", alpha=0.9)
```


```{r echo=FALSE, micronesia}

### MICRONESIA AREA MAPS
#geocode("Songsong", output = "latlon") #center of Marianna Islands (MIsle)
MIsle_center = c(145.1459, 14.14199)
MIsleMap = ggmap(get_googlemap(center=MIsle_center, zoom=8, extent="normal")) #+
  scale_y_continuous(limit = c(,)) #+
  scale_x_continuous(limit = c())
MIsleMap

mariannas <- noncontinental %>% filter(between(lon,140,150))
MIsleMap  + geom_point(data = mariannas, aes(x= lon, y= lat),
                    col="deeppink", alpha=0.9)
```
```


```{r}
#How Many Region Values are there?
regions <- filter(rad_data, grepl('REGION', rad_data$CITY_NAME))
table(regions$MAT_ID)

```

```{r}
#OVERVIEW OF HI DATA
#Remove observations with no RESULT_AMOUNT
count(HI_data, !is.na(HI_data$RESULT_AMOUNT))
HI_data_filt <- HI_data[complete.cases(HI_data[,11]),]
count(HI_data_filt, !is.na(HI_data_filt$RESULT_AMOUNT))

table(droplevels(HI_data_filt$ANALYTE_NAME))
ggplot(HI_data_filt, aes(RESULT_DATE, fill = MAT_ID)) + 
  geom_histogram(bins = 50, alpha = 0.6)

ggplot(HI_data_filt, aes(RESULT_DATE, fill = ANALYTE_NAME)) + 
  geom_histogram(bins = 50, alpha = 0.6)

ggplot(subset(HI_data_filt, (
                              RESULT_UNIT == 'PCI/L' & 
                              MAT_ID == 'PASTEURIZED MILK')),
              aes(RESULT_DATE, RESULT_AMOUNT)) +
        geom_point()

length(unique(HI_data_filt$SAMP_ID))
str(HI_data_filt$SAMP_ID)
```
Of the 504,092 observations in the combined data 81,921 do not actually have a result amount 
```{r echo=FALSE, Univariate_Plots}

# Find observations without Results 81,291 of the 504,092 observations have
# NA for RESULT AMOUNT
count(rad_data, !is.na(rad_data$RESULT_AMOUNT))
head(filter(rad_data, is.na(rad_data$RESULT_AMOUNT)), 3)

# Filter out the NA rows
rad_data_filt <- rad_data[complete.cases(rad_data[,"RESULT_AMOUNT"]),]
# count(rad_data_filt, !is.na(rad_data_filt$RESULT_AMOUNT))

# Count the responses based on the categories within the 
# different variables, top counts listed in comments
count(rad_data_filt, MAT_ID)       # Predominantly AIR-Filter 244,590      
count(rad_data_filt, ANALYTE_NAME) # Mostly Gross Beta 232,387
count(rad_data_filt, ANA_PROC_NUM) # Procedure #1 217744
count(rad_data_filt, CITY_NAME)    # varied top 5000- 8000
count(rad_data_filt, STATE_ABBR)   # varied top 10K - 15K
count(rad_data_filt, LOC_NUM)
count(rad_data_filt, ANA_UNIT)     # M3 followed by L
count(rad_data_filt, RESULT_UNIT)  # PCI/M3 followed by PCI/L
 
#MAT_ID, count of observations Fill by MAT_ID
ggplot(rad_data_filt, aes(RESULT_DATE, fill = MAT_ID)) + 
  geom_histogram(bins = 50)
ggplot(rad_data_filt, aes(MAT_ID)) + 
  geom_bar()
ggplot(rad_data_filt, aes(CITY_NAME)) + 
  geom_bar()

#Analytes, count of analytes
analytes_rad <- rad_data_filt %>% group_by(ANALYTE_ID) %>% 
  summarise(number = n())
summary(analytes_rad)
# test <- analytes_rad[order(analytes_rad$number),]

ggplot(HI_data_filt, aes(ANALYTE_ID)) + geom_bar()

#places: Cities and States
head(filter(rad_data_filt, STATE_ABBR == "ON"), 3)

##########################################################
### PLOTS Preliminary Overview with limited HI DATA  #####
##########################################################

## (1) Plot of Result Amounts from all sources, units 
min(HI_data_filt[,"RESULT_AMOUNT"])
summary(HI_data_filt$RESULT_logmod)
HI_data_filt$RESULT_logmod <- (HI_data_filt$RESULT_AMOUNT + 
                               abs(min(HI_data_filt[,"RESULT_AMOUNT"])) + 
                               1.1)
ggplot(HI_data_filt, aes(x=RESULT_DATE, 
                    y= RESULT_logmod, colour = MAT_ID)) +
  geom_point() +
  scale_y_log10() + 
  geom_hline(yintercept = 201) +
  labs(title = 'Results of RAD Analyses in HI', 
       subtitle = 
         'log transformation of data (log(Result +min(Result) + 1)',
       caption = 'EPA RadNet Data',
       colour = 'Material Type') +
   xlab('Date') +
   ylab('Result Amount (NO normalization of units)') +
   theme(plot.caption = element_text(size = rel(0.5)),
         plot.title = element_text(hjust=0.5),
         plot.subtitle = element_text(size=rel(.75), hjust = 0.5))

## (2) Plot of Result Amounts Faceted by Unit Measure 
ggplot(HI_data_filt, aes(x=RESULT_DATE, y= RESULT_AMOUNT, 
                         color = MAT_ID)) +
  geom_point() +
  facet_wrap(~HI_data_filt$RESULT_UNIT, scales = "free") +
  labs(title = 'Results of RAD Analyses in HI by Unit Measure', 
       caption = 'EPA RadNet Data',
       colour = 'Material Type') +
   xlab('Date') +
   ylab('Result Amount') +
   theme(plot.caption = element_text(size = rel(0.5)),
         plot.title = element_text(hjust=0.5),
         legend.text = element_text(size = rel(0.5)),
         legend.position = c(0.85,0.2),
         legend.title.align = 0.5)

## (3) Plot of Beta Results (pCi/m3)
ggplot(subset(HI_data_filt, RESULT_UNIT == "PCI/M3" & ANALYTE_ID =="BETA"),
     aes(x=RESULT_DATE, y= RESULT_AMOUNT)) + 
     geom_point()+
     annotate("text", label = "0.894 Mar '11", x= as.Date("2011-03-25"),
              y=0.894-0.035) +
  annotate("text", label = "0.3758 May '86", x= as.Date("1986-05-18"),
              y=0.3758 + 0.04) +
  labs(title = 'Gross Beta in HI Air Filter Samples 1978-2017', 
       caption = 'EPA RadNet Data') +
  scale_x_date(date_breaks = "5 years", labels = date_format("%Y")) +
   xlab('Date') +
   ylab('pCi/m3') +
   theme(plot.caption = element_text(size = rel(0.5)),
         plot.title = element_text(hjust=0.5))

## Filter date for High Points in above chart
test <- HI_data_filt %>% filter(RESULT_AMOUNT > 0.25 & 
                                  ANALYTE_ID == "BETA" &
                                  RESULT_UNIT == "PCI/M3" &
                                  RESULT_DATE < "2011-01-01") %>% 
                         select(RESULT_AMOUNT, RESULT_DATE)

## Counts of results by year, faceted by City, coloured by Material Type
ggplot(HI_data, aes(RESULT_DATE, fill = MAT_ID)) + 
  geom_histogram() +
  facet_wrap(~HI_data$CITY_NAME) +
  labs(title = 'Number of RAD Analyses in HI Cities', 
       caption = 'EPA RadNet Data', cex.main = 0.50,
       fill = 'Material Type') +
  xlab('Date') +
  ylab('count') +
  theme(plot.caption = element_text(size = rel(0.5)))

## HISTOGRAMS

ggplot(HI_data, aes(RESULT_DATE, fill = ANALYTE_ID)) + 
  geom_histogram() +
  facet_wrap(~HI_data$MAT_ID)

ggplot(subset(HI_data, HI_data$ANALYTE_ID =="BETA"), 
       aes(RESULT_AMOUNT, fill = RESULT_UNIT)) +   
       geom_histogram(bins = 15)
ggplot(subset(HI_data, HI_data$RESULT_UNIT =="PCI/L"), 
       aes(RESULT_AMOUNT, fill = MAT_ID)) +   
       geom_histogram(bins = 15)
ggplot(subset(HI_data, !is.na(HI_data$RESULT_AMOUNT)), 
       aes(ANA_PROC_NUM) )+   
       geom_histogram(bins = 25)
ggplot(subset(HI_data, !is.na(HI_data$RESULT_AMOUNT)), 
       aes(ANA_PROC_NUM) )+   
       geom_histogram(bins = 25)
ggplot(HI_data, aes(ANA_PROC_NUM, RESULT_AMOUNT)) + geom_col()


summary(HI_data$ANA_PROC_NUM)

ggplot(HI_data, aes(table, price, color = cut)) +
    geom_point()+
    scale_color_brewer(type='qual')+
  scale_x_continuous(name ="Table", breaks=seq(50,80,2), limits = c(50,80))
```



```{r}

#
```




```{r}

##CSU Combined Standard Uncertainty (95% = 2xCSU)
summary(rad_data_filt$CSU)
ggplot(subset(rad_data_filt, RESULT_UNIT == "PCI/L"), aes(x=RESULT_DATE, y= CSU, color = ANALYTE_ID)) +
  geom_point() +
  labs(title = "CSU by year",
       y="pCi/L", x =  "YEAR")

##MDC Minimum Detectable Concentration
summary(rad_data_filt$MDC)
ggplot(rad_data_filt, aes(x=RESULT_DATE, y= MDC, color = ANALYTE_ID)) +
  geom_point() +
  labs(title = "MDC by year",
       y="all units", x =  "YEAR")

ggplot(subset(rad_data_filt, !is.na(RESULT_UNIT) & RESULT_UNIT == "PCI/L"), aes(x=RESULT_DATE, y= MDC, color = ANALYTE_ID)) +
  geom_point() +
  labs(title = "MDC (pCi/L) by year",
       y="pCi/L", x =  "YEAR")


## Half-Life
summary(rad_data_filt$HALF_LIFE)
count(rad_data_filt, HALF_LIFE_TIME_UNIT)

ggplot(rad_data_filt, aes(x=RESULT_DATE, y= HALF_LIFE, color = ANALYTE_ID)) +
  geom_point() +
  labs(title = "Half-Lives of Analytes",
       y="all units", x =  "YEAR")
ggplot(subset(rad_data_filt, HALF_LIFE_TIME_UNIT =="Y"), aes(x=RESULT_DATE, y= HALF_LIFE, color = ANALYTE_ID)) +
  geom_point() +
  labs(title = "Half-Lives of Analytes",
       y="Half-life Years", x =  "YEAR")

##Duration (count time in minutes)
summary(rad_data_filt$DURATION)
count(rad_data_filt, DURATION)

rad_data_filt$LOC_NUM <- as.numeric(rad_data_filt$LOC_NUM)
hist(rad_data_filt$LOC_NUM)

#  geom_histogram() +
#  labs(title = "Testing Duration",
#       y="?", x =  "Duration")
```
```{r}
ggplot(rad_data_filt, aes(LOC_NUM)) +
  geom_histogram()
```

# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!
